### 2021-10-13 sports_analytics.Rmd
initial commit

---
title: "Tools for Sports Analytics"
authors:
  - name: Benjamin S. Baumer
    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
abstract: |
  The abstract should be a concise (less than 250 words) description of the article and its implications. It should include all keywords associated with your article, as keywords increase its discoverability.
keywords:
  - sports analytics
  - R packages
  - sports data
  - datasets
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   # use apa for WIRE
output: rticles::arxiv_article
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  fig.align = 'center'
)
library(tidyverse)
```

```{r pkg}
knitr::write_bib(
  c(
    "Lahman", "pitchRx", "retrosheet",
    "teamcolors",
    "ggplot2"
  ),
  file = "pkgs.bib"
)
```

```{=latex}
\newcommand{\pkg}[1]{\textbf{#1}}
```

**Remember that you are writing for an interdisciplinary audience. Please be sure to discuss interdisciplinary themes, issues, debates, etc. where appropriate.** Note that the WIREs are forums for review articles, rather than primary literature describing the results of original research.

## Article Title

The title should not exceed 20 words. Please be original and try to include keywords, especially before a colon if applicable, as they will increase the discoverability of your article. [Tips on Search Engine Optimization](https://authorservices.wiley.com/author-resources/Journal-Authors/Prepare/writing-for-seo.html)

## Article Category

- Overview

## Graphical/Visual Abstract and Caption

Include an attractive full color image to go under the text abstract and in the online Table of Contents.
You will also need to upload this as a separate file during submission.
It may be a figure or panel from the article or may be specifically designed as a visual summary. While original images are preferred, if you need to look for a thematically appropriate stock image, you can go to pixabay.com (not affiliated with Wiley) to find a free stock image with a CC0 license. Another option you have is to utilize professional illustrators with Wiley’s Graphical Abstract Design service.

Size: The minimum resolution is 300 dpi. Please keep the image as simple as possible because it will be displayed in multiple sizes. Multiple panels and text other than labels are strongly discouraged.

Caption: This is a narrative sentence to convey the article's essence and wider implications to a non-specialist audience. The maximum length is 50 words, but consider using 280 characters or less to facilitate social media sharing, which can increase the discoverability of your article.

# Introduction

Introduce your topic in ~2 paragraphs, ~750 words.

While Wiley does consider articles on preprint servers (ArXiv, bioRxiv, psyArXiv, SocArXiv, engrXiv, etc.) for submission to primary research journals, preprint articles should not be cited in WIREs manuscripts as review articles should discuss and draw conclusions only from peer-reviewed research. Remember that original research/unpublished work should also not be included as it has not yet been peer-reviewed and could put the work in jeopardy of getting published in the primary press.

- [CRAN Task View SportsAnalytics](https://github.com/beanumber/ctv-sportsanalytics)

# Baseball

@marchi2018analyzing

@R-Lahman, @R-pitchRx, @R-retrosheet

# Basketball

# Football

## American

## Australian Rules

## Rugby

# Hockey

# Soccer

# Chess

Figure \ref{fig:chess-startup} shows a rendering of the starting chess board obtained via the \pkg{chess} package.

```{r chess-startup, fig.cap="The starting chess board printed via the \\pkg{chess} package."}
library(chess)
game() %>%
  plot()
```

# Other sports

# APIs

# Color palettes

@R-teamcolors provides color palettes for professional and amateur sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
Figure \ref{fig:teamcolors} illustrates team color palettes for professional sports leagues.

```{r teamcolors, warning=FALSE, fig.cap="Professional sports team color palettes provided by the \\pkg{teamcolors} package. ", fig.height=9}
teamcolors::show_team_col()
```

# Conclusion

Sum up the key conclusions of your review, highlighting the most promising scientific developments, directions for future research, applications, etc. The conclusion should be ~2 paragraphs, ~750 words total.

# Acknowledgments

List contributions from individuals who do not meet the criteria for authorship (for example, to recognize people who provided technical help, collation of data, writing assistance, acquisition of funding, or a department chairperson who provided general support), with permission from the individual. Thanks to anonymous reviewers are not appropriate.

# References {-}

### 2021-11-01 sports_analytics.Rmd
drafted start of initial section

@@
 ---
-title: "Tools for Sports Analytics"
+title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
 authors:
@@
 abstract: |
-  The abstract should be a concise (less than 250 words) description of the article and its implications.
-It should include all keywords associated with your article, as keywords increase its discoverability.
+  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years.
+We connect four(?) big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport.
+While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data.
+Where appropriate, we outline how data, models, and knowledge of the sport combine to generate actionable insights.
+This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics.
 keywords:
@@
 
+> Charge: Given your expertise and background, a review article from your group related to the topic of Statistical tools for sports analytics (e.g.
+baseball) would significantly contribute to WIREs Computational Statistics.
+>
+> The article should critically present the current state-of-the-art research, include opposing viewpoints, and identify challenges and opportunities.
+
 **Remember that you are writing for an interdisciplinary audience.
@@
 where appropriate.** Note that the WIREs are forums for review articles, rather than primary literature describing the results of original research.
-
-## Article Title
-
-The title should not exceed 20 words.
-Please be original and try to include keywords, especially before a colon if applicable, as they will increase the discoverability of your article.
-[Tips on Search Engine Optimization](https://authorservices.wiley.com/author-resources/Journal-Authors/Prepare/writing-for-seo.html)
 
@@
 
-Include an attractive full color image to go under the text abstract and in the online Table of Contents.
+> Include an attractive full color image to go under the text abstract and in the online Table of Contents.
 You will also need to upload this as a separate file during submission.
@@
 
-Size: The minimum resolution is 300 dpi.
+> Size: The minimum resolution is 300 dpi.
 Please keep the image as simple as possible because it will be displayed in multiple sizes.
@@
 
-Caption: This is a narrative sentence to convey the article's essence and wider implications to a non-specialist audience.
+> Caption: This is a narrative sentence to convey the article's essence and wider implications to a non-specialist audience.
 The maximum length is 50 words, but consider using 280 characters or less to facilitate social media sharing, which can increase the discoverability of your article.
@@
 
-Introduce your topic in ~2 paragraphs, ~750 words.
+> Introduce your topic in ~2 paragraphs, ~750 words.
 
-While Wiley does consider articles on preprint servers (ArXiv, bioRxiv, psyArXiv, SocArXiv, engrXiv, etc.) for submission to primary research journals, preprint articles should not be cited in WIREs manuscripts as review articles should discuss and draw conclusions only from peer-reviewed research.
+> While Wiley does consider articles on preprint servers (ArXiv, bioRxiv, psyArXiv, SocArXiv, engrXiv, etc.) for submission to primary research journals, preprint articles should not be cited in WIREs manuscripts as review articles should discuss and draw conclusions only from peer-reviewed research.
 Remember that original research/unpublished work should also not be included as it has not yet been peer-reviewed and could put the work in jeopardy of getting published in the primary press.
@@
 
-# Baseball
+# The expected value of a possession
+
+In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a possession at any given point in the game. 
+@lindsey1963investigation makes what is likely the most impactful such line of inquiry in baseball, leading to the analysis of all types of common strategies.
+
+For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.
+@lindsey1963investigation used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these two situations. 
+His, and many subsequent analyses [@tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
+
+Mathematically, we define $X$ to be a random variable indicating the number of runs (points) that a team will score in the remainder of a period, or game. 
+Let $s$ be a tuple that encodes the *state* of a game. 
+Then our task is to estimate:
+$$
+  \mathbb{E}[X | s] \,,
+$$
+for any state $s$. 
+In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
+Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball. 
+The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* and it is foundational in baseball analytics. 
+See [blah] for examples of analytical findings that are based on the expected run matrix.
+
+However, this concept is easily extended to other sports. 
+In American football, $s$ is determined by down, yardage to the next first down, and field position. 
+Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.
+
+Even in sports where the concept of a state is not so easy to define, like basketball, the value of a possession can be estimated with player tracking data [@cervone2016multiresolution].
+
+## Win probability
+
+A related, but different, concept is the notion of *win probability*, which is simply an estimate of the probability that either team will win the game, given its current state.
+
+## Other stuff
 
@@
 @R-Lahman, @R-pitchRx, @R-retrosheet
-
-# Basketball
-
-# Football
-
-## American
-
-## Australian Rules
-
-## Rugby
-
-# Hockey
-
-# Soccer
-
-# Chess
 
@@
 ```
-
-# Other sports
-
-# APIs
 

### 2021-11-01 sports_analytics.Rmd
Quang's author info

@@
     ORCID: 0000-0002-8413-5097
+  - name: Quang Nguyen
+    department: Mathematics and Statistics
+    affiliation: Loyola University Chicago
+    location: Chicago, IL 60660
+    email: nnguyen22@luc.edu
 abstract: |

### 2021-11-15 sports_analytics.Rmd
I had to use reticulate.

@@
 library(chess)
-game() %>%
-  plot()
+Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python")
+game() %>% plot()
 ```

### 2021-11-16 sports_analytics.Rmd
Update sports_analytics.Rmd
added self as author

@@
     email: nnguyen22@luc.edu
+  - name: Katherine Evans
+    affiliation: Monumental Basketball
+    location: Washington, DC 20003
+    email: kevans@monumentalsports.com
 abstract: |

### 2021-11-18 sports_analytics.Rmd
g

--- 2021-11-01 sports_analytics.Rmd
@@
 library(chess)
-game() %>%
-  plot()
+Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
+game() %>% plot()
 ```

### 2021-11-18 sports_analytics.Rmd

--- 2021-11-16 sports_analytics.Rmd
@@
 library(chess)
-Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python")
+Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
 game() %>% plot()

### 2021-12-02 sports_analytics.Rmd
added somew football references.

--- 2021-11-16 sports_analytics.Rmd
@@
 for any state $s$.
+
 In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
 Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball.
+
 The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* and it is foundational in baseball analytics.
+
 See [blah] for examples of analytical findings that are based on the expected run matrix.
@@
 In American football, $s$ is determined by down, yardage to the next first down, and field position.
+
 Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.
+
+@Pattani2012 describes ESPN's expected points metric.
+
+@CarrollPalmer1998 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.
+
+@nflwar, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM).
+
+@CarterMachol1971 estimate the expected points for 1st and 10 given a yard line (broken up into 10 yard chunks centered at a given yardline (e.g.
+5, 15, 25, 35, etc.
+)).
+This is wild that they did this in 1971!
 
@@
 library(chess)
-Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python")
+#Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
 game() %>% plot()

### 2022-05-17 sports_analytics.Rmd
soccer refs
WP refs

@@
 
+@fernandez2021epvsoccer A framework for the fine-grained evaluation of the instantaneous expected value of soccer possessions
+
+@decroos2019playeraction Actions Speak Louder than Goals: Valuing Player Actions in Soccer
+
 ## Win probability
@@
 A related, but different, concept is the notion of *win probability*, which is simply an estimate of the probability that either team will win the game, given its current state.
+
+Some refs (cite later)
+
+https://www.degruyter.com/document/doi/10.1515/jqas-2013-0100/html
+https://www.taylorfrancis.com/chapters/edit/10.1201/9781315166070-14/situational-success-evaluating-decision-making-football-keith-goldner
+http://stat.wharton.upenn.edu/~stjensen/papers/shanejensen.basketball2016.pdf
+https://arxiv.org/pdf/2204.11777.pdf
 

### 2022-05-25 sports_analytics.Rmd
Updates.

--- 2021-12-02 sports_analytics.Rmd
@@
 
-In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a possession at any given point in the game. 
-@lindsey1963investigation makes what is likely the most impactful such line of inquiry in baseball, leading to the analysis of all types of common strategies.
+In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a game state at any given point in a sporting contest.
 
-For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.
-@lindsey1963investigation used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these two situations. 
-His, and many subsequent analyses [@tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
-
-Mathematically, we define $X$ to be a random variable indicating the number of runs (points) that a team will score in the remainder of a period, or game. 
+Mathematically, we define $X$ to be a random variable indicating the the value of a particular game state for a team over some determined amount of time (e.g.
+remainder of game, quarter, period, or inning).
 Let $s$ be a tuple that encodes the *state* of a game.
@@
 
+For simplicity, we start with baseball due to the discrete nature of the sport.
 In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities).
-Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball.
+Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball (25, if you include the absorbing state of 3 outs), and we are often interested in the number of runs that will be scored from some state until the end of the inning.
+Thus in this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$.
+The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix*, and it is foundational in baseball analytics.
 
-The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* and it is foundational in baseball analytics.
+Early work on this topic can be found in @lindsey1963investigation, which used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning.
+This line of work led to work analyzing all types of common baseball strategies.
+For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.
+Theirs, and many subsequent analyses [@tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
+
+(Comment: Page 10 of this paper (https://www.jstor.org/stable/pdf/167996.pdf) actually have the expected values that were estimated.
+Do we want to include an example of this table?
+Might be cool to compare to like the numbers we got from open WAR.)
 
@@
 ```{r chess-startup, fig.cap="The starting chess board printed via the \\pkg{chess} package."}
-library(chess)
+#library(chess)
 #Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
-game() %>% plot()
+#game() %>% plot()
 ```

### 2022-05-27 sports_analytics.Rmd
more win prob refs (will cite eventually)

--- 2022-05-25 sports_analytics.Rmd
@@
 A related, but different, concept is the notion of *win probability*, which is simply an estimate of the probability that either team will win the game, given its current state.
+
+Some refs (cite later)
+
+https://www.degruyter.com/document/doi/10.1515/jqas-2013-0100/html
+
+https://www.taylorfrancis.com/chapters/edit/10.1201/9781315166070-14/situational-success-evaluating-decision-making-football-keith-goldner
+
+http://stat.wharton.upenn.edu/~stjensen/papers/shanejensen.basketball2016.pdf
+
+https://arxiv.org/pdf/2204.11777.pdf
+
+https://content.iospress.com/download/journal-of-sports-analytics/jsa231?id=journal-of-sports-analytics%2Fjsa231
+
+https://content.iospress.com/download/journal-of-sports-analytics/jsa279?id=journal-of-sports-analytics%2Fjsa279
+
+https://content.iospress.com/articles/journal-of-sports-analytics/jsa190294?resultNumber=10&totalResults=1420&start=10&q=win+probability&resultsPageSize=10&rows=10#jsa-5-jsa190294-g009
+
+https://www.degruyter.com/document/doi/10.1515/jqas-2019-0096/html
+
+https://www.degruyter.com/document/doi/10.1515/jqas-2019-0089/html
+
+https://www.degruyter.com/document/doi/10.1515/jqas-2019-0056/html
+
+https://wires.onlinelibrary.wiley.com/doi/epdf/10.1002/wics.1358?saml_referrer
+
+https://wires.onlinelibrary.wiley.com/doi/epdf/10.1002/wics.1483
 

### 2022-05-31 sports_analytics.Rmd
started on win prob and team strength

--- 2022-05-25 sports_analytics.Rmd
@@
     "teamcolors",
-    "ggplot2"
+    "ggplot2",
+    "BradleyTerry2"
   ),
@@
 For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.
-Theirs, and many subsequent analyses [@tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
+Theirs, and many subsequent analyses [@Tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
 
@@
 
-@CarrollPalmer1998 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.
+@CarrollPalmer1988 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.
 
-@nflwar, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM).
+@Yurko2019, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM).
 
@@
 
-## Win probability
+@fernandez2021epvsoccer A framework for the fine-grained evaluation of the instantaneous expected value of soccer possessions
+
+@decroos2019playeraction Actions Speak Louder than Goals: Valuing Player Actions in Soccer
+
+# Win probability
 
 A related, but different, concept is the notion of *win probability*, which is simply an estimate of the probability that either team will win the game, given its current state.
+Win probability provides a measure of the impact of sports plays and individual player contributions, as discussed in @Albert2015.
+Given its popularity, recent books on sports analytics often offers multiple chapters entirely on win probability.
+These include @Albert2001curve, @Schwarz2005numbers, @Tango2007book, @Albert2017handbook, and @Winston2022.
+In this section, we discuss notable previous work on win probability in baseball, football, basketball, and several other sports.
+
+## Baseball
+
+@Albert2015 provided a historical overview of the use of win probability in baseball.
+This notion goes back to at least as early as @Lindsey1961, where the expected win probability after each inning was calculated based on the distribution of runs scored in each inning.
+@Mills1970 utilized win probability to introduce Player Win Average (PWA), a measure of a player's contribute to the game outcome.
+In particular, PWA was computed as
+$$
+PWA = \frac{Win \ Points}{Win \ Points + Loss \ Points},
+$$
+where the win and loss points represent how much the player positively and negatively impact their team's probability of winning after each play.
+
+## Football
+
+In football,
+
+Burke WPA
+
+http://www.advancedfootballanalytics.com/2010/01/win-probability-added-wpa-explained.html
+
+@Lock2014 used random forests as a means for estimating WP before each play in a football game.
+
+@Yurko2019
+
+@Yam2019 causal
+
+nflscrapr, nflfastR, nflverse
+
+@Yurko2020 going deep
+
+## Basketball
+
+@Stern1994 brownian
+
+@Deshpande2016 bayesian
+
+@McFarlane2019 nba logistic
+
+## Other sports
+
+https://arxiv.org/pdf/2204.11777.pdf
+
+@Brenzel2019 curling markov
+
+@Maymin2020 esport!
+
+# Team strength
+
+A popular method for estimating team strength in sports is through pairwise evaluations.
+Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952].
+For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$.
+Then the BTM is a logistic regression model with parameters $\{\beta_i\}$ such that
+
+$$
+log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j
+$$
+
+Here, $exp(\beta_i)$ is often viewed as a representation of player $i$'s ability. The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2].
+
+@Elo1978
+
+@Glickman1998 state-space model
+
+@Lopez2018 extension to @Glickman1998's framework
 

### 2022-06-01 sports_analytics.Rmd
added to section 1

@@
 remainder of game, quarter, period, or inning).
-Let $s$ be a tuple that encodes the *state* of a game.
+Let $s \in S$ be a tuple that encodes the *state* of a game. 
 Then our task is to estimate:
@@
 $$
-for any state $s$.
+for any state $s \in S$, where $S$ is the set of all possible states.
 
-For simplicity, we start with baseball due to the discrete nature of the sport.
+## In baseball, the expected run matrix
+
+For clarity, we start with baseball due to the discrete nature of the sport. 
 In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
-Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball (25, if you include the absorbing state of 3 outs), and we are often interested in the number of runs that will be scored from some state until the end of the inning.
-Thus in this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$.
-The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix*, and it is foundational in baseball analytics.
+Thus, there $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball (25, if you include the absorbing state of 3 outs), and we are often interested in the number of runs that will be scored from some state until the end of the inning.
+In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
+The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
+Figure \ref{fig:lindsey} shows a reproduction of Lindsey's original expected run matrix.
 
-Early work on this topic can be found in @lindsey1963investigation, which used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning.
-This line of work led to work analyzing all types of common baseball strategies.
+[^matrix]: There is no inherent dimensionality to $\mathbb{E}[X | s]$.
+The *matrix* nomenclature stems from its values typically being displayed in $8 \times 3$ grid.
+However, when computing with $\mathbb{E}[X | s]$, it is most often convenient to treat it as a $24 \times 1$ vector.
+
+```{r lindsey, out.width="90%", fig.cap="Table 1 from Lindsey's original paper.
+The column labelled $E(T, B)$ gives the expected run matrix as a vector, based on Lindsey's analysis of Major League Baseball data from 1959 and 1960.
+"}
+knitr::include_graphics("lindsey_table1.png")
+```
+
+Early work on this topic can be found in @lindsey1963investigation, who used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning.
+This line of work led to analysis of all types of common baseball strategies. 
 For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.
-Theirs, and many subsequent analyses [@Tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
+@Tango2007book (and many subsequent analyses), conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
 
-(Comment: Page 10 of this paper (https://www.jstor.org/stable/pdf/167996.pdf) actually have the expected values that were estimated.
-Do we want to include an example of this table?
-Might be cool to compare to like the numbers we got from open WAR.)
+It is worth emphasizing that the values in $\mathbb{E}[X | s]$ are estimates, and the precision of those estimates has many subtleties.
+
+First, the values within the expected run matrix change over time. 
+For example, any estimation of the values in the expected run matrix based on data from a high-scoring era (e.g., the early 2000s) will yield different values than equivalent analysis in a low-scoring era.
+In a high run-scoring environment, where there are many home runs, the value of a walk may be higher, since a player who walks is more likely to score on a subsequent home run. 
+Conversely, in a low run-scoring environment, stolen bases and sacrifice bunts may be comparatively more valuable. 
+Thus, a careful estimate of $\mathbb{E}[X | s]$ would include a time parameter $t$, indicating when the estimate is appropriate.
+
+Second, the characterization of $S$ as having 24 states is only the simplest possible. 
+The inning, or the score of the game, or even the weather, might reasonably be incorporated into $S$, as those conditions might reasonably affect the estimate of $\mathbb{E}[X | s]$. 
+More definitively, the identity of the current batter, pitcher, or batter on deck, might also affect the estimate of $\mathbb{E}[X | s]$. 
+Indeed, @Tango2007book show that when a particularly weak-hitting batter is up (i.e., the pitcher), a sacrifice bunt becomes much more effective strategy.
 
@@
 
-However, this concept is easily extended to other sports. 
+## In other sports
+
+The concept of estimating the value of the state of a game is easily extended to other sports. 
 In American football, $s$ is determined by down, yardage to the next first down, and field position.

### 2022-06-01 sports_analytics.Rmd
BTM

@@
 For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$.
-Then the BTM is a logistic regression model with parameters $\{\beta_i\}$ such that
+Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
 
@@
 
-Here, $exp(\beta_i)$ is often viewed as a representation of player $i$'s ability. The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2].
+Here, $\exp{\beta_i}$ is often viewed as a representation of player $i$'s ability. The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2].
+
+For example, ...
+
+```{r btm, echo=TRUE}
+library(Lahman)
+library(BradleyTerry2)
+# show BTM for Pirates vs. Yankees in 1960
+# or Red Sox vs. Yankees in 2003/2004
+```
 

### 2022-06-01 sports_analytics.Rmd
added Albert reference
added Marchi reference
added sections

@@
 
-See [blah] for examples of analytical findings that are based on the expected run matrix.
+See @Albert2001curve for a fuller discussion of the use of the expected run matrix in baseball and @marchi2018analyzing for examples of how to estimate the expected run matrix using Retrosheet data.
 
-## In other sports
+## In other sports with discrete states
 
@@
 This is wild that they did this in 1971!
+
+## Sports with continuous states
 

### 2022-06-01 sports_analytics.Rmd
added Handbook references
added BTM computations

@@
 Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy. 
+@goldner2017situational uses a similar framework.
 
@@
 Even in sports where the concept of a state is not so easy to define, like basketball, the value of a possession can be estimated with player tracking data [@cervone2016multiresolution].
+
+See also @bornn2017studying.
 
@@
 
+@goldner2017situational
+
 ## Basketball
@@
 ```{r btm, echo=TRUE}
-library(Lahman)
+library(tidyverse)
 library(BradleyTerry2)
@@
 # or Red Sox vs. Yankees in 2003/2004
+library(retrosheet)
+games <- getRetrosheet("game", 2004)
+
+bt_games <- games |>
+  group_by(HmTm, VisTm) |>
+  summarize(
+    home_w = sum(HmRuns > VisRuns),
+    away_w = sum(VisRuns > HmRuns),
+  ) |>
+  mutate(
+    HmTm = factor(HmTm),
+    VisTm = factor(VisTm, levels = levels(HmTm))
+  )
+
+btm <- BTm(
+  outcome = cbind(home_w, away_w), 
+  player1 = HmTm, 
+  player2 = VisTm,
+  data = bt_games, 
+  id = "team"
+)
+btm
+
+bt_games2 <- bt_games |>
+  mutate(
+    HmTm = data.frame(team = HmTm, at_home = 1),
+    VisTm = data.frame(team = VisTm, at_home = 0)
+  )
+
+btm2 <- btm |>
+  update(formula = ~ team + at_home, data = bt_games2)
+btm2
+```
+
+```{r}
+btm |>
+  broom::tidy() |>
+  arrange(desc(estimate))
+btm2 |>
+  broom::tidy() |>
+  arrange(desc(estimate))
 ```
@@
 
-@Glickman1998 state-space model
+@Glickman1998 state-space model. 
+See also @glickman2017estimating.
 
 @Lopez2018 extension to @Glickman1998's framework
+
+In soccer, @koning2017rating
 

### 2022-06-06 sports_analytics.Rmd
added section on probability of scoring one run

@@
 output: rticles::arxiv_article
+header-includes:
+  - \usepackage{amsmath}
 ---
@@
   c(
+    "base",
     "Lahman", "pitchRx", "retrosheet",
@@
 
-- [CRAN Task View SportsAnalytics](https://github.com/beanumber/ctv-sportsanalytics)
+See @cochran2017oxford and @Albert2017handbook for overview readers.
 
-# The expected value of a possession
+See @baumer2022ctv for R packages useful for sports analytics.
+
+# The expected value of a possession {#sec:ev}
 
@@
 
-See @Albert2001curve for a fuller discussion of the use of the expected run matrix in baseball and @marchi2018analyzing for examples of how to estimate the expected run matrix using Retrosheet data.
+See @Albert2001curve for a fuller discussion of the use of the expected run matrix in baseball and @marchi2018analyzing for examples of how to estimate the expected run matrix using Retrosheet data and the R statistical computing language [@R-base].
 
@@
 @decroos2019playeraction Actions Speak Louder than Goals: Valuing Player Actions in Soccer
+
+## Optimal strategies that don't maximize expected points
+
+In Section \ref{sec:ev}, we defined the expected value of a possession based on the state $s$ of the game in terms of the expected number of points (runs) $X$ that would be scored in the remainder of some period of time.
+We then showed how this value could be used to analyze the relative effectiveness of certain strategies, with the simple idea that strategies that yield higher expected values are preferable. 
+Generally, the goal of any sport is to score more points than the other team, which most often means trying to score as many points as possible, leading to a general strategy of maximizing expected points.
+However, there are situations in which maximizing the number of expected points is *not* the desired strategy.
+
+For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs.
+Mathematically, we distinguish between 
+$$
+  \max{\Pr[X > 0 | s]} \, \text{ and } \, \max{\mathbb{E}[X | s]} \,.
+$$
+Maximizing the probability of scoring is optimal in any sudden-death situation, which has (but currently does not) include overtime in American football [@martin2018markov].
+
+The situation gets even more interesting when teams modify both their offensive and defensive strategies simultaneously. 
+For example, in hockey teams will often pull their goalie when trailing in the final period. 
+This strategy severely weakens their defense, but strengthens their offense. 
+The hope is to score a quick goal to get back in the game, but the risk is falling further behind. 
+@beaudoin2010strategies show that NHL coaches do not always employ the optimal strategies. 
+@skinner2011scoring develops a general framework for these desperation strategies, which include the onside kick in American football, pulling the infield and/or outfield in in baseball, and of course, the fabled Hack-a-Shaq strategy in basketball.
 

### 2022-06-07 sports_analytics.Rmd
added more math

@@
 
-# The expected value of a possession {#sec:ev}
+# The expected value of a state {#sec:ev}
 
@@
 
-Mathematically, we define $X$ to be a random variable indicating the the value of a particular game state for a team over some determined amount of time (e.g.
+Mathematically, we define $X$ to be a random variable indicating the the number of points (or runs) that a team will score over some determined amount of time (e.g.
 remainder of game, quarter, period, or inning).
@@
 $$
-  \mathbb{E}[X | s] \,,
+  \mathbb{E}[X | s] = \sum_{X \geq 0} \Pr[X | s] \cdot X\,,
 $$
-for any state $s \in S$, where $S$ is the set of all possible states.
+for any state $s \in S$, where $\Pr[X | s]$ is the probability of scoring $X$ points given that the game is in state $s$ and $S$ is the set of all possible states.
 
@@
 For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs.
-Mathematically, we distinguish between 
+If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy that maximizes the expected number of points will maximize the probability of winning:
 $$
-  \max{\Pr[X > 0 | s]} \, \text{ and } \, \max{\mathbb{E}[X | s]} \,.
+  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s]} \,.
 $$
-Maximizing the probability of scoring is optimal in any sudden-death situation, which has (but currently does not) include overtime in American football [@martin2018markov].
+Maximizing the probability of scoring is optimal in any sudden-death situation, which has (but currently does not) included overtime in American football [@martin2018markov].
 
@@
 
-A related, but different, concept is the notion of *win probability*, which is simply an estimate of the probability that either team will win the game, given its current state.
-Win probability provides a measure of the impact of sports plays and individual player contributions, as discussed in @Albert2015.
+A related, but different, concept is the notion of *win probability*.
+Win probability is simply an estimate of the probability that a team will win the game, given its current state $s$. 
+Extending the mathematical framework we defined in Section \ref{sec:ev}, let $W$ be a binary random variable that indicates a win for a given team. 
+Then, 
+$$
+  \Pr[W | s] \,,
+$$
+is the win probability in the state $s$.
+
+This win probability is closely related to the expected value of a state. 
+@Albert2015 defines the win probability as:
+$$
+  \Pr[W | s] = \sum_{X \geq 0} \Pr[X | s] \cdot \Pr[W | X, s] \,,
+$$
+where $\Pr[W | X, s]$ is the probability that the team will win the game given that they score $X$ points from state $s$.
+
+Win probability is easily extended to provide a measure of the impact of sports plays and individual player contributions, as discussed in @Albert2015. 
 Given its popularity, recent books on sports analytics often offers multiple chapters entirely on win probability. 

### 2022-06-07 sports_analytics.Rmd
added Lindsey ERM

@@
 The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
-Figure \ref{fig:lindsey} shows a reproduction of Lindsey's original expected run matrix.
+Figure \ref{fig:lindsey} shows a reproduction of Lindsey's original calculations and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
 
@@
 knitr::include_graphics("lindsey_table1.png")
+```
+
+```{r erm, echo=FALSE}
+lindsey <- c(
+  .461, .243, .102,
+  .813, .498, .219,
+  1.194, .671, .297,
+  1.39, .98, .355,
+  1.471, .939, .403,
+  1.94, 1.115, .532,
+  1.96, 1.56, .687,
+  2.22, 1.642, .823
+)
+erm <- tibble(
+  base = rep(0:7, each = 3),
+  out = rep(0:2, times = 8),
+  runs = lindsey
+)
+```
+
+```{r lindsey-erm}
+int2bin <- function(x) {
+  x |>
+    map(intToBits) |>
+    map(rev) |>
+    map(tail, 3) |>
+    map(str_sub, 2) |>
+    map_chr(paste, collapse = "")
+}
+#int2bin(0:7)
+
+erm |>
+  pivot_wider(names_from = out, values_from = runs) |>
+  mutate(base = int2bin(base)) |>
+  rename(`base\\\\out` = base) |>
+  knitr::kable(caption = "George Lindsey's expected run matrix.
+Note how (when reading across the rows) the expected runs decrease as outs increase for the same configuration of baserunners, while (when reading down the columns) expected runs generally increase as baserunners advance.
+\\label{tab:erm}")
 ```
@@
 $$
-log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j
+\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j
 $$
@@
 ```{r btm, echo=TRUE}
-library(tidyverse)
 library(BradleyTerry2)
@@
 library(retrosheet)
-games <- getRetrosheet("game", 2004)
+games <- get_retrosheet("game", 2004)
 

### 2022-06-07 sports_analytics.Rmd
added probability of not scoring

@@
 $$
+Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
+Table \ref{tab:tied} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the the probability of scoring zero runs is 0.13. 
+The defense is in a tight spot, facing an 87% probability of losing the game. 
+However, by walking the hitter to load the bases, they create the opportunity to force the lead runner at home and thus reduce the chances of scoring to 82%, even though they raise the expected number of runs scored to 2.22.
+In this case, the defensive team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.
+
+```{r}
+erm <- erm |>
+  mutate(
+    pr_0 = c(
+      .747, .855, .933,
+      .604, .734, .886, 
+      .381, .610, .788,
+      .120, .307, .738,
+      .395, .571, .791,
+      .130, .367, .717,
+      .180, .270, .668,
+      .180, .303, .671
+    )
+  )
+
+erm |>
+  filter(out == 0, base %in% c(5, 7)) |>
+  mutate(base = int2bin(base)) |>
+  knitr::kable(caption = "Loading the bases can be a rational strategy.
+In this case, loading the bases maximizes both the expected number of runs scored *and* the probability of not scoring at all.
+\\label{tab:tied}")
+```
+
 Maximizing the probability of scoring is optimal in any sudden-death situation, which has (but currently does not) included overtime in American football [@martin2018markov].

### 2022-06-14 sports_analytics.Rmd
a paragraph on nfl wp

@@
     "ggplot2",
-    "BradleyTerry2"
+    "BradleyTerry2",
+    "nflfastR",
+    "nflverse"
   ),
@@
 
-# Win probability
+# Win probability {#sec:wp}
 
@@
 
-In football,
+In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football.
+@Lock2014 used random forests as a means for estimating win probability before each play in a football game.
+In addition, @Yurko2019 estimated win probability in the NFL using a generalized additive model, as part of the `nflscrapr` package and nflwar framework.
+A major highlight of this framework is it is fully reproducible and uses publicly available data.
+One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work", which ultimately inspired \pkg{nflfastR}, known as the successor to `nflscrapr`
 
-Burke WPA
-
-http://www.advancedfootballanalytics.com/2010/01/win-probability-added-wpa-explained.html
-
-@Lock2014 used random forests as a means for estimating WP before each play in a football game.
-
-@Yurko2019
+(add a wp example from nflfastR)
 
 @Yam2019 causal
-
-nflscrapr, nflfastR, nflverse
 
@@
 @Maymin2020 esport!
+
+ann appl stat march 2022 paper on wp in rugby
+
+https://projecteuclid.org/journals/annals-of-applied-statistics/volume-16/issue-1/In-game-win-probabilities-for-the-National-Rugby-League/10.1214/21-AOAS1514.short
 

### 2022-06-18 sports_analytics.Rmd
wp in football now looks decent

@@
 In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football.
+These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require little assumptions, and produce feature importance scores.
+
 @Lock2014 used random forests as a means for estimating win probability before each play in a football game.
-In addition, @Yurko2019 estimated win probability in the NFL using a generalized additive model, as part of the `nflscrapr` package and nflwar framework.
+Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yardline, and yards to go for a first down.
+
+In addition, @Yurko2019 estimated win probability in the NFL using a generalized additive model (GAM), as part of the `nflscrapr` package and nflWAR framework.
+This model takes into account the estimated expected points obtained from the model described in \ref{sec:ev}, along with other predictors for time, current half, and timeouts.
 A major highlight of this framework is it is fully reproducible and uses publicly available data.
@@
 
-(add a wp example from nflfastR)
+These two frameworks were also implemented in @Yam2019 with minimal modifications.
+Specifically, the authors combined both @Lock2014 and @Yurko2019's approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision making in gridiron football.
 
-@Yam2019 causal
-
-@Yurko2020 going deep
-
-@goldner2017situational
+(add a wp example from nflfastR or nflscapr)
 

### 2022-06-20 sports_analytics.Rmd
some work on team strength

@@
 
-ann appl stat march 2022 paper on wp in rugby
-
-https://projecteuclid.org/journals/annals-of-applied-statistics/volume-16/issue-1/In-game-win-probabilities-for-the-National-Rugby-League/10.1214/21-AOAS1514.short
+@Guan2022 wp for rugby
 
@@
 
-@Elo1978
+Another widely known tool for measuring team strength is the Elo rating system [@Elo1978].
+The Elo rating was originally developed for chess, but also has been used to estimate team strength in other sports.
 
-@Glickman1998 state-space model. 
-See also @glickman2017estimating.
+See @koning2017rating for more information on applications of the Elo rating in soccer.
 
-@Lopez2018 extension to @Glickman1998's framework
+@Glickman1998 proposed a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time.
+In particular, this paper modeled point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths.
+See also @glickman2017estimating for more discussion on estimating team strengths in American football.
 
-In soccer, @koning2017rating
+@Lopez2018 extended @Glickman1998's state-space model to understand randomness in the four major American sports leagues.
+Betting moneylines were used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
 

### 2022-06-20 sports_analytics.Rmd
some notes on wp in basketball and other sports

@@
 
-@Stern1994 brownian
+@Stern1994 used Brownian motion to investigate the scoring process of basketball.
+The model introduced by @Stern1994 is a probit regression model, which provides an estimate for the probability of winning for a home team, taking into account the score difference between the teams and the current game time.
 
-@Deshpande2016 bayesian
+@Deshpande2016 extended @Stern1994's framework by applying it in a Bayesian setting.
 
-@McFarlane2019 nba logistic
+@McFarlane2019 used logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA
 
@@
 
-https://arxiv.org/pdf/2204.11777.pdf
+Examples of win probability can also be found in other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team.
 
-@Brenzel2019 curling markov
+@Brenzel2019 used a three-dimensional Markov model to estimate win probability in curling.
 
-@Maymin2020 esport!
+In esport, @Maymin2020 proposed a well-calibrated in-game win probability model for each specific moment during a game of League of Legends.
 
-@Guan2022 wp for rugby
+More recently, @Guan2022 developed an in-game win probability model for the National Rugby League using a functional data analysis approach.
 

### 2022-06-29 sports_analytics.Rmd
a few paragraphs on basketball wp

@@
 A major highlight of this framework is it is fully reproducible and uses publicly available data.
-One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work", which ultimately inspired \pkg{nflfastR}, known as the successor to `nflscrapr`
+One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work", which ultimately inspired \pkg{nflfastR}, known as the successor to `nflscrapr`.
 
@@
 @Stern1994 used Brownian motion to investigate the scoring process of basketball.
-The model introduced by @Stern1994 is a probit regression model, which provides an estimate for the probability of winning for a home team, taking into account the score difference between the teams and the current game time.
+Let $p(l, t)$ represents the win probability for the home team given an $l$-point lead after $t$ seconds of game time.
+The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$.
+In particular,
 
-@Deshpande2016 extended @Stern1994's framework by applying it in a Bayesian setting.
+$$
+p(l, t) = \Phi\left(\frac{l + (1-t)\mu}{\sqrt{(1-t)\sigma^2}}\right)
+,
+$$
 
-@McFarlane2019 used logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA
+Here, a Brownian motion process with drift $\mu$ points advantage for the home team and variance $\sigma^2$ is used to model the score difference between the home and away teams.
+
+On a related note, @Deshpande2016 extended @Stern1994's framework by applying it in a Bayesian setting.
+In particular, a Bayesian lasso regression model was proposed to assess the impact of individual players on their team's win probability at a given time of a basketball game.
+This model assumes independence of observation, constant variability in win probability, and also addresses multicollinearity.
+
+@McFarlane2019 used logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA.
+The approach takes into account the remaining game time, score difference, and point spread.
+This basketball win probability model is then applied to the calculation of the End-of-game Tactics Metric (ETM), measuring how the win probability differs between the optimal and on-court actual decisions.
 

### 2022-06-30 sports_analytics.Rmd
added wp plot using `nflfastR`

@@
 
-(add a wp example from nflfastR or nflscapr)
+Figure \ref{fig:nflfastr_wp}
+
+```{r}
+# win prob for Minneapolis Miracle
+library(teamcolors)
+
+vikings <- teamcolors |> 
+  filter(league == "nfl", name == "Minnesota Vikings") |> 
+  pull(primary)
+
+saints <- teamcolors |> 
+  filter(league == "nfl", name == "New Orleans Saints") |> 
+  pull(primary)
+```
+
+```{r nflfastr_wp, fig.cap="WP for MIN vs NO (add more later)"}
+library(nflfastR)
+load_pbp(2017) |> 
+  filter(season_type == "POST",
+         home_team == "MIN",
+         away_team == "NO",
+         !is.na(home_wp),
+         !is.na(away_wp)) |>
+  select(game_seconds_remaining, home_wp, away_wp) |> 
+  pivot_longer(!game_seconds_remaining,
+               names_to = "team",
+               values_to = "wp") |> 
+  ggplot(aes(game_seconds_remaining, wp, color = team)) +
+  geom_line(size = 2) +
+  scale_x_reverse(breaks = seq(0, 3600, 900)) + 
+  scale_color_manual(values = c(saints, vikings), 
+                     labels = c("New Orleans Saints", "Minnesota Vikings")) +
+  # geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
+  # geom_vline(xintercept = seq(0, 2700, 900), linetype = "dashed", color = "gray") + 
+  labs(x = "Time Remaining (seconds)",
+       y = "Win Probability",
+       color = NULL) +
+  theme_bw() +
+  theme(legend.position = "top")
+```
 

### 2022-06-30 sports_analytics.Rmd
team strength: BT example

--- 2022-06-29 sports_analytics.Rmd
@@
   message = FALSE,
+  warning = FALSE,
   fig.align = 'center'
@@
     "base",
-    "Lahman", "pitchRx", "retrosheet",
+    "Lahman", 
+    "pitchRx", 
+    "retrosheet",
     "teamcolors",
@@
 Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yardline, and yards to go for a first down.
-
 In addition, @Yurko2019 estimated win probability in the NFL using a generalized additive model (GAM), as part of the `nflscrapr` package and nflWAR framework.
 This model takes into account the estimated expected points obtained from the model described in \ref{sec:ev}, along with other predictors for time, current half, and timeouts.
-A major highlight of this framework is it is fully reproducible and uses publicly available data.
+The two win probability frameworks proposed by @Lock2014 and @Yurko2019 were also implemented in @Yam2019 with minimal modifications.
+Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision making in gridiron football.
+
+A vital highlight of @Yurko2019's win probability model is that it is fully reproducible and uses publicly available data.
 One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work", which ultimately inspired \pkg{nflfastR}, known as the successor to `nflscrapr`.
+Figure \ref{fig:nflfastr_wp} shows a win probability graph for the 2017-18 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings.
+The chance of winning for each team is obtained using the \pkg{nflfastR} R package, which implemented a gradient boosting model via the `xgboost` library for estimating win probabilities.
+Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, 0.742 at the end of the first, second, and third quarters, respectively.
+The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 seconds left in the game.
+The last play of this game, famously known as the Minneapolis Miracle, resulted in a drastic swing in win probabilities for both teams.
+With 10 seconds remaining of game time, the Vikings begin the final possession with a 25.3% chance of winning.
+Their probability increased to a perfect 1 when Stefon Diggs scored a game-winning 61-yard receiving touchdown as the game clock expired.
 
-These two frameworks were also implemented in @Yam2019 with minimal modifications.
-Specifically, the authors combined both @Lock2014 and @Yurko2019's approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision making in gridiron football.
+```{r}
+# win prob for Minneapolis Miracle
+library(teamcolors)
 
-(add a wp example from nflfastR or nflscapr)
+vikings <- teamcolors |> 
+  filter(league == "nfl", name == "Minnesota Vikings") |> 
+  pull(primary)
+
+saints <- teamcolors |> 
+  filter(league == "nfl", name == "New Orleans Saints") |> 
+  pull(primary)
+```
+
+```{r nflfastr_wp, fig.cap="Win probability graph for Saints vs Vikings in the 2017-18 NFL Playoffs."}
+library(nflfastR)
+load_pbp(2017) |> 
+  filter(season_type == "POST",
+         home_team == "MIN",
+         away_team == "NO",
+         !is.na(home_wp),
+         !is.na(away_wp)) |>
+  select(game_seconds_remaining, home_wp, away_wp) |> 
+  pivot_longer(!game_seconds_remaining,
+               names_to = "team",
+               values_to = "wp") |>
+  ggplot(aes(game_seconds_remaining, wp, color = team)) +
+  geom_line(size = 1.5) +
+  scale_x_reverse(breaks = seq(0, 3600, 900)) + 
+  scale_color_manual(values = c(saints, vikings), 
+                     labels = c("New Orleans Saints", "Minnesota Vikings")) +
+  # geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
+  # geom_vline(xintercept = seq(0, 2700, 900), linetype = "dashed", color = "gray") + 
+  labs(x = "Time Remaining (seconds)",
+       y = "Win Probability",
+       color = NULL) +
+  theme_bw() +
+  theme(legend.position = "top")
+```
 
@@
 In particular,
-
 $$
@@
 $$
-
 Here, a Brownian motion process with drift $\mu$ points advantage for the home team and variance $\sigma^2$ is used to model the score difference between the home and away teams.
@@
 
-Examples of win probability can also be found in other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team.
-
-@Brenzel2019 used a three-dimensional Markov model to estimate win probability in curling.
-
-In esport, @Maymin2020 proposed a well-calibrated in-game win probability model for each specific moment during a game of League of Legends.
-
-More recently, @Guan2022 developed an in-game win probability model for the National Rugby League using a functional data analysis approach.
+The idea of win probability is also applied to other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team.
+@Brenzel2019 used three-dimensional Markov models to estimate win probability throughout a curling match.
+In particular, the authors proposed both a homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game.
+In esports, @Maymin2020 used logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends.
+More recently, @Guan2022 developed an in-game win probability model for the National Rugby League using functional data analysis.
+In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.
 
@@
 
+Another important idea in sports analytics is the estimation of team strength.
 A popular method for estimating team strength in sports is through pairwise evaluations.
@@
 
-Here, $\exp{\beta_i}$ is often viewed as a representation of player $i$'s ability. The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2].
+Here, $\exp{\beta_i}$ is often viewed as a representation of player $i$'s ability.
 
-For example, ...
+The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2].
+For example, we consider the data given in @agresti2018introduction, page 247 on tennis results from 2014 to 2018 for five men's professional players Djokovic, Federer, Murray, Nadal, and Wawrinka.
+We fitted a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.
 
-```{r btm, echo=TRUE}
+```{r}
+tennis <- dplyr::tibble(
+  Djokovic = c(NA, 9, 14, 9, 4),
+  Federer = c(6, NA, 5, 5, 7),
+  Murray = c(3, 0, NA, 2, 2),
+  Nadal = c(2, 1, 4, NA, 4),
+  Wawrinka = c(3, 2, 2, 3, NA)
+) |>
+  t() |>
+  BradleyTerry2::countsToBinomial()
+
+tennis_bt <- BradleyTerry2::BTm(cbind(win1, win2), player1, player2, data = tennis)
+
+tennis_bt |>
+  update(refcat = "Wawrinka") |> 
+  BradleyTerry2::BTabilities() |> 
+  round(3) |> 
+  knitr::kable(caption = "Estimated abilities for each player, relative to Wawrinka, obtained from the fitted BTM. \\label{tab:tennisbtm}")
+```
+
+Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM.
+According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Murray, 5) Nadal.
+To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is
+$$
+\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768
+$$
+
+```{r btm, eval=FALSE}
 library(BradleyTerry2)
@@
 btm2
-```
 
-```{r}
 btm |>
@@
 The Elo rating was originally developed for chess, but also has been used to estimate team strength in other sports.
-
 See @koning2017rating for more information on applications of the Elo rating in soccer.
@@
 See also @glickman2017estimating for more discussion on estimating team strengths in American football.
-
-@Lopez2018 extended @Glickman1998's state-space model to understand randomness in the four major American sports leagues.
+More recently, @Lopez2018 extended @Glickman1998's state-space model to understand randomness in the four major American sports leagues.
 Betting moneylines were used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.

### 2022-06-30 sports_analytics.Rmd
started on epv for continuous sports

@@
 
-Even in sports where the concept of a state is not so easy to define, like basketball, the value of a possession can be estimated with player tracking data [@cervone2016multiresolution].
+Even in sports where the concept of a state is not so easy to define, like basketball, the value of a possession can be estimated with player tracking data.
+@cervone2016multiresolution introduces expected possession value (EPV), providing a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations.
 
@@
 
+@Yurko2020going football
+
 @fernandez2021epvsoccer A framework for the fine-grained evaluation of the instantaneous expected value of soccer possessions
 
-@decroos2019playeraction Actions Speak Louder than Goals: Valuing Player Actions in Soccer
+(These papers shared the same definitions of EPV)
 
@@
 Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM.
-According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Murray, 5) Nadal.
-To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is
+According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is
 $$

### 2022-07-07 sports_analytics.Rmd
outlined epv for sports with continuous states
edits after f3d76af; #5

@@
     "nflfastR",
-    "nflverse"
+    "nflscrapr",
+    "nflverse",
+    "xgboost"
   ),
@@
 Even in sports where the concept of a state is not so easy to define, like basketball, the value of a possession can be estimated with player tracking data.
+
+(a few sentences on tracking data)
+
 @cervone2016multiresolution introduces expected possession value (EPV), providing a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations.
@@
 Win probability is simply an estimate of the probability that a team will win the game, given its current state $s$. 
-Extending the mathematical framework we defined in Section \ref{sec:ev}, let $W$ be a binary random variable that indicates a win for a given team. 
+Extending the mathematical framework we defined in Section \ref{sec:ev}, let $W_i$ be a binary random variable that indicates a win for team $i$. 
 Then, 
 $$
-  \Pr[W | s] \,,
+  \Pr[W_i | s] \,,
 $$
-is the win probability in the state $s$.
+is the win probability for team $i$ in the state $s$.
 
@@
 $$
-  \Pr[W | s] = \sum_{X \geq 0} \Pr[X | s] \cdot \Pr[W | X, s] \,,
+  \Pr[W_i | s] = \sum_{X \geq 0} \Pr[X | s] \cdot \Pr[W_i | X, s] \,,
 $$
-where $\Pr[W | X, s]$ is the probability that the team will win the game given that they score $X$ points from state $s$.
+where $\Pr[W_i | X, s]$ is the probability that team $i$ will win the game given that they score $X$ points from state $s$.
 
 Win probability is easily extended to provide a measure of the impact of sports plays and individual player contributions, as discussed in @Albert2015. 
-Given its popularity, recent books on sports analytics often offers multiple chapters entirely on win probability. 
+Given its popularity, recent books on sports analytics often dedicate multiple chapters entirely to win probability. 
 These include @Albert2001curve, @Schwarz2005numbers, @Tango2007book, @Albert2017handbook, and @Winston2022.
+
 In this section, we discuss notable previous work on win probability in baseball, football, basketball, and several other sports.
@@
 
-@Albert2015 provided a historical overview of the use of win probability in baseball.
+@Albert2015 provides a historical overview of the use of win probability in baseball. 
 This notion goes back to at least as early as @Lindsey1961, where the expected win probability after each inning was calculated based on the distribution of runs scored in each inning. 
-@Mills1970 utilized win probability to introduce Player Win Average (PWA), a measure of a player's contribute to the game outcome.
-In particular, PWA was computed as
+@Mills1970 utilize win probability to introduce Player Win Average (PWA), a measure of a player's contribution to the game outcome. 
+In particular, PWA is computed as 
 $$
@@
 where the win and loss points represent how much the player positively and negatively impact their team's probability of winning after each play.
+In effect, the win points are the sum of the changes in $\Pr[W_i | s]$ from one state to the next.
 
@@
 In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football. 
-These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require little assumptions, and produce feature importance scores.
+These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require few assumptions, and produce feature importance scores.
 
-@Lock2014 used random forests as a means for estimating win probability before each play in a football game.
+@Lock2014 use random forests as a means for estimating win probability before each play in a football game. 
 Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yardline, and yards to go for a first down.
-In addition, @Yurko2019 estimated win probability in the NFL using a generalized additive model (GAM), as part of the `nflscrapr` package and nflWAR framework.
-This model takes into account the estimated expected points obtained from the model described in \ref{sec:ev}, along with other predictors for time, current half, and timeouts.
+
+In addition, @Yurko2019 estimate win probability in the NFL using a generalized additive model (GAM), as part of the \pkg{nflscrapr} package (@R-nflscrapr) and nflWAR framework. 
+This model takes into account the estimated expected points obtained from the model described in Section \ref{sec:ev}, along with other predictors for time, current half, and timeouts. 
 The two win probability frameworks proposed by @Lock2014 and @Yurko2019 were also implemented in @Yam2019 with minimal modifications.
-Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision making in gridiron football.
+Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in gridiron football.
 
 A vital highlight of @Yurko2019's win probability model is that it is fully reproducible and uses publicly available data. 
-One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work", which ultimately inspired \pkg{nflfastR}, known as the successor to `nflscrapr`.
-Figure \ref{fig:nflfastr_wp} shows a win probability graph for the 2017-18 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings.
-The chance of winning for each team is obtained using the \pkg{nflfastR} R package, which implemented a gradient boosting model via the `xgboost` library for estimating win probabilities.
+One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work," which ultimately inspired \pkg{nflfastR} (@R-nflfastR), now considered the successor to \pkg{nflscrapr}.
+
+Figure \ref{fig:nflfastr_wp} shows a win probability graph for the January 14, 2018 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. 
+We obtain the estimated probability of winning for each team using the \pkg{nflfastR} R package, which implements a gradient boosting model via the \pkg{xgboost} library (@R-xgboost) for estimating win probabilities.
 Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, 0.742 at the end of the first, second, and third quarters, respectively. 
-The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 seconds left in the game.
-The last play of this game, famously known as the Minneapolis Miracle, resulted in a drastic swing in win probabilities for both teams.
-With 10 seconds remaining of game time, the Vikings begin the final possession with a 25.3% chance of winning.
+The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 left in the game. 
+The last play of this game, famously known as the [Minneapolis Miracle](https://en.wikipedia.org/wiki/Minneapolis_Miracle), resulted in a drastic swing in win probabilities for both teams. 
+With 10 seconds remaining in the game, the Vikings begin the final possession with a 25.3% chance of winning. 
 Their probability increased to a perfect 1 when Stefon Diggs scored a game-winning 61-yard receiving touchdown as the game clock expired.
@@
 
-```{r nflfastr_wp, fig.cap="Win probability graph for Saints vs Vikings in the 2017-18 NFL Playoffs."}
+```{r nflfastr_wp, fig.cap="Win probability graph for Saints vs. Vikings in the 2017--18 NFL Playoffs."}
 library(nflfastR)
-load_pbp(2017) |> 
-  filter(season_type == "POST",
+x <- load_pbp(2017) |> 
+  filter(
+    season_type == "POST",
     home_team == "MIN",
@@
     !is.na(home_wp),
-         !is.na(away_wp)) |>
+    !is.na(away_wp)
+  )
+x |>
   select(game_seconds_remaining, home_wp, away_wp) |> 
-  pivot_longer(!game_seconds_remaining,
+  pivot_longer(
+    !game_seconds_remaining,
     names_to = "team",
-               values_to = "wp") |>
+    values_to = "wp"
+  ) |>
   ggplot(aes(game_seconds_remaining, wp, color = team)) +
-  geom_line(size = 1.5) +
   scale_x_reverse(breaks = seq(0, 3600, 900)) + 
-  scale_color_manual(values = c(saints, vikings), 
-                     labels = c("New Orleans Saints", "Minnesota Vikings")) +
-  # geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
-  # geom_vline(xintercept = seq(0, 2700, 900), linetype = "dashed", color = "gray") + 
-  labs(x = "Time Remaining (seconds)",
+  scale_color_manual(
+    values = c(saints, vikings), 
+    labels = c("New Orleans Saints", "Minnesota Vikings")
+  ) +
+  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
+  geom_vline(
+    xintercept = seq(0, 3600, 900), 
+    linetype = "dashed", 
+    color = "gray"
+  ) + 
+  geom_line(size = 1.5, alpha = 0.8) +
+  labs(
+    x = "Time Remaining (seconds)",
     y = "Win Probability",
-       color = NULL) +
+    color = NULL
+  ) +
   theme_bw() +
@@
 
-@Stern1994 used Brownian motion to investigate the scoring process of basketball.
-Let $p(l, t)$ represents the win probability for the home team given an $l$-point lead after $t$ seconds of game time.
+@Stern1994 uses Brownian motion to investigate the scoring process of basketball. 
+Let $p(l, t)$ represent the win probability for the home team given an $l$-point lead after $t$ seconds of game time. 
 The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$. 
@@
 
-On a related note, @Deshpande2016 extended @Stern1994's framework by applying it in a Bayesian setting.
-In particular, a Bayesian lasso regression model was proposed to assess the impact of individual players on their team's win probability at a given time of a basketball game.
+On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
+In particular, they propose a Bayesian lasso regression model to assess the impact of individual players on their team's win probability at a given time of a basketball game. 
 This model assumes independence of observation, constant variability in win probability, and also addresses multicollinearity.
 
-@McFarlane2019 used logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA.
+@McFarlane2019 uses logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA. 
 The approach takes into account the remaining game time, score difference, and point spread. 
@@
 The idea of win probability is also applied to other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team. 
-@Brenzel2019 used three-dimensional Markov models to estimate win probability throughout a curling match.
-In particular, the authors proposed both a homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game.
-In esports, @Maymin2020 used logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends.
-More recently, @Guan2022 developed an in-game win probability model for the National Rugby League using functional data analysis.
+@Brenzel2019 uses three-dimensional Markov models to estimate win probability throughout a curling match. 
+In particular, the authors propose both a homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game.
+In esports, @Maymin2020 uses logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
+More recently, @Guan2022 develops an in-game win probability model for the National Rugby League using functional data analysis. 
 In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.
@@
 The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. 
-For example, we consider the data given in @agresti2018introduction, page 247 on tennis results from 2014 to 2018 for five men's professional players Djokovic, Federer, Murray, Nadal, and Wawrinka.
-We fitted a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.
+As an example, we consider the data given in @agresti2018introduction (page 247) on tennis results from 2014--2018 for five men's professional players Novak Djokovic, Roger Federer, Andy Murray, Rafael Nadal, and Stan Wawrinka.
+We fit a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.
 
@@
 Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM. 
-According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is
+According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray.
+To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is:
 $$
@@
 
-Another widely known tool for measuring team strength is the Elo rating system [@Elo1978].
-The Elo rating was originally developed for chess, but also has been used to estimate team strength in other sports.
+Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
 See @koning2017rating for more information on applications of the Elo rating in soccer.
@@
 @Glickman1998 proposed a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
-In particular, this paper modeled point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths.
+In particular, they modeled point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths. 
 See also @glickman2017estimating for more discussion on estimating team strengths in American football. 

### 2022-07-08 sports_analytics.Rmd
edits, style, bib fix

@@
   - name: Quang Nguyen
-    department: Mathematics and Statistics
-    affiliation: Loyola University Chicago
-    location: Chicago, IL 60660
-    email: nnguyen22@luc.edu
+    department: Statistics & Data Science
+    affiliation: Carnegie Mellon University
+    location: Pittsburgh, PA 15213
+    email: nmquang@cmu.edu
   - name: Katherine Evans
@@
 bibliography: [refs.bib, pkgs.bib]
-biblio-style: unsrt   # use apa for WIRE
+biblio-style: unsrt   
+# use apa for WIRE
+csl: apa.csl
 output: rticles::arxiv_article
@@
     "nflfastR",
-    "nflscrapr",
+    # "nflscrapr", no longer supported, can no longer be installed
     "nflverse",
@@
 
-In addition, @Yurko2019 estimate win probability in the NFL using a generalized additive model (GAM), as part of the \pkg{nflscrapr} package (@R-nflscrapr) and nflWAR framework. 
+In addition, @Yurko2019 estimate win probability in the NFL using a generalized additive model (GAM), as part of the **nflscrapR** package [@R-nflscrapR] and nflWAR framework. 
 This model takes into account the estimated expected points obtained from the model described in Section \ref{sec:ev}, along with other predictors for time, current half, and timeouts. 
@@
 A vital highlight of @Yurko2019's win probability model is that it is fully reproducible and uses publicly available data. 
-One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work," which ultimately inspired \pkg{nflfastR} (@R-nflfastR), now considered the successor to \pkg{nflscrapr}.
+One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work," which ultimately inspired \pkg{nflfastR} (@R-nflfastR), now considered the successor to **nflscrapR**.
 
@@
 
+Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM. 
+According to the abilities, between 2014 and 2018 the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray.
+To obtain win probabilities, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is:
+$$
+\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768
+$$
+
 ```{r}
@@
   round(3) |> 
-  knitr::kable(caption = "Estimated abilities for each player, relative to Wawrinka, obtained from the fitted BTM. \\label{tab:tennisbtm}")
+  knitr::kable(caption = "Estimated abilities for each player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}")
 ```
-
-Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM. 
-According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray.
-To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is:
-$$
-\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768
-$$
 
@@
 
-@Glickman1998 proposed a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
-In particular, they modeled point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths. 
+@Glickman1998 propose a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
+In particular, they model point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths. 
 See also @glickman2017estimating for more discussion on estimating team strengths in American football. 
-More recently, @Lopez2018 extended @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
-Betting moneylines were used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
+More recently, @Lopez2018 extend @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
+Betting moneylines are used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
 
@@
 ```
-
-# Color palettes
 

### 2022-07-10 sports_analytics.Rmd
a few paragraphs on ep for continuous sports

@@
 
-Even in sports where the concept of a state is not so easy to define, like basketball, the value of a possession can be estimated with player tracking data.
+Even in sports where the concept of a state is not so easy to define, the value of a possession can be estimated, with the help of tracking data.
+Over the past decade or so, professional sports leagues have collected tracking data which record the locations of all players and the ball (or puck) throughout a game.
+This sophisticated type of data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game, which is much more difficult to be accomplished with pre-existing resources such as box-score results and play-by-play data.
 
-(a few sentences on tracking data)
+In basketball, @cervone2014pointwise and @cervone2016multiresolution introduces expected possession value (EPV) as an assessment of a player's on-court performance.
+This metric is a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations.
+The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision.
+See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
 
-@cervone2016multiresolution introduces expected possession value (EPV), providing a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations.
+more basketball @Sicilia2019DeepHoops
 
-See also @bornn2017studying.
-
-@Yurko2020going football
+@Yurko2020going uses tracking data provided by the 2019 NFL Big Data Bowl (citation) to model the expected yards gained for a ball-carrier during the course of a play.
+As opposed to only giving a single estimate for the expected yards gained, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play.
 
 @fernandez2021epvsoccer A framework for the fine-grained evaluation of the instantaneous expected value of soccer possessions
-
-(These papers shared the same definitions of EPV)
 

### 2022-07-10 sports_analytics.Rmd
more writing on sports with continuous states

@@
 
-In basketball, @cervone2014pointwise and @cervone2016multiresolution introduces expected possession value (EPV) as an assessment of a player's on-court performance.
+In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as an assessment of a player's on-court performance.
 This metric is a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations.
 The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision.
+
+Another framework for estimating expected points in basketball was proposed by @Sicilia2019DeepHoops.
+The authors first consider a classification model which returns the probabilities for whether a player would commit a foul (shooting and non-shooting), turnover, or attempt a shot.
+The values associated with each of those "terminal actions" are then utilized to compute the expected points within a basketball play.
+
 See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
 
-more basketball @Sicilia2019DeepHoops
+In American football, @Yurko2020going use tracking data provided by the 2019 NFL Big Data Bowl (citation) to model the expected yards gained for a ball-carrier during the course of a play.
+As an extension to pre-existing approaches, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play, rather than only producing a single estimate for the expected yards gained.
+Accordingly, the probability of various types of outcome at the end of a play such as a touchdown or a first-down gain can be computed from the yards gained distribution.
 
-@Yurko2020going uses tracking data provided by the 2019 NFL Big Data Bowl (citation) to model the expected yards gained for a ball-carrier during the course of a play.
-As opposed to only giving a single estimate for the expected yards gained, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play.
-
-@fernandez2021epvsoccer A framework for the fine-grained evaluation of the instantaneous expected value of soccer possessions
+The notion of EPV has also been extended to soccer recently.
+@fernandez2021framework uses deep learning methods to examine the  instantaneous expected value of soccer possessions.
+This approach considers passes, ball drives, and shots in soccer as the main set of actions used to compute expected possession value.
 

### 2022-07-11 sports_analytics.Rmd
some notes on discrete sports

@@
 
-Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy. 
-@goldner2017situational uses a similar framework.
-
-@Pattani2012 describes ESPN's expected points metric.
-
-@CarrollPalmer1988 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.
-
-@Yurko2019, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM).
-
 @CarterMachol1971 estimate the expected points for 1st and 10 given a yard line (broken up into 10 yard chunks centered at a given yardline (e.g.
@@
 
+Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy. 
+@goldner2017situational uses a similar framework.
+
+@CarrollPalmer1988 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.
+
+@Yurko2019, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM)
+
+multinomial logistic regression
+next score r.v.
+
 ## Sports with continuous states
@@
 Over the past decade or so, professional sports leagues have collected tracking data which record the locations of all players and the ball (or puck) throughout a game. 
-This sophisticated type of data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game, which is much more difficult to be accomplished with pre-existing resources such as box-score results and play-by-play data.
+This sophisticated type of data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game, which is much more difficult to be accomplished with pre-existing resources such as traditional box-score results and play-by-play data.
 
@@
 The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision. 
+Consequently, the authors derive a metric called EPV-Added (EPVA), measuring a player's EPV contribution in a given situation relative to a league-average player.
+
+(include code demo? https://github.com/dcervone/EPVDemo)
 

### 2022-07-11 sports_analytics.Rmd
several paragraphs on sports with discrete states

@@
 The concept of estimating the value of the state of a game is easily extended to other sports. 
-In American football, $s$ is determined by down, yardage to the next first down, and field position.
+For example, in American football, $s$ is determined by factors such as down, yardage to the next first down, and field position.
 
-@CarterMachol1971 estimate the expected points for 1st and 10 given a yard line (broken up into 10 yard chunks centered at a given yardline (e.g.
-5, 15, 25, 35, etc.
-)).
-This is wild that they did this in 1971!
+The first effort of estimating expected points of possession in football goes back to @Carter1971operations. 
+In this technical note, the authors estimate the expected points for 1st and 10 plays in the NFL given any yard line on the football field. 
+Due to limitations regarding the amount of data collected, the football field is broken into 10-yard strips, centered at their midpoints (e.g.
+5, 15, 25, 35, etc.), and the expected points are calculated for each of those field positions.
+Early work on expected point values in football also includes @Carroll1988hidden, which popularizing the idea of play evaluation using expected points and expected points added.
+
+@goldner2012markov and @goldner2017situational proposes a Markov model for estimating expected points in football. 
+In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
+For any given play, the expected points is calculated using the absorption probabilities for different scoring events.
+
+A more in-depth overview of the history of expected points is provided in @Yurko2019nflwar (Section 1.1). 
+Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football. 
+The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes.
+This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.
+The expected points for a play can then be calculated accordingly, by taking the sum of the products of the scoring event point values and their associated probabilities.
+
+(A paragraph on applications of expected points on decision making, fourth down strategy, with an example)
 
 Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.
-@goldner2017situational uses a similar framework.
-
-@CarrollPalmer1988 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.
-
-@Yurko2019, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM)
-
-multinomial logistic regression
-next score r.v.
 
@@
 The notion of EPV has also been extended to soccer recently.
-@fernandez2021framework uses deep learning methods to examine the instantaneous expected value of soccer possessions. 
+@fernandez2021framework implement deep learning methods to examine the instantaneous expected value of soccer possessions. 
 This approach considers passes, ball drives, and shots in soccer as the main set of actions used to compute expected possession value.
@@
 
-In addition, @Yurko2019 estimate win probability in the NFL using a generalized additive model (GAM), as part of the **nflscrapR** package [@R-nflscrapR] and nflWAR framework. 
+In addition, @Yurko2019nflwar estimate win probability in the NFL using a generalized additive model (GAM), as part of the **nflscrapR** package [@R-nflscrapR] and nflWAR framework. 
 This model takes into account the estimated expected points obtained from the model described in Section \ref{sec:ev}, along with other predictors for time, current half, and timeouts. 
-The two win probability frameworks proposed by @Lock2014 and @Yurko2019 were also implemented in @Yam2019 with minimal modifications.
+The two win probability frameworks proposed by @Lock2014 and @Yurko2019nflwar were also implemented in @Yam2019 with minimal modifications.
 Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in gridiron football.
 
-A vital highlight of @Yurko2019's win probability model is that it is fully reproducible and uses publicly available data. 
-One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work," which ultimately inspired \pkg{nflfastR} (@R-nflfastR), now considered the successor to **nflscrapR**.
+A vital highlight of @Yurko2019nflwar's win probability model is that it is fully reproducible and uses publicly available data. 
+One of @Yurko2019nflwar's goals was also to encourage researchers to "use, explore, and improve upon our work," which ultimately inspired \pkg{nflfastR} (@R-nflfastR), now considered the successor to **nflscrapR**.
 
@@
 
-Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
+Another widely known tool for measuring team strength is the Elo rating system [@Elo1978].
+The Elo rating was originally developed for chess, but also has been used to estimate team strength in other sports.
 See @koning2017rating for more information on applications of the Elo rating in soccer.

### 2022-07-13 sports_analytics.Rmd
forgot to push these

@@
 Another framework for estimating expected points in basketball was proposed by @Sicilia2019DeepHoops. 
-The authors first consider a classification model which returns the probabilities for whether a player would commit a foul (shooting and non-shooting), turnover, or attempt a shot. 
+The authors offer a different point of view on expected points, where they first consider a classification model which returns the probabilities for whether a player would commit a foul (shooting and non-shooting), turnover, or attempt a shot.
 The values associated with each of those "terminal actions" are then utilized to compute the expected points within a basketball play.

### 2022-07-13 sports_analytics.Rmd
Small updates.

@@
 The concept of estimating the value of the state of a game is easily extended to other sports.
-For example, in American football, $s$ is determined by factors such as down, yardage to the next first down, and field position.
+For example, in American football, $s$ is determined by factors such as down, yardage to the next first down, time remaining in the game, and field position.
 
@@
 5, 15, 25, 35, etc.), and the expected points are calculated for each of those field positions.
-Early work on expected point values in football also includes @Carroll1988hidden, which popularizing the idea of play evaluation using expected points and expected points added.
+Other early work on expected point values in American football can be found in @Carroll1988hidden, which popularized the idea of play evaluation using expected points and expected points added.
 
@@
 Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
+
 The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes.
+
 This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.
@@
 (A paragraph on applications of expected points on decision making, fourth down strategy, with an example)
+
+4th down strategy papers based on expected points.  
+@Yam2019: 
+@Romer2006: Examines 4th down decisions in the NFL using expected points.
+Focuses only on the first quarter.
+Teams don't go for it enough is they are trying to maximize their probability of winning the game.
+@pelechrinis2019 Present an expected points framework where they adjust for the strength of the defence.  
+<!-- New york times article: https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html?_r=0 -->
 

### 2022-07-13 sports_analytics.Rmd
Small updates.

@@
 Teams don't go for it enough is they are trying to maximize their probability of winning the game.
-@pelechrinis2019 Present an expected points framework where they adjust for the strength of the defence.  
+@pelechrinis2019 Present an expected points framework where they adjust for the strength of the defense.  
 <!-- New york times article: https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html?_r=0 -->
+
+@Alamar2010 Evaluation of play calling in the NFL based on expected points.  
+https://www.degruyter.com/document/doi/10.2202/1559-0410.1235/html
+
+@urschel2011 Examines decsions on kick offs (surprise on sides or regular kick off AND accept touchback or return kick) using an expected points frame work.
+Shows that teams are moderately risk and loss averse.
+
+@White2002 Evaluated quaraterback usein gtiered multinomial regression in an expected points framework. They use a simple expected points model with down, to go, and to goal as predictors.   
+https://www.eng.buffalo.edu/~jzhuang/Papers/JQAS_NFL_2011.pdf
 

### 2022-07-14 sports_analytics.Rmd
added logos to Figure 2; #10

@@
   - sports data
+  - pairwise comparisons
   - datasets
@@
 
+<!--
+
 > Charge: Given your expertise and background, a review article from your group related to the topic of Statistical tools for sports analytics (e.g.
@@
 
-## Graphical/Visual Abstract and Caption
-
-> Include an attractive full color image to go under the text abstract and in the online Table of Contents.
-You will also need to upload this as a separate file during submission.
-It may be a figure or panel from the article or may be specifically designed as a visual summary.
-While original images are preferred, if you need to look for a thematically appropriate stock image, you can go to pixabay.com (not affiliated with Wiley) to find a free stock image with a CC0 license.
-Another option you have is to utilize professional illustrators with Wiley’s Graphical Abstract Design service.
-
-> Size: The minimum resolution is 300 dpi.
-Please keep the image as simple as possible because it will be displayed in multiple sizes.
-Multiple panels and text other than labels are strongly discouraged.
-
-> Caption: This is a narrative sentence to convey the article's essence and wider implications to a non-specialist audience.
-The maximum length is 50 words, but consider using 280 characters or less to facilitate social media sharing, which can increase the discoverability of your article.
+-->
 
@@
 
-vikings <- teamcolors |> 
-  filter(league == "nfl", name == "Minnesota Vikings") |> 
-  pull(primary)
-
-saints <- teamcolors |> 
-  filter(league == "nfl", name == "New Orleans Saints") |> 
-  pull(primary)
+miracle <- teamcolors |> 
+  filter(league == "nfl", grepl("Vikings|Saints", name)) |>
+  arrange()
 ```
 
-```{r nflfastr_wp, fig.cap="Win probability graph for Saints vs. Vikings in the 2017--18 NFL Playoffs."}
+```{r nflfastr_wp, fig.cap="Win probability graph for New Orleans Saints vs. Minnesota Vikings in the 2017--18 NFL Playoffs."}
 library(nflfastR)
@@
     !is.na(away_wp)
-  )
-x |>
+  ) |>
   select(game_seconds_remaining, home_wp, away_wp) |> 
@@
     values_to = "wp"
-  ) |>
-  ggplot(aes(game_seconds_remaining, wp, color = team)) +
-  scale_x_reverse(breaks = seq(0, 3600, 900)) + 
+  )
+
+wp_plot <- ggplot(x, aes(game_seconds_remaining, wp, color = team)) +
   scale_color_manual(
-    values = c(saints, vikings), 
-    labels = c("New Orleans Saints", "Minnesota Vikings")
+    values = setNames(miracle$primary, c("home_wp", "away_wp")), 
+    labels = miracle$name
   ) +
@@
   theme(legend.position = "top")
+
+library(magick)
+logos <- miracle |>
+  pull(logo) |>
+  map(image_read) |>
+  map(as.raster)
+
+wp_plot + 
+  annotation_raster(logos[[1]], -1010, -1490, 0.76, 0.865) + 
+  annotation_raster(logos[[2]], -1010, -1490, 0.135, 0.24) + 
+  scale_x_reverse()
 ```
@@
 @R-teamcolors provides color palettes for professional and amateur sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
+The team colors and logos shown in Figure \ref{fig:nflfastr_wp} were provided by the \pkg{teamcolors} package. 
 Figure \ref{fig:teamcolors} illustrates team color palettes for professional sports leagues.

### 2022-07-14 sports_analytics.Rmd
drafted introduction; #11

@@
   - name: Benjamin S. Baumer
-    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
+#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
     department: Statistical & Data Sciences
@@
 
+<!--
+
 > Introduce your topic in ~2 paragraphs, ~750 words.
@@
 
-See @cochran2017oxford and @Albert2017handbook for overview readers.
+-->
 
-See @baumer2022ctv for R packages useful for sports analytics.
+Insights derived from the analysis of data have transformed the world of sports over the last few decades. 
+While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the deepest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact.
 
-# The expected value of a state {#sec:ev}
+In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
+Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot.
+[CITATION NEEDED?]
+In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative.
+[LOPEZ?]
+And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact.
+
+These are just the most obvious examples of strategic changes that are fueled by insights extracted from data by practitioners of sports analytics.
+Similar insights are now being made in less obvious settings, including esports [@clark2020bayesian; @maymin2021smart]. 
+These insights come both from academia, where researchers typically use public data to produce high-caliber, peer-reviewed scientific work, as well as from industry, where highly-trained analysts work with with players, coaches, and team officials to gain immediate effect thanks to high-resolution, often proprietary data.
+A growing pool of people move seamlessly between these two worlds, leading to the formation of partnerships and the cross-pollination of ideas.
+
+Every sport is different, with its own set of rules, strategies, methods of data collection, number of players, and the magnitude of the role of chance. 
+At the same time, many sports are similar, either because one evolved from the other, or the structure of the game shares certain attributes. 
+Sports that are closely related anthropologically (NEED EXAMPLE) may or may not share common applications of analytical methods. 
+Conversely, with just a few small tweaks, analytical metrics might work just as well across sports that are unrelated and quite different (e.g., an Elo rating could be equally valid for chess players and ice hockey teams).
+
+In this paper, we explore three key ideas that have widespread applicability across many sports: the expected value of a game state (Section \ref{sec:ev}), win probability (Section \ref{sec:wp}), and measures of team strength (Section \ref{sec:strength}).
+In each case, we define the concept mathematically, explain how it originated, and give examples of how it is applied in multiple sports. 
+Our goal is to unify the conceptual threads, while doing some justice to the customizations necessary to make a metric meaningful in a particular sport. 
+We include copious references to original works of scholarship.
+
+Doing the work of sports analytics requires computing with data. 
+While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible.
+
+We encourage readers to explore @cochran2017oxford and @Albert2017handbook for collections of articles in sports analytics that provide broad coverage of the field.
+
+# The expected value of a game state {#sec:ev}
 
@@
 
-# Team strength
+# Team strength {#sec:strength}
 
@@
 
-## Other stuff
+# Tools {#sec:tools}
+
+See @baumer2022ctv for R packages useful for sports analytics.
 

### 2022-07-14 sports_analytics.Rmd
editing and references

@@
     email: nmquang@cmu.edu
-  - name: Katherine Evans
-    affiliation: Monumental Basketball
-    location: Washington, DC 20003
-    email: kevans@monumentalsports.com
 abstract: |
@@
 In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
-Thus, there $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball (25, if you include the absorbing state of 3 outs), and we are often interested in the number of runs that will be scored from some state until the end of the inning.
+Thus, there $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
 In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
 The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
-Figure \ref{fig:lindsey} shows a reproduction of Lindsey's original calculations and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
+Figure \ref{fig:lindsey} shows a reproduction of George Lindsey's original calculations and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
+
+[^state]: 25, if you include the absorbing state of 3 outs.
 
@@
 More definitively, the identity of the current batter, pitcher, or batter on deck, might also affect the estimate of $\mathbb{E}[X | s]$. 
-Indeed, @Tango2007book show that when a particularly weak-hitting batter is up (i.e., the pitcher), a sacrifice bunt becomes much more effective strategy.
+Indeed, @Tango2007book show that when a particularly weak-hitting batter is up (i.e., the pitcher), a sacrifice bunt becomes a more effective strategy.
 
@@
 
-## In other sports with discrete states
+## In American football
 
@@
 
-@goldner2012markov and @goldner2017situational proposes a Markov model for estimating expected points in football.
-In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes.
+@goldner2012markov and @goldner2017situational propose a Markov model for estimating expected points in football. 
+In particular, the author consider a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
 For any given play, the expected points is calculated using the absorption probabilities for different scoring events.
@@
 The expected points for a play can then be calculated accordingly, by taking the sum of the products of the scoring event point values and their associated probabilities.
+
+### 4th down strategy
 
@@
 
-## Sports with continuous states
+## In sports with continuous states
 
-Even in sports where the concept of a state is not so easy to define, the value of a possession can be estimated, with the help of tracking data. 
+Even in sports where the concept of a state is not so easy to define, the value of a possession can be estimated with the help of tracking data. 
 Over the past decade or so, professional sports leagues have collected tracking data which record the locations of all players and the ball (or puck) throughout a game. 
-This sophisticated type of data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game, which is much more difficult to be accomplished with pre-existing resources such as traditional box-score results and play-by-play data.
+This high-resolution data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game. 
+This is a great leap forward from older resources such as traditional box-score results and play-by-play data.
 
-In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as an assessment of a player's on-court performance. 
+In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as means toward an assessment of a player's on-court performance. 
 This metric is a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations. 
@@
 (include code demo? https://github.com/dcervone/EPVDemo)
+> It might be cool to include one of their graphics here, if can. Or regenerate one.
 
-Another framework for estimating expected points in basketball was proposed by @Sicilia2019DeepHoops. 
+Another framework for estimating expected points in basketball is proposed by @Sicilia2019DeepHoops. 
 The authors offer a different point of view on expected points, where they first consider a classification model which returns the probabilities for whether a player would commit a foul (shooting and non-shooting), turnover, or attempt a shot.
-The values associated with each of those "terminal actions" are then utilized to compute the expected points within a basketball play.
+The values associated with each of those "terminal actions" are then used to compute the expected points within a basketball play.
 
@@
 
-In American football, @Yurko2020going use tracking data provided by the 2019 NFL Big Data Bowl (citation) to model the expected yards gained for a ball-carrier during the course of a play. 
+In American football, @Yurko2020going use tracking data provided by the [2019 NFL Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) to model the expected yards gained for a ball-carrier during the course of a play.
 As an extension to pre-existing approaches, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play, rather than only producing a single estimate for the expected yards gained.
-Accordingly, the probability of various types of outcome at the end of a play such as a touchdown or a first-down gain can be computed from the yards gained distribution.
+Accordingly, the probability of various types of outcomes at the end of a play such as a touchdown or a first-down gain can be computed from the yards gained distribution.
 
-The notion of EPV has also been extended to soccer recently.
+The notion of EPV has also been extended to soccer.
 @fernandez2021framework implement deep learning methods to examine the instantaneous expected value of soccer possessions. 
@@
 For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs.
-If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy that maximizes the expected number of points will maximize the probability of winning:
+If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy $u$ that maximizes the expected number of points will maximize the probability of winning:
 $$
-  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s]} \,.
+  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s, u]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s, u]} \,.
 $$
@@
 
-A related, but different, concept is the notion of *win probability*.
+A related, but different concept is the notion of *win probability*.
 Win probability is simply an estimate of the probability that a team will win the game, given its current state $s$. 
@@
 
-In this section, we discuss notable previous work on win probability in baseball, football, basketball, and several other sports.
+In this section, we discuss notable previous work on win probability in baseball, American football, basketball, and several other sports.
 
@@
 
-## Football
+## American football
 
@@
 The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 left in the game. 
-The last play of this game, famously known as the [Minneapolis Miracle](https://en.wikipedia.org/wiki/Minneapolis_Miracle), resulted in a drastic swing in win probabilities for both teams. 
+The last play of this game---famously known as the [Minneapolis Miracle](https://en.wikipedia.org/wiki/Minneapolis_Miracle)---resulted in a drastic swing in win probabilities for both teams. 
 With 10 seconds remaining in the game, the Vikings begin the final possession with a 25.3% chance of winning. 
@@
 p(l, t) = \Phi\left(\frac{l + (1-t)\mu}{\sqrt{(1-t)\sigma^2}}\right)
-,
+\,.
 $$
@@
 @Brenzel2019 uses three-dimensional Markov models to estimate win probability throughout a curling match. 
-In particular, the authors propose both a homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game.
-In esports, @Maymin2020 uses logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
-More recently, @Guan2022 develops an in-game win probability model for the National Rugby League using functional data analysis. 
+In particular, the authors propose both homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game.
+In esports, @maymin2021smart uses logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
+@Guan2022 develops an in-game win probability model for the National Rugby League using functional data analysis. 
 In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.
@@
 
-Another important idea in sports analytics is the estimation of team strength. 
+A third crucial idea in sports analytics is the estimation of team strength. 
 A popular method for estimating team strength in sports is through pairwise evaluations. 
@@
 $$
-\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j
+\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
 $$
@@
 
-The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. 
-As an example, we consider the data given in @agresti2018introduction (page 247) on tennis results from 2014--2018 for five men's professional players Novak Djokovic, Roger Federer, Andy Murray, Rafael Nadal, and Stan Wawrinka.
+The BTM can be implemented in `R` [@R-base] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. 
+As an example, we consider the data given in @agresti2018introduction (page 247) on tennis results from 2014--2018 for five men's professional players: Novak Djokovic, Roger Federer, Andy Murray, Rafael Nadal, and Stan Wawrinka.
 We fit a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.
@@
 
-List contributions from individuals who do not meet the criteria for authorship (for example, to recognize people who provided technical help, collation of data, writing assistance, acquisition of funding, or a department chairperson who provided general support), with permission from the individual.
-Thanks to anonymous reviewers are not appropriate.
+We are grateful to Michael Lopez and Katherine Evans for their thoughts on early versions of this paper.
 

### 2022-07-15 sports_analytics.Rmd
added epv basketball figure

@@
 @goldner2012markov and @goldner2017situational propose a Markov model for estimating expected points in football. 
-In particular, the author consider a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
+In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
 For any given play, the expected points is calculated using the absorption probabilities for different scoring events.
@@
 Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
-
 The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes.
-
 This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.
-The expected points for a play can then be calculated accordingly, by taking the sum of the products of the scoring event point values and their associated probabilities.
+The expected points for a play can then be calculated accordingly, by summing up the products of the scoring event point values and their associated probabilities.
 
@@
 
-(include code demo? https://github.com/dcervone/EPVDemo)
-> It might be cool to include one of their graphics here, if can. Or regenerate one.
+A demonstration of the EPV model presented in @cervone2016multiresolution's paper is available on the first author's GitHub at <https://github.com/dcervone/EPVDemo>.
+As an illustration, using the provided tracking data, Figure \ref{fig:epv-demo} shows a snapshot of a possession during the 2013-14 NBA regular season matchup between the Miami Heat and the Brooklyn Nets on November 1, 2013.
+Miami is the team on offense in this play, and the possession outcome is a 26-foot three-point miss by Mario Chalmers of the Heat.
+The plot consists of two elements: 1) (bottom) the player locations on the court at a particular moment in this possession--when the ball just left Chalmers' hands, and 2) (top) a line graph showing how the EPV has continuously changed throughout the play until the three-point attempt.
+For this entire possession, the estimated EPV for the Miami Heat peaks at this single time point at $1.276$.
+
+<!-- (include code demo? https://github.com/dcervone/EPVDemo) -->
+<!-- > It might be cool to include one of their graphics here, if can. Or regenerate one.  -->
+
+```{r epv-demo, out.width="100%", fig.cap="Player locations and estimated EPV for a possession during Miami Heat (red) vs.
+Brooklyn Nets (black) on November 1, 2013.
+The captured moment is when Miami's Mario Chalmers just releases a three-point shot, which ends up being a miss."}
+# Source: https://github.com/dcervone/EPVDemo/blob/master/gifs/poss_12.gif
+# library(magick)
+# epv <- image_read("https://github.com/dcervone/EPVDemo/blob/master/gifs/poss_12.gif?raw=true")
+# image_write(epv[529], "epv_snap.png")
+knitr::include_graphics("epv_snap.png")
+```
 
@@
 
-Figure \ref{fig:nflfastr_wp} shows a win probability graph for the January 14, 2018 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. 
+Figure \ref{fig:nflfastr-wp} shows a win probability graph for the January 14, 2018 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. 
 We obtain the estimated probability of winning for each team using the \pkg{nflfastR} R package, which implements a gradient boosting model via the \pkg{xgboost} library (@R-xgboost) for estimating win probabilities.
@@
 
-```{r nflfastr_wp, fig.cap="Win probability graph for New Orleans Saints vs. Minnesota Vikings in the 2017--18 NFL Playoffs."}
+```{r nflfastr-wp, fig.cap="Win probability graph for New Orleans Saints vs. Minnesota Vikings in the 2017--18 NFL Playoffs."}
 library(nflfastR)
@@
 The idea of win probability is also applied to other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team. 
-@Brenzel2019 uses three-dimensional Markov models to estimate win probability throughout a curling match. 
+@Brenzel2019 use three-dimensional Markov models to estimate win probability throughout a curling match. 
 In particular, the authors propose both homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game.
 In esports, @maymin2021smart uses logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
-@Guan2022 develops an in-game win probability model for the National Rugby League using functional data analysis. 
+Moreover, @Guan2022 develop an in-game win probability model for the National Rugby League using functional data analysis. 
 In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.
@@
 
-Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM. 
-According to the abilities, between 2014 and 2018 the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray.
-To obtain win probabilities, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is:
-$$
-\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768
-$$
-
 ```{r}
@@
 ```
+
+Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM. 
+According to the abilities, between 2014 and 2018 the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray.
+To obtain win probabilities, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is
+$$
+\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768.
+$$
 

### 2022-07-15 sports_analytics.Rmd
minor edits

@@
 In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
-Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot.
-[CITATION NEEDED?]
+Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot [@Schuhmann2021nba3point].
 In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative.
@@
 A demonstration of the EPV model presented in @cervone2016multiresolution's paper is available on the first author's GitHub at <https://github.com/dcervone/EPVDemo>.
-As an illustration, using the provided tracking data, Figure \ref{fig:epv-demo} shows a snapshot of a possession during the 2013-14 NBA regular season matchup between the Miami Heat and the Brooklyn Nets on November 1, 2013.
+As an illustration, using the provided tracking data, Figure \ref{fig:epv-demo} displays a snapshot of a possession during the 2013-14 NBA regular season matchup between the Miami Heat and the Brooklyn Nets on November 1, 2013.
 Miami is the team on offense in this play, and the possession outcome is a 26-foot three-point miss by Mario Chalmers of the Heat.
-The plot consists of two elements: 1) (bottom) the player locations on the court at a particular moment in this possession--when the ball just left Chalmers' hands, and 2) (top) a line graph showing how the EPV has continuously changed throughout the play until the three-point attempt.
-For this entire possession, the estimated EPV for the Miami Heat peaks at this single time point at $1.276$.
+The plot consists of two elements: 1) (bottom) the player locations on the court at a particular moment in this possession--when the ball just left Chalmers' hands, and 2) (top) a line graph showing how the EPV is continuously changing throughout the play until the three-point attempt.
+For this entire possession, the estimated EPV for the Miami Heat reaches its peak at $1.276$ at this specific time point.
 

### 2022-07-15 sports_analytics.Rmd
chess

@@
 ```{r chess-startup, fig.cap="The starting chess board printed via the \\pkg{chess} package."}
-#library(chess)
-#Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
-#game() %>% plot()
+# library(chess)
+# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
+# game() %>% plot()
+
+# we could plot greg's chess moves
+# doesn't work for me... python/reticulate issues
+# library(chessR)
+# greg <- get_raw_chessdotcom(usernames = "fatgreggy",
+#                             year_month = "202206")
+# m <- extract_moves_as_game(greg[1, ])
+# plot_moves(m)
 ```

### 2022-07-15 sports_analytics.Rmd
minor things

@@
 The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision. 
+The computation of the EPV statistic is done using a Markov model.
 Consequently, the authors derive a metric called EPV-Added (EPVA), measuring a player's EPV contribution in a given situation relative to a league-average player.
@@
 Miami is the team on offense in this play, and the possession outcome is a 26-foot three-point miss by Mario Chalmers of the Heat.
-The plot consists of two elements: 1) (bottom) the player locations on the court at a particular moment in this possession--when the ball just left Chalmers' hands, and 2) (top) a line graph showing how the EPV is continuously changing throughout the play until the three-point attempt.
+The plot consists of two elements: 
+1) (bottom) the player locations on the court at a particular moment in this possession--when the ball just left Chalmers' hands, 
+and 2) (top) a line graph showing how the EPV is continuously changing throughout the play until the three-point attempt.
 For this entire possession, the estimated EPV for the Miami Heat reaches its peak at $1.276$ at this specific time point.
@@
 
-```{r epv-demo, out.width="100%", fig.cap="Player locations and estimated EPV for a possession during Miami Heat (red) vs.
-Brooklyn Nets (black) on November 1, 2013.
+```{r epv-demo, out.width="100%", fig.cap="Player locations and estimated EPV for a possession during the Miami Heat (red) vs.
+Brooklyn Nets (black) NBA game on November 1, 2013.
 The captured moment is when Miami's Mario Chalmers just releases a three-point shot, which ends up being a miss."}
@@
 
-@Stern1994 uses Brownian motion to investigate the scoring process of basketball. 
+@Stern1994 provides an investigation of in-game win probability and the scoring process in basketball using a Brownian motion model. 
 Let $p(l, t)$ represent the win probability for the home team given an $l$-point lead after $t$ seconds of game time. 
 The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$. 
-In particular, 
+Specifically, 
 $$
@@
 On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
-In particular, they propose a Bayesian lasso regression model to assess the impact of individual players on their team's win probability at a given time of a basketball game. 
-This model assumes independence of observation, constant variability in win probability, and also addresses multicollinearity.
+In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's win probability at any given time of a basketball game. 
+This model assumes independence of observations and constant variability in win probability.
 
-@McFarlane2019 uses logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA. 
+In addition, @McFarlane2019 uses logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA. 
 The approach takes into account the remaining game time, score difference, and point spread. 

### 2022-07-15 sports_analytics.Rmd
small notes

@@
 As an extension to pre-existing approaches, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play, rather than only producing a single estimate for the expected yards gained.
-Accordingly, the probability of various types of outcomes at the end of a play such as a touchdown or a first-down gain can be computed from the yards gained distribution.
+Accordingly, the probability of various types of outcomes at the end of a play such as a touchdown or a first-down gain can be computed from the distribution of the end-of-play yard line.
 
@@
 
+<!-- I think we should also mention the sportsdataverse (https://github.com/sportsdataverse, https://sportsdataverse.org)
+
+maybe some kaggle competitions (march madness, cite mike and greg's paper; and big data bowl)
+-->
+
 @marchi2018analyzing

### 2022-07-15 sports_analytics.Rmd
refs for #11

@@
 Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot [@Schuhmann2021nba3point].
-In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative.
-[LOPEZ?]
+In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
 And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact.

### 2022-07-15 sports_analytics.Rmd
tweaks to EPV section

@@
 
-In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as means toward an assessment of a player's on-court performance. 
+In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as a means toward an assessment of a player's on-court performance. 
 This metric is a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations. 
@@
 
-A demonstration of the EPV model presented in @cervone2016multiresolution's paper is available on the first author's GitHub at <https://github.com/dcervone/EPVDemo>. 
-As an illustration, using the provided tracking data, Figure \ref{fig:epv-demo} displays a snapshot of a possession during the 2013-14 NBA regular season matchup between the Miami Heat and the Brooklyn Nets on November 1, 2013.
-Miami is the team on offense in this play, and the possession outcome is a 26-foot three-point miss by Mario Chalmers of the Heat.
+A demonstration of the EPV model presented in @cervone2016multiresolution is available at <https://github.com/dcervone/EPVDemo>. 
+Figure \ref{fig:epv-demo} illustrates how the provided tracking data informs the evolution of EPV throughout the play. 
+It displays a snapshot of a possession during the NBA regular season matchup between the Miami Heat and the Brooklyn Nets on November 1, 2013. 
+Miami is the team on offense in this possession, whose outcome is a 26-foot three-point miss by Mario Chalmers. 
 The plot consists of two elements: 
-1) (bottom) the player locations on the court at a particular moment in this possession--when the ball just left Chalmers' hands, 
-and 2) (top) a line graph showing how the EPV is continuously changing throughout the play until the three-point attempt.
-For this entire possession, the estimated EPV for the Miami Heat reaches its peak at $1.276$ at this specific time point.
-
-<!-- (include code demo? https://github.com/dcervone/EPVDemo) -->
-<!-- > It might be cool to include one of their graphics here, if can. Or regenerate one.  -->
+1) (bottom) the player locations on the court at a particular moment in this possession: when the ball just left Chalmers' hands, 
+and 2) (top) a line graph showing how the EPV changes continuously throughout the play until the three-point attempt. 
+For this possession, the estimated EPV for the Miami Heat reaches its peak at $1.276$ points at the moment the shot is taken.
 
@@
 ```
+
+Note that Miami starts the play with an EPV of approximately 1.0 points.
+This indicates their implied average points per possession. 
+Chalmers's shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
+A breakthrough in this work is that this estimate is conditional on the location of the other 9 players on the court!
 

### 2022-07-15 sports_analytics.Rmd
added chess figures

@@
 
-```{r chess-startup, fig.cap="The starting chess board printed via the \\pkg{chess} package."}
-# library(chess)
+```{r chess, fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
+library(chess)
 # Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
-# game() %>% plot()
+game() |> plot()
 
@@
 # doesn't work for me... python/reticulate issues
-# library(chessR)
-# greg <- get_raw_chessdotcom(usernames = "fatgreggy",
-#                             year_month = "202206")
-# m <- extract_moves_as_game(greg[1, ])
-# plot_moves(m)
+library(chessR)
+us <- c("fatgreggy")
+greg <- get_raw_chessdotcom(
+  usernames = us,
+  year_month = 202206
+)
+
+m <- extract_moves_as_game(greg[6, ])
+# plot_moves(m, interactive = FALSE)
+plot(m)
 ```

### 2022-07-15 sports_analytics.Rmd
wrote chess paragraph; #6

@@
     "nflverse",
-    "xgboost"
+    "xgboost",
+    "chess", 
+    "chessR"
   ),
@@
 
-Figure \ref{fig:chess-startup} shows a rendering of the starting chess board obtained via the \pkg{chess} package.
+Many tools in sports analytics provide the ability to read, write, and plot data stored in a sport-specific format. 
+For example, consider chess, where the sequence of moves in games is often recorded in [Portable Game Notation](https://en.wikipedia.org/wiki/Portable_Game_Notation).
+Software tools can then be built around this well-defined format. 
+The \pkg{chess} package [@R-chess] provides R users with the ability to read, write, display, and manipulate chess data in PGN format. 
+[Application programming interfaces](https://en.wikipedia.org/wiki/API) (APIs) are also common. 
+In chess, the \pkg{chessR} package [@R-chessR] allows R users to download game data from the Chess.com API. 
+This type of infrastructure, where one package is the "workhorse" that facilitates common generic data operations, and other packages layer on specific functionality, is common in sports analytics.
+
+Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in a game recently won by one of the authors (downloaded via the \pkg{chessR} package).
 

### 2022-07-15 sports_analytics.Rmd
first paragraph in tools; #6

@@
   Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years.
-We connect four(?) big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport.
+We connect three big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport.
 While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data.
@@
 
-See @baumer2022ctv for R packages useful for sports analytics.
+Analytical work in sports requires facility with an ever-changing set of computational tools for working with data. 
+Sources of authoritative data about sports are myriad, and are too numerous to list here. 
+Software tools for sports analytics are similarly numerous. 
+For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network and organizes them by sport [@baumer2022ctv].
+In the remainder of this section, we highlight a few tools for sports analytics that are of general interest and illustrate a common paradigm for how these tools can be used in conjunction.
 

### 2022-07-15 sports_analytics.Rmd
teamcolors; #6

@@
 For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network and organizes them by sport [@baumer2022ctv].
+A more general collection of software tools is being curated by the SportsDataverse.org (<https://sportsdataverse.org>).
+
 In the remainder of this section, we highlight a few tools for sports analytics that are of general interest and illustrate a common paradigm for how these tools can be used in conjunction.
 
-<!-- I think we should also mention the sportsdataverse (https://github.com/sportsdataverse, https://sportsdataverse.org)
-
-maybe some kaggle competitions (march madness, cite mike and greg's paper; and big data bowl)
--->
-
-@marchi2018analyzing
-
-@R-Lahman, @R-pitchRx, @R-retrosheet
+## Case study in how tools fit together: chess
 
@@
 
-@R-teamcolors provides color palettes for professional and amateur sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
-The team colors and logos shown in Figure \ref{fig:nflfastr_wp} were provided by the \pkg{teamcolors} package. 
-Figure \ref{fig:teamcolors} illustrates team color palettes for professional sports leagues.
+## Colors
+
+Each professional sports team has its own brand, most obviously identified by a team logo and set of colors. 
+The \pkg{teamcolors} package [@R-teamcolors] provides color palettes and logos for men's and women's professional and collegiate sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
+The team colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
+Figure \ref{fig:teamcolors} illustrates the full array of team color palettes for professional sports leagues.
 
@@
 ```
+
+# Notable competitions and public events
+
+<!--
+
+maybe some kaggle competitions (march madness, cite mike and greg's paper; and big data bowl)
+-->
+
+@marchi2018analyzing
+
+@R-Lahman, @R-pitchRx, @R-retrosheet
 

### 2022-07-15 sports_analytics.Rmd
skeleton of competitions; #6
competitions and conferences
added repo URL

@@
 
-<!--
+## Competitions
 
-maybe some kaggle competitions (march madness, cite mike and greg's paper; and big data bowl)
--->
+- Kaggle March Madness [@lopez2015building]
+- Big Data Bowl
+- Big Data Cup
+- SABR Case competition
+
+## Conferences
+
+- NESSIS/CASSIS
+- Sloan
+- CMU
+- UConn sports analytics
 
@@
 
+The R Markdown file that generated this paper, including all R code, is available at <https://github.com/beanumber/wire21>.
+
 # References {-}

### 2022-07-15 sports_analytics.Rmd
small fix

@@
   annotation_raster(logos[[2]], -1010, -1490, 0.135, 0.24) + 
-  scale_x_reverse()
+  scale_x_reverse(breaks = seq(0, 3600, 900))
 ```
@@
 For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network and organizes them by sport [@baumer2022ctv].
-A more general collection of software tools is being curated by the SportsDataverse.org (<https://sportsdataverse.org>).
+A more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].
 
@@
 
-# we could plot greg's chess moves
-# doesn't work for me... python/reticulate issues
 library(chessR)

### 2022-07-16 sports_analytics.Rmd
some formatting

@@
 library(tidyverse)
+library(kableExtra)
 ```
@@
 erm <- tibble(
-  base = rep(0:7, each = 3),
-  out = rep(0:2, times = 8),
-  runs = lindsey
+  Base = rep(0:7, each = 3),
+  Out = rep(0:2, times = 8),
+  Runs = lindsey
 )
@@
 erm |>
-  pivot_wider(names_from = out, values_from = runs) |>
-  mutate(base = int2bin(base)) |>
-  rename(`base\\\\out` = base) |>
-  knitr::kable(caption = "George Lindsey's expected run matrix.
+  pivot_wider(names_from = Out, values_from = Runs) |>
+  mutate(Base = int2bin(Base)) |>
+  kable(booktabs = TRUE, 
+        caption = "George Lindsey's expected run matrix.
 Note how (when reading across the rows) the expected runs decrease as outs increase for the same configuration of baserunners, while (when reading down the columns) expected runs generally increase as baserunners advance.
-\\label{tab:erm}")
+\\label{tab:erm}") |>
+  add_header_above(c("", "Out" = 3))
 ```
@@
 Brooklyn Nets (black) NBA game on November 1, 2013.
-The captured moment is when Miami's Mario Chalmers just releases a three-point shot, which ends up being a miss."}
+The captured moment is when Miami's Mario Chalmers just releases a three-point shot, which ends up as a missed field goal."}
 # Source: https://github.com/dcervone/EPVDemo/blob/master/gifs/poss_12.gif
@@
   mutate(
-    pr_0 = c(
+    Pr_0 = c(
       .747, .855, .933,
@@
 erm |>
-  filter(out == 0, base %in% c(5, 7)) |>
-  mutate(base = int2bin(base)) |>
-  knitr::kable(caption = "Loading the bases can be a rational strategy.
+  filter(Out == 0, Base %in% c(5, 7)) |>
+  mutate(Base = int2bin(Base)) |>
+  kable(booktabs = TRUE, 
+        caption = "Loading the bases can be a rational strategy.
 In this case, loading the bases maximizes both the expected number of runs scored *and* the probability of not scoring at all.
@@
 @Lock2014 use random forests as a means for estimating win probability before each play in a football game. 
-Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yardline, and yards to go for a first down.
+Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yard line, and yards to go for a first down.
 
@@
 Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
-
 $$
@@
 $$
-
 Here, $\exp{\beta_i}$ is often viewed as a representation of player $i$'s ability.
@@
   round(3) |> 
-  knitr::kable(caption = "Estimated abilities for each player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}")
+  kable(booktabs = TRUE, 
+        col.names = c("Ability", "SE"),
+        caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}")
 ```

### 2022-07-16 sports_analytics.Rmd
jqas special issues; #11

@@
 At the same time, many sports are similar, either because one evolved from the other, or the structure of the game shares certain attributes. 
-Sports that are closely related anthropologically (NEED EXAMPLE) may or may not share common applications of analytical methods. 
-Conversely, with just a few small tweaks, analytical metrics might work just as well across sports that are unrelated and quite different (e.g., an Elo rating could be equally valid for chess players and ice hockey teams).
+Sports that are closely related anthropologically may or may not share common applications of analytical methods. 
+For example, although belonging to the same bat-and-ball family, baseball and cricket differ in strategies such as batting order or sacrifice plays.
+Conversely, with just a few small tweaks, analytical metrics might work just as well across sports that are unrelated and quite different.
+For instance, an Elo rating could be equally valid for chess players and ice hockey teams.
 
@@
 In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a game state at any given point in a sporting contest. 
-
 Mathematically, we define $X$ to be a random variable indicating the the number of points (or runs) that a team will score over some determined amount of time (e.g.
@@
 Chalmers's shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
-A breakthrough in this work is that this estimate is conditional on the location of the other 9 players on the court!
+A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court!
 
@@
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
-library(chess)
-# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
-game() |> plot()
-
-library(chessR)
-us <- c("fatgreggy")
-greg <- get_raw_chessdotcom(
-  usernames = us,
-  year_month = 202206
-)
-
-m <- extract_moves_as_game(greg[6, ])
-# plot_moves(m, interactive = FALSE)
-plot(m)
+# library(chess)
+# # Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
+# game() |> plot()
+# 
+# library(chessR)
+# us <- c("fatgreggy")
+# greg <- get_raw_chessdotcom(
+#   usernames = us,
+#   year_month = 202206
+# )
+# 
+# m <- extract_moves_as_game(greg[6, ])
+# # plot_moves(m, interactive = FALSE)
+# plot(m)
 ```
@@
 
-# Notable competitions and public events
+# Opportunities
 
@@
 
-- Kaggle March Madness [@lopez2015building]
-- Big Data Bowl
+- Kaggle March Madness [@lopez2015building] (jqas special issue: https://www.degruyter.com/journal/key/jqas/11/1/html)
+- Big Data Bowl (special issue https://www.degruyter.com/journal/key/jqas/16/2/html)
 - Big Data Cup

### 2022-07-16 sports_analytics.Rmd
1 additional sentence on epv soccer

@@
   kable(booktabs = TRUE, 
+        linesep = "",
         caption = "George Lindsey's expected run matrix.
@@
 
-The notion of EPV has also been extended to soccer.
+The notion of EPV has also been extended to association football (soccer.)
 @fernandez2021framework implement deep learning methods to examine the instantaneous expected value of soccer possessions. 
 This approach considers passes, ball drives, and shots in soccer as the main set of actions used to compute expected possession value. 
+Many applications can be derived from this framework, including predicting which footballer is most likely to receive the next pass from the current on-ball player.
 

### 2022-07-16 sports_analytics.Rmd
python

@@
 Sports that are closely related anthropologically may or may not share common applications of analytical methods. 
-For example, although belonging to the same bat-and-ball family, baseball and cricket differ in strategies such as batting order or sacrifice plays.
+For example, despite belonging to the same bat-and-ball family, baseball and cricket differ in strategies such as batting order or sacrifice plays.
 Conversely, with just a few small tweaks, analytical metrics might work just as well across sports that are unrelated and quite different.
@@
 
-A more in-depth overview of the history of expected points is provided in @Yurko2019nflwar (Section 1.1). 
+A more in-depth overview of the history of expected points in sports is provided in @Yurko2019nflwar (Section 1.1). 
 Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
@@
 This approach considers passes, ball drives, and shots in soccer as the main set of actions used to compute expected possession value. 
-Many applications can be derived from this framework, including predicting which footballer is most likely to receive the next pass from the current on-ball player.
+Many applications can be derived from this framework, including predicting which footballer on the pitch is most likely to receive the next pass from the current on-ball player.
 
@@
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
-# library(chess)
-# # Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
-# game() |> plot()
-# 
-# library(chessR)
-# us <- c("fatgreggy")
-# greg <- get_raw_chessdotcom(
-#   usernames = us,
-#   year_month = 202206
-# )
-# 
-# m <- extract_moves_as_game(greg[6, ])
-# # plot_moves(m, interactive = FALSE)
-# plot(m)
+library(chess)
+# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
+game() |> plot()
+
+library(chessR)
+us <- c("fatgreggy")
+greg <- get_raw_chessdotcom(
+  usernames = us,
+  year_month = 202206
+)
+
+m <- extract_moves_as_game(greg[6, ])
+# plot_moves(m, interactive = FALSE)
+plot(m)
 ```

### 2022-07-18 sports_analytics.Rmd
added to wp

@@
 
-@Albert2015 provides a historical overview of the use of win probability in baseball. 
-This notion goes back to at least as early as @Lindsey1961, where the expected win probability after each inning was calculated based on the distribution of runs scored in each inning. 
-@Mills1970 utilize win probability to introduce Player Win Average (PWA), a measure of a player's contribution to the game outcome. 
+The notion of win probability in baseball goes back to at least as early as @Lindsey1961.
+In this paper, the author calculates the expected win probability after each inning based on the distribution of runs scored in each inning. 
+Inspired by @Lindsey1961's work, @Mills1970 utilize win probability to introduce Player Win Average (PWA), a measure of a player's contribution to the game outcome. 
 In particular, PWA is computed as 
@@
 
+Additionally, a mathematical model for estimating win probability in baseball is presented in @Tango2007book. 
+The authors use Markov chains to look at win expectancy throughout the course of a baseball game. 
+This approach considers different states of the game such as base, inning, outs and score, and outputs win probabilities accordingly.
+
+See @Albert2015 for a more detailed historical overview of the use of win probability in baseball.
+
 ## American football
@@
 
-@Lock2014 use random forests as a means for estimating win probability before each play in a football game. 
-Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yard line, and yards to go for a first down.
+@Lock2014 use a random forest method to provide a win probability estimate before each play in a football game. 
+Covariates included in this tree-based win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yard line, and yards to go for a first down.
+According to this model, the difference in score between the two teams is the most important feature for predicting win probabilities at any given moment in an NFL game.
 
@@
 We obtain the estimated probability of winning for each team using the \pkg{nflfastR} R package, which implements a gradient boosting model via the \pkg{xgboost} library (@R-xgboost) for estimating win probabilities.
-Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, 0.742 at the end of the first, second, and third quarters, respectively. 
+Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, and 0.742 at the end of the first, second, and third quarters, respectively. 
 The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 left in the game. 
@@
 On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
-In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's win probability at any given time of a basketball game. 
+In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
 This model assumes independence of observations and constant variability in win probability.
@@
 The approach takes into account the remaining game time, score difference, and point spread. 
-This basketball win probability model is then applied to the calculation of the End-of-game Tactics Metric (ETM), measuring how the win probability differs between the optimal and on-court actual decisions.
+This win probability model is then applied to the calculation of the End-of-game Tactics Metric (ETM), measuring how the chance of winning a basketball game differs between the optimal and on-court actual decisions.
 
@@
         col.names = c("Ability", "SE"),
-        caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}")
+        caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}") |> 
+  kable_styling(latex_options = "hold_position")
 ```

### 2022-07-18 sports_analytics.Rmd
eval chess chunk based on username

@@
 
-```{r chess, fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
-# library(chess)
-# # Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
-# game() |> plot()
-# 
-# library(chessR)
-# us <- c("fatgreggy")
-# greg <- get_raw_chessdotcom(
-#   usernames = us,
-#   year_month = 202206
-# )
-# 
-# m <- extract_moves_as_game(greg[6, ])
-# # plot_moves(m, interactive = FALSE)
-# plot(m)
+library(chess)
+# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
+game() |> plot()
+
+library(chessR)
+us <- c("fatgreggy")
+greg <- get_raw_chessdotcom(
+  usernames = us,
+  year_month = 202206
+)
+
+m <- extract_moves_as_game(greg[6, ])
+# plot_moves(m, interactive = FALSE)
+plot(m)
 ```

### 2022-07-18 sports_analytics.Rmd
merge commit

@@
 We connect three big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport.
-While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data.
-Where appropriate, we outline how data, models, and knowledge of the sport combine to generate actionable insights.
+While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and sources of data.
+Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights.
 This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics.
@@
 
+<!--
 Sum up the key conclusions of your review, highlighting the most promising scientific developments, directions for future research, applications, etc.
 The conclusion should be ~2 paragraphs, ~750 words total.
+-->
+
+As an applied science, sports analytics may lack a [grand unified theory](https://en.wikipedia.org/wiki/Grand_Unified_Theory) that succinctly characterizes game play across all sports. 
+However, as a maturing discipline, sports analytics has been able to address fundamental questions common to many sports. 
+In this paper, we explicate three of those big questions:
+
+- Who are the best teams and how good are they? 
+- What is the likelihood of each team winning the game at any given juncture? 
+- Is there a generic framework for evaluating strategies at any given juncture in a game?
+
+Other fundamental questions, such as:
+
+- How significant is the element of chance in a particular sport? 
+- Given that we know how the best teams are, who are the best players and how can we quantify their relative contributions? 
+- What combinations of players work best together in a particular sport?
+
+are addressed elsewhere. 
+In particular, see @Lopez2018 for estimations of the element of chance across four major sports. 
+The second question is often addressed using a formulation of *wins above replacement* (WAR)---see @baumer2015openwar and @Yurko2019nflwar for details in baseball and American football. 
+The third question is most compelling in sports like basketball, ice hockey, and soccer, where substitutions are common and it is obvious that different combinations of players with different sets of skills will result in squad of varying strengths and weaknesses.
+The concept of *plus-minus*, and then *adjusted* plus-minus is frequently applied to address this question (see @hvattum2019comprehensive for a comprehensive overview of applications).
+
+In drawing together these three big ideas in sports analytics, we have also drawn attention to some computational tools for doing the work and opportunities to showcase that work. 
+It is through these exchanges of ideas, tools, models, and data that analytics moves our collective understanding of sports forward.
 

### 2022-07-18 sports_analytics.Rmd
merge commit

@@
   file = "pkgs.bib"
+)
+
+xfun::gsub_file(
+  "pkgs.bib",
+  "\\{Jason Zivkovic\\}",
+  "Zivkovic, Jason"
   )

### 2022-07-18 sports_analytics.Rmd
Some edits.

--- 2022-07-18 sports_analytics.Rmd
@@
 This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.
+
 The expected points for a play can then be calculated accordingly, by summing up the products of the scoring event point values and their associated probabilities.
@@
 ### 4th down strategy
+The concept of expected points in American football has many applications.
+One of the most notable and well-studied is the use of expected points in the evaluation of 4th down strategy, and there is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.
+@Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on example from the first quarter of a game (to avoid issues with end of half and end of game decision making).
+They conclude that teams don't go for it enough if they are trying to maximize their probability of winning the game.
+
+Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.
+@Romer2006 Examines 4th down decisions in the NFL using expected points.
+Focuses only on the first quarter.
+Teams don't go for it enough is they are trying to maximize their probability of winning the game.
+Later work in @Yam2019 used win probability, rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams were to conservative in going for it on 4th down.
+In addition, they estimate that a better strategy would have been worth about 0.4 wins per season on average, which is substantial in a season with only 16 games.
+
+Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.
 
@@
 4th down strategy papers based on expected points.  
-@Yam2019: 
-@Romer2006: Examines 4th down decisions in the NFL using expected points.
+@lopez2020bigger: Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.  
+<!-- @Yam2019: Uses Causal inference techniques to examine the effects of 4th down decision making in the NFL.   -->
+<!-- @Romer2006: Examines 4th down decisions in the NFL using expected points.
 Focuses only on the first quarter.
 Teams don't go for it enough is they are trying to maximize their probability of winning the game.
+-->
 @pelechrinis2019 Present an expected points framework where they adjust for the strength of the defense.  
@@
 https://www.eng.buffalo.edu/~jzhuang/Papers/JQAS_NFL_2011.pdf
-
-Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.
 

### 2022-07-25 sports_analytics.Rmd
expected pts updates

@@
 In this technical note, the authors estimate the expected points for 1st and 10 plays in the NFL given any yard line on the football field. 
-Due to limitations regarding the amount of data collected, the football field is broken into 10-yard strips, centered at their midpoints (e.g.
-5, 15, 25, 35, etc.), and the expected points are calculated for each of those field positions.
-Other early work on expected point values in American football can be found in @Carroll1988hidden, which popularized the idea of play evaluation using expected points and expected points added.
+Due to limitations regarding the amount of data collected, the football field is broken into 10-yard buckets, centered at their midpoints (e.g. 5, 15, 25, 35, etc.).
+The estimated value for first and 10 situation on those field positions is calculated by averaging the points scored on the next scoring instance.
+
+Other early work on expected point values in American football can be found in @Carroll1988hidden. 
+In particular, the authors extend @Carter1971operations's work and propose a linear model for expected points in the NFL.
+It is determined that every extra 25 yards is associated with 2 more points scored on average for a football team.
 
@@
 This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.
-
 The expected points for a play can then be calculated accordingly, by summing up the products of the scoring event point values and their associated probabilities.

### 2022-07-25 sports_analytics.Rmd
more on EP

@@
 The concept of estimating the value of the state of a game is easily extended to other sports. 
-For example, in American football, $s$ is determined by factors such as down, yardage to the next first down, time remaining in the game, and field position.
+For example, in American football, $s$ is determined by situational variables at the start of a play such as down, yardage to the next first down, time remaining in the game, and field position.
 
@@
 In this technical note, the authors estimate the expected points for 1st and 10 plays in the NFL given any yard line on the football field. 
-Due to limitations regarding the amount of data collected, the football field is broken into 10-yard buckets, centered at their midpoints (e.g. 5, 15, 25, 35, etc.).
-The estimated value for first and 10 situation on those field positions is calculated by averaging the points scored on the next scoring instance.
+Due to limitations regarding the amount of data collected, the football field is broken into 10-yard buckets, centered at their midpoints (e.g.
+5, 15, 25, 35, etc.), before averaging the value of the next scoring instance across the field to obtain the expected points.
 

### 2022-07-28 sports_analytics.Rmd
Some greg football changes.

--- 2022-07-18 sports_analytics.Rmd
@@
 Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
+
 The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes.
@@
 The expected points for a play can then be calculated accordingly, by summing up the products of the scoring event point values and their associated probabilities.
+
+@pelechrinis2019 develops an expected points framework in the same spirit as previous work, but accounts for the strength of the opponents in their method.
+They state that by failing to account for opponent strength appropriately, about 124.8 points per team each season (or about 3.8 wins per season) are not credited correctly.
 
@@
 
-Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.
+Recently, @lopez2020bigger presented an introduction to NFL tracking data, and as an example they examine 4th down behavior.
+Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.
+They demonstrate than teams are still conservative when choosing to go for it or punt on 4th down, but the effect may be overstated.
 
-(A paragraph on applications of expected points on decision making, fourth down strategy, with an example)
+### Other applications of expected points in American football
+@White2002 presents a tiered logistic regression method that can be, in general, applied to regression setting with a polychotomous response.
+Using this technique, they present an example where they estimate the value of NFL plays using a simple expected points model with down, to go, and to goal as predictors.
+They then use these results to evaluate and rank NFL quarterbacks.
 
-4th down strategy papers based on expected points.  
-@lopez2020bigger: Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.  
+@Alamar2010 uses an expected points framework to evaluate play calling in the NFL.
+However, rather than evaluate each play on its own, the framework presented here evaluates the play in the context of the drive.
+Based on play-by-play data from 2005 through 2008, they find that teams are underutilizing passing plays in some situations.
+
+<!-- (A paragraph on applications of expected points on decision making, fourth down strategy, with an example) -->
+
+<!-- 4th down strategy papers based on expected points.   -->
+<!-- @lopez2020bigger: Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.   -->
 <!-- @Yam2019: Uses Causal inference techniques to examine the effects of 4th down decision making in the NFL.   -->
@@
 -->
-@pelechrinis2019 Present an expected points framework where they adjust for the strength of the defense.  
+<!-- @pelechrinis2019 Present an expected points framework where they adjust for the strength of the defense.   -->
 <!-- New york times article: https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html?_r=0 -->
 
-@Alamar2010 Evaluation of play calling in the NFL based on expected points.  
-https://www.degruyter.com/document/doi/10.2202/1559-0410.1235/html
+<!-- @Alamar2010 Evaluation of play calling in the NFL based on expected points.   -->
+<!-- https://www.degruyter.com/document/doi/10.2202/1559-0410.1235/html -->
 
@@
 Shows that teams are moderately risk and loss averse.
-
-@White2002 Evaluated quaraterback usein gtiered multinomial regression in an expected points framework. They use a simple expected points model with down, to go, and to goal as predictors.   
 https://www.eng.buffalo.edu/~jzhuang/Papers/JQAS_NFL_2011.pdf
@@
 In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football.
+
 These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require few assumptions, and produce feature importance scores.
@@
 In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game.
+
 This model assumes independence of observations and constant variability in win probability.

### 2022-07-28 sports_analytics.Rmd

@@
 In this technical note, the authors estimate the expected points for 1st and 10 plays in the NFL given any yard line on the football field. 
-Due to limitations regarding the amount of data collected, the football field is broken into 10-yard strips, centered at their midpoints (e.g.
-5, 15, 25, 35, etc.), and the expected points are calculated for each of those field positions.
-Other early work on expected point values in American football can be found in @Carroll1988hidden, which popularized the idea of play evaluation using expected points and expected points added.
+Due to limitations regarding the amount of data collected, the football field is broken into 10-yard buckets, centered at their midpoints (e.g.
+5, 15, 25, 35, etc.), before averaging the value of the next scoring instance across the field to obtain the expected points.
+
+Other early work on expected point values in American football can be found in @Carroll1988hidden. 
+In particular, the authors extend @Carter1971operations's work and propose a linear model for expected points in the NFL.
+It is determined that every extra 25 yards is associated with 2 more points scored on average for a football team.
 

### 2022-07-28 sports_analytics.Rmd
NFL expected points section updates.

@@
 
+@urschel2011 uses expected points to evaluate NFL coaching decisions on kickoffs.
+Specifically, they look at surprise on-sides kicks vs regular kickoff and the decision to accept a touchback vs returning kickoff.
+Using data from the 2009 NFL season, the conclude, as many have, that coaches in the NFL tend to make conservative decisions.
+
+<!-- decsions on kick offs (surprise on sides or regular kick off AND accept touchback or return kick) using an expected points frame work.  Shows that teams are moderately risk and loss averse. -->
+<!-- https://www.eng.buffalo.edu/~jzhuang/Papers/JQAS_NFL_2011.pdf -->
+
 <!-- (A paragraph on applications of expected points on decision making, fourth down strategy, with an example) -->
@@
 
-@urschel2011 Examines decsions on kick offs (surprise on sides or regular kick off AND accept touchback or return kick) using an expected points frame work.
+<!-- @urschel2011 Examines decsions on kick offs (surprise on sides or regular kick off AND accept touchback or return kick) using an expected points frame work.
 Shows that teams are moderately risk and loss averse.
-https://www.eng.buffalo.edu/~jzhuang/Papers/JQAS_NFL_2011.pdf
+-->
+<!-- https://www.eng.buffalo.edu/~jzhuang/Papers/JQAS_NFL_2011.pdf -->
 

### 2022-07-28 sports_analytics.Rmd
light edits

@@
 ### 4th down strategy
+
 The concept of expected points in American football has many applications.
 One of the most notable and well-studied is the use of expected points in the evaluation of 4th down strategy, and there is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.
-@Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on example from the first quarter of a game (to avoid issues with end of half and end of game decision making).
+@Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on examples from the first quarter of a game (to avoid issues with end-of-half and end-of-game decision making).
 They conclude that teams don't go for it enough if they are trying to maximize their probability of winning the game.
 
-Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.
-@Romer2006 Examines 4th down decisions in the NFL using expected points.
-Focuses only on the first quarter.
-Teams don't go for it enough is they are trying to maximize their probability of winning the game.
-Later work in @Yam2019 used win probability, rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams were to conservative in going for it on 4th down.
-In addition, they estimate that a better strategy would have been worth about 0.4 wins per season on average, which is substantial in a season with only 16 games.
+Numerous papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy. 
+In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down.
+In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.
 
@@
 ### Other applications of expected points in American football
-@White2002 presents a tiered logistic regression method that can be, in general, applied to regression setting with a polychotomous response.
-Using this technique, they present an example where they estimate the value of NFL plays using a simple expected points model with down, to go, and to goal as predictors.
+
+@White2002 present a tiered logistic regression method that can be, in general, applied to any regression setting with a polychotomous response.  
+Using this technique, they estimate the value of NFL plays using a simple expected points model with down, yards to go, and yards to goal as predictors.
 They then use these results to evaluate and rank NFL quarterbacks.
@@
 @Alamar2010 uses an expected points framework to evaluate play calling in the NFL. 
-However, rather than evaluate each play on its own, the framework presented here evaluates the play in the context of the drive.
+However, rather than evaluate each play on its own, they evaluate the play in the context of the drive. 
 Based on play-by-play data from 2005 through 2008, they find that teams are underutilizing passing plays in some situations.
 
-@urschel2011 uses expected points to evaluate NFL coaching decisions on kickoffs.
-Specifically, they look at surprise on-sides kicks vs regular kickoff and the decision to accept a touchback vs returning kickoff.
+@urschel2011 use expected points to evaluate NFL coaching decisions on kickoffs. 
+Specifically, they look at surprise on-sides kicks vs. regular kickoff and the decision to accept a touchback vs. returning the kickoff. 
 Using data from the 2009 NFL season, the conclude, as many have, that coaches in the NFL tend to make conservative decisions.
@@
 
-```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+```{r chess, eval=whoami::username() == 'bbadumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}

### 2022-07-28 sports_analytics.Rmd
first draft of #13

@@
 
-## Competitions
+Public research in sports analytics is driven in part by several notable competitions. 
+These competitions have been an important source of new ideas and have contributed to the diversification of the field by breaking down barriers to entry.
 
-- Kaggle March Madness [@lopez2015building] (jqas special issue: https://www.degruyter.com/journal/key/jqas/11/1/html)
-- Big Data Bowl (special issue https://www.degruyter.com/journal/key/jqas/16/2/html)
-- Big Data Cup
-- SABR Case competition
+In 2014, the *Journal of Quantitative Analysis in Sports* teamed up with Kaggle to create a March Madness competition to predict the winners of the NCAA men's basketball tournament. 
+243 entrants competed for the $15,000 cash prize by submitting predicted probabilities for every possible pairwise matchup among the 68 teams in the tournament [@GlickmanSonas2015]. 
+In the [special issue](https://www.degruyter.com/journal/key/jqas/11/1/html) that *JQAS* subsequently published, we learned that the winning entry was based on a fairly simple logistic regression model trained on betting market data [@lopez2015building].
+Thus, the competition not only sparked interest in sports analytics, but resulted in peer-reviewed research that, in that case, demonstrated the value of superior data over sophisticated modeling.
 
-## Conferences
+Perhaps motivated by his success in the Kaggle March Madness competition, Michael Lopez joined the National Football League and launched the [Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/) in 2019.
+This annual competition has similarly fueled new research directions in American football and a [special issue](https://www.degruyter.com/journal/key/jqas/16/2/html) of *JQAS* [@lopez2020bigger].
+Successful entries and their corresponding publications [@DeshpandeEvans2020; @Yurko2020going; @ChuReyersThomsonWu2020] have launched the careers of several of the most prominent early-career researchers in sports analytics.
 
-- NESSIS/CASSIS
-- Sloan
-- CMU
-- UConn sports analytics
-
-@marchi2018analyzing
-
-@R-Lahman, @R-pitchRx, @R-retrosheet
+Similar competitions include the [Big Data Cup](https://www.stathletes.com/big-data-cup/) for ice hockey and the [SABR Diamond Dollars Case Competition](https://sabr.org/analytics/case) for baseball.
 
@@
 - How significant is the element of chance in a particular sport? 
-- Given that we know how the best teams are, who are the best players and how can we quantify their relative contributions? 
+- Given that we know who the best teams are, who are the best players and how can we quantify their relative contributions? 
 - What combinations of players work best together in a particular sport?

### 2022-07-29 sports_analytics.Rmd
Q edits

@@
 
-@pelechrinis2019 develops an expected points framework in the same spirit as previous work, but accounts for the strength of the opponents in their method.
+In addition, @pelechrinis2019 develops an expected points framework in the same spirit as previous work, but accounts for the strength of the opponents in their method.
 They state that by failing to account for opponent strength appropriately, about 124.8 points per team each season (or about 3.8 wins per season) are not credited correctly.
@@
 One of the most notable and well-studied is the use of expected points in the evaluation of 4th down strategy, and there is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.
+
 @Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on examples from the first quarter of a game (to avoid issues with end-of-half and end-of-game decision making).
-They conclude that teams don't go for it enough if they are trying to maximize their probability of winning the game.
+The author concludes that teams don't go for it enough if they are trying to maximize their probability of winning the game.
 
@@
 
+Researchers have also applied the notion of expected points to evaluate other aspects of the game of football, including quarterback performance and coaching decision.
+
 @White2002 present a tiered logistic regression method that can be, in general, applied to any regression setting with a polychotomous response.
 Using this technique, they estimate the value of NFL plays using a simple expected points model with down, yards to go, and yards to goal as predictors.
-They then use these results to evaluate and rank NFL quarterbacks.
+Accordingly, the model results can be used to obtain ratings and rankings for NFL passers.
 
-@Alamar2010 uses an expected points framework to evaluate play calling in the NFL. 
-However, rather than evaluate each play on its own, they evaluate the play in the context of the drive. 
-Based on play-by-play data from 2005 through 2008, they find that teams are underutilizing passing plays in some situations.
+In addition, @Alamar2010 implements an expected points framework to examine play calling in the NFL. 
+However, rather than assessing each play on its own, the author evaluates the play in the context of the drive. 
+Based on play-by-play data from 2005 through 2008, it is determined that teams are under-utilizing passing plays in some situations.
 
-@urschel2011 use expected points to evaluate NFL coaching decisions on kickoffs. 
-Specifically, they look at surprise on-sides kicks vs. regular kickoff and the decision to accept a touchback vs. returning the kickoff. 
+Another application of expected points is to evaluate kickoff decisions made by football coaches, as demonstrated by @urschel2011.
+Specifically, they look at surprise on-sides kicks versus regular kickoff and the decision to accept a touchback versus returning the kickoff. 
 Using data from the 2009 NFL season, the conclude, as many have, that coaches in the NFL tend to make conservative decisions.
@@
 
-The notion of EPV has also been extended to association football (soccer.)
+The notion of EPV has also been extended to association football (soccer).
 @fernandez2021framework implement deep learning methods to examine the instantaneous expected value of soccer possessions. 
@@
 In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football. 
-
 These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require few assumptions, and produce feature importance scores.
@@
 In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
-
 This model assumes independence of observations and constant variability in win probability.
 
-In addition, @McFarlane2019 uses logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA. 
+Moreover, @McFarlane2019 uses logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA. 
 The approach takes into account the remaining game time, score difference, and point spread. 
@@
 Many tools in sports analytics provide the ability to read, write, and plot data stored in a sport-specific format. 
-For example, consider chess, where the sequence of moves in games is often recorded in [Portable Game Notation](https://en.wikipedia.org/wiki/Portable_Game_Notation).
+For example, consider chess, where the sequence of moves in games is often recorded in [Portable Game Notation](https://en.wikipedia.org/wiki/Portable_Game_Notation) (PGN).
 Software tools can then be built around this well-defined format. 
 The \pkg{chess} package [@R-chess] provides R users with the ability to read, write, display, and manipulate chess data in PGN format.
-[Application programming interfaces](https://en.wikipedia.org/wiki/API) (APIs) are also common. 
+
+[Application programming interfaces](https://en.wikipedia.org/wiki/API) (APIs) are also a common source for data retrieval. 
 In chess, the \pkg{chessR} package [@R-chessR] allows R users to download game data from the Chess.com API. 
@@
 The \pkg{teamcolors} package [@R-teamcolors] provides color palettes and logos for men's and women's professional and collegiate sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
-The team colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
+The NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
 Figure \ref{fig:teamcolors} illustrates the full array of team color palettes for professional sports leagues.
@@
 
-In 2014, the *Journal of Quantitative Analysis in Sports* teamed up with Kaggle to create a March Madness competition to predict the winners of the NCAA men's basketball tournament. 
-243 entrants competed for the $15,000 cash prize by submitting predicted probabilities for every possible pairwise matchup among the 68 teams in the tournament [@GlickmanSonas2015]. 
-In the [special issue](https://www.degruyter.com/journal/key/jqas/11/1/html) that *JQAS* subsequently published, we learned that the winning entry was based on a fairly simple logistic regression model trained on betting market data [@lopez2015building].
-Thus, the competition not only sparked interest in sports analytics, but resulted in peer-reviewed research that, in that case, demonstrated the value of superior data over sophisticated modeling.
+In 2014, Kaggle launched its first March Machine Learning Mania competition for predicting the outcome of the NCAA men's basketball tournament. 
+243 entrants competed for the $15,000 cash prize by submitting predicted probabilities for every possible pairwise matchup among the 68 college basketball teams in the tournament [@GlickmanSonas2015].
+Subsequently, the *Journal of Quantitative Analysis in Sports* (JQAS) released a [special issue on prediction methodology for the NCAA men's basketball tournament](https://www.degruyter.com/journal/key/jqas/11/1/html).
+Among the published papers, we learned that the winning entry was based on a fairly simple logistic regression model trained on betting market data [@lopez2015building].
+Thus, the competition not only sparked interest in sports analytics, but also resulted in peer-reviewed research which, in that case, demonstrated the value of superior data over sophisticated modeling.
 
 Perhaps motivated by his success in the Kaggle March Madness competition, Michael Lopez joined the National Football League and launched the [Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/) in 2019.
-This annual competition has similarly fueled new research directions in American football and a [special issue](https://www.degruyter.com/journal/key/jqas/16/2/html) of *JQAS* [@lopez2020bigger].
+This annual competition has similarly fueled new research directions in American football and a [special issue on player tracking data in the National Football League (NFL)](https://www.degruyter.com/journal/key/jqas/16/2/html) of *JQAS* [@lopez2020bigger].
 Successful entries and their corresponding publications [@DeshpandeEvans2020; @Yurko2020going; @ChuReyersThomsonWu2020] have launched the careers of several of the most prominent early-career researchers in sports analytics.
 
-Similar competitions include the [Big Data Cup](https://www.stathletes.com/big-data-cup/) for ice hockey and the [SABR Diamond Dollars Case Competition](https://sabr.org/analytics/case) for baseball.
+Similar competitions that provide opportunities for aspiring researchers to tackle sports analytics problems include the [Big Data Cup](https://www.stathletes.com/big-data-cup/) for ice hockey and the [SABR Diamond Dollars Case Competition](https://sabr.org/analytics/case) for baseball.
 

### 2022-07-29 sports_analytics.Rmd
added to expected points

@@
 
-Researchers have also applied the notion of expected points to evaluate other aspects of the game of football, including quarterback performance and coaching decision.
+Researchers have also applied the notion of expected points to investigate other aspects of the game of football, including quarterback performance and coaching decision.
 
-@White2002 present a tiered logistic regression method that can be, in general, applied to any regression setting with a polychotomous response.
+As for quarterback evaluation, @White2002 present a tiered logistic regression method that can be, in general, applied to any regression setting with a polychotomous response.
 Using this technique, they estimate the value of NFL plays using a simple expected points model with down, yards to go, and yards to goal as predictors.
 Accordingly, the model results can be used to obtain ratings and rankings for NFL passers.
+
+Expected point value is also the main component of a novel quarterback evaluation metric introduced by @Reyers2021quarterback. 
+The authors take advantage of player tracking data to account for different passing and running options available to the quarterback.
+The expected points and probabilities associated with the available options are estimated using the method of ensemble learning via stacking.
 
@@
 
-The BTM can be implemented in `R` [@R-base] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. 
+The BTM can be implemented in R via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. 
 As an example, we consider the data given in @agresti2018introduction (page 247) on tennis results from 2014--2018 for five men's professional players: Novak Djokovic, Roger Federer, Andy Murray, Rafael Nadal, and Stan Wawrinka.
@@
 Perhaps motivated by his success in the Kaggle March Madness competition, Michael Lopez joined the National Football League and launched the [Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/) in 2019.
-This annual competition has similarly fueled new research directions in American football and a [special issue on player tracking data in the National Football League (NFL)](https://www.degruyter.com/journal/key/jqas/16/2/html) of *JQAS* [@lopez2020bigger].
+This annual competition has similarly fueled new research directions in American football and a [*JQAS* special issue on player tracking data in the National Football League (NFL)](https://www.degruyter.com/journal/key/jqas/16/2/html) of [@lopez2020bigger].
 Successful entries and their corresponding publications [@DeshpandeEvans2020; @Yurko2020going; @ChuReyersThomsonWu2020] have launched the careers of several of the most prominent early-career researchers in sports analytics.

### 2022-07-29 sports_analytics.Rmd
typos

@@
 In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a game state at any given point in a sporting contest. 
-Mathematically, we define $X$ to be a random variable indicating the the number of points (or runs) that a team will score over some determined amount of time (e.g.
+Mathematically, we define $X$ to be a random variable indicating the number of points (or runs) that a team will score over some determined amount of time (e.g.
 remainder of game, quarter, period, or inning).
@@
 ```{r lindsey, out.width="90%", fig.cap="Table 1 from Lindsey's original paper.
-The column labelled $E(T, B)$ gives the expected run matrix as a vector, based on Lindsey's analysis of Major League Baseball data from 1959 and 1960.
+The column labeled $E(T, B)$ gives the expected run matrix as a vector, based on Lindsey's analysis of Major League Baseball data from 1959 and 1960.
 "}
@@
 For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.
-@Tango2007book (and many subsequent analyses), conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
+@Tango2007book (and many subsequent analyses) conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
 
@@
 In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
-For any given play, the expected points is calculated using the absorption probabilities for different scoring events.
+For any given play, the expected points are calculated using the absorption probabilities for different scoring events.
 
@@
 
-The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes.
+The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal down situation and whether there are less than two minutes remaining in the half.
 This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.
@@
 
-Recently, @lopez2020bigger presented an introduction to NFL tracking data, and as an example they examine 4th down behavior.  
+Recently, @lopez2020bigger presented an introduction to NFL tracking data, and as an example, they examine 4th down behavior.
 Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.
-They demonstrate than teams are still conservative when choosing to go for it or punt on 4th down, but the effect may be overstated.
+They demonstrate that teams are still conservative when choosing to go for it or punt on 4th down, but the effect may be overstated.
 
@@
 
-Researchers have also applied the notion of expected points to investigate other aspects of the game of football, including quarterback performance and coaching decision.
+Researchers have also applied the notion of expected points to investigate other aspects of the game of football, including quarterback performance and coaching decisions.
 
@@
 Using this technique, they estimate the value of NFL plays using a simple expected points model with down, yards to go, and yards to goal as predictors.
-Accordingly, the model results can be used to obtain ratings and rankings for NFL passers.
+Accordingly, the model results are utilized to obtain ratings and rankings for NFL passers.
 
@@
 Specifically, they look at surprise on-sides kicks versus regular kickoff and the decision to accept a touchback versus returning the kickoff. 
-Using data from the 2009 NFL season, the conclude, as many have, that coaches in the NFL tend to make conservative decisions.
+Using data from the 2009 NFL season, they conclude, as many have, that coaches in the NFL tend to make conservative decisions.
 
@@
 
-The notion of EPV has also been extended to association football (soccer).
+The notion of expected possession value has also been extended to association football (soccer).
 @fernandez2021framework implement deep learning methods to examine the instantaneous expected value of soccer possessions. 
-This approach considers passes, ball drives, and shots in soccer as the main set of actions used to compute expected possession value. 
+This approach considers passes, ball drives, and shots in soccer as the main set of actions used to compute the EPV metric. 
 Many applications can be derived from this framework, including predicting which footballer on the pitch is most likely to receive the next pass from the current on-ball player.
@@
 Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
-Table \ref{tab:tied} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the the probability of scoring zero runs is 0.13. 
+Table \ref{tab:tied} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13. 
 The defense is in a tight spot, facing an 87% probability of losing the game. 
@@
 For example, in hockey teams will often pull their goalie when trailing in the final period. 
-This strategy severely weakens their defense, but strengthens their offense. 
+This strategy severely weakens their defense but strengthens their offense. 
 The hope is to score a quick goal to get back in the game, but the risk is falling further behind. 
 @beaudoin2010strategies show that NHL coaches do not always employ the optimal strategies. 
-@skinner2011scoring develops a general framework for these desperation strategies, which include the onside kick in American football, pulling the infield and/or outfield in in baseball, and of course, the fabled Hack-a-Shaq strategy in basketball.
+@skinner2011scoring develops a general framework for these desperation strategies, which include the onside kick in American football, pulling the infield and/or outfield in baseball, and of course, the fabled Hack-a-Shaq strategy in basketball.
 
@@
 
-A related, but different concept is the notion of *win probability*.
+A related, but different concept to expected points is the notion of *win probability*.
 Win probability is simply an estimate of the probability that a team will win the game, given its current state $s$. 
@@
 $$
-where the win and loss points represent how much the player positively and negatively impact their team's probability of winning after each play.
+where the win and loss points represent how much the player positively and negatively impacts their team's probability of winning after each play.
 In effect, the win points are the sum of the changes in $\Pr[W_i | s]$ from one state to the next.
@@
 
-@Lock2014 use a random forest method to provide a win probability estimate before each play in a football game. 
+@Lock2014 implement a random forest method to provide a win probability estimate before each play in a football game. 
 Covariates included in this tree-based win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yard line, and yards to go for a first down.
@@
 
-Figure \ref{fig:nflfastr-wp} shows a win probability graph for the January 14, 2018 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. 
+Figure \ref{fig:nflfastr-wp} shows a win probability graph for the 2018 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings on January 14, 2018. 
 We obtain the estimated probability of winning for each team using the \pkg{nflfastR} R package, which implements a gradient boosting model via the \pkg{xgboost} library (@R-xgboost) for estimating win probabilities.
@@
 @Glickman1998 propose a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
-In particular, they model point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths. 
+In particular, they model point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources of variation in team strengths. 
 See also @glickman2017estimating for more discussion on estimating team strengths in American football. 
@@
 
-```{r chess, eval=whoami::username() == 'bbadumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}

### 2022-07-31 sports_analytics.Rmd
more light edits

@@
   Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years.
-  We connect three big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
-  While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and sources of data. 
+  In this paper, we connect three big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
+  While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources. 
   Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
@@
 While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the deepest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact.
-
 In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
@@
 In this paper, we explore three key ideas that have widespread applicability across many sports: the expected value of a game state (Section \ref{sec:ev}), win probability (Section \ref{sec:wp}), and measures of team strength (Section \ref{sec:strength}).
-In each case, we define the concept mathematically, explain how it originated, and give examples of how it is applied in multiple sports. 
+In each case, we define the concept mathematically, explain how it originated, and give examples of its applications in multiple sports. 
 Our goal is to unify the conceptual threads, while doing some justice to the customizations necessary to make a metric meaningful in a particular sport. 
@@
 The concept of estimating the value of the state of a game is easily extended to other sports. 
-For example, in American football, $s$ is determined by situational variables at the start of a play such as down, yardage to the next first down, time remaining in the game, and field position.
+For example, in American football, $s$ is determined by situational variables such as down, yardage to the next first down, time remaining in the game, and field position.
 
 The first effort of estimating expected points of possession in football goes back to @Carter1971operations. 
-In this technical note, the authors estimate the expected points for 1st and 10 plays in the NFL given any yard line on the football field. 
-Due to limitations regarding the amount of data collected, the football field is broken into 10-yard buckets, centered at their midpoints (e.g.
+In this technical note, the expected points for 1st and 10 plays in the NFL are estimated, given any yard line on the football field. 
+Due to limitations regarding the amount of data collected, the authors divide football field into 10-yard buckets, centered at their midpoints (e.g.
 5, 15, 25, 35, etc.), before averaging the value of the next scoring instance across the field to obtain the expected points.
 
-Other early work on expected point values in American football can be found in @Carroll1988hidden. 
+Early work on expected point values in American football can also be found in @Carroll1988hidden.
 In particular, the authors extend @Carter1971operations's work and propose a linear model for expected points in the NFL.
@@
 
-@goldner2012markov and @goldner2017situational propose a Markov model for estimating expected points in football. 
+Another attempt at modeling expected points in football is done by @goldner2012markov and @goldner2017situational, which proposes a Markov framework.
 In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
@@
 Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
-
 The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal down situation and whether there are less than two minutes remaining in the half.
@@
 The concept of expected points in American football has many applications.
-One of the most notable and well-studied is the use of expected points in the evaluation of 4th down strategy, and there is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.
+For example, one of the most notable and well-studied topic is the evaluation of 4th down strategy.
+There is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.
 
@@
 Accordingly, the model results are utilized to obtain ratings and rankings for NFL passers.
-
-Expected point value is also the main component of a novel quarterback evaluation metric introduced by @Reyers2021quarterback. 
-The authors take advantage of player tracking data to account for different passing and running options available to the quarterback.
-The expected points and probabilities associated with the available options are estimated using the method of ensemble learning via stacking.
 
@@
 Accordingly, the probability of various types of outcomes at the end of a play such as a touchdown or a first-down gain can be computed from the distribution of the end-of-play yard line.
+
+Expected point value is also the main component of a novel NFL quarterback evaluation metric introduced by @Reyers2021quarterback. 
+The authors take advantage of player tracking data to account for different passing and running options on the football field that are available to the quarterback.
+The expected points and probabilities associated with the possible quarterback options are estimated using the method of ensemble learning via stacking.
 
@@
 In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football. 
-These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require few assumptions, and produce feature importance scores.
+These are flexible algorithms that have high predictability, can account for nonlinear interactions between the explanatory variables, require few assumptions, and produce feature importance scores.
 
-@Lock2014 implement a random forest method to provide a win probability estimate before each play in a football game. 
-Covariates included in this tree-based win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yard line, and yards to go for a first down.
+@Lock2014 implement a random forest framework to provide a win probability estimate before each play in a football game. 
+Covariates included in this tree-based method are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yard line, and yards to go for a first down.
 According to this model, the difference in score between the two teams is the most important feature for predicting win probabilities at any given moment in an NFL game.
@@
 In particular, the authors propose both homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game.
-In esports, @maymin2021smart uses logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
+In esports, @maymin2021smart relies on logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
 Moreover, @Guan2022 develop an in-game win probability model for the National Rugby League using functional data analysis. 
-In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.
+In this approach, the rugby play-by-play event data are treated as functional, and the win probability is expressed as a function of the match time.
 
@@
 Each professional sports team has its own brand, most obviously identified by a team logo and set of colors. 
-The \pkg{teamcolors} package [@R-teamcolors] provides color palettes and logos for men's and women's professional and collegiate sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
-The NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
+The \pkg{teamcolors} R package [@R-teamcolors] provides color palettes and logos for men's and women's professional and collegiate sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
+For instance, the NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
 Figure \ref{fig:teamcolors} illustrates the full array of team color palettes for professional sports leagues.
@@
 Among the published papers, we learned that the winning entry was based on a fairly simple logistic regression model trained on betting market data [@lopez2015building].
-Thus, the competition not only sparked interest in sports analytics, but also resulted in peer-reviewed research which, in that case, demonstrated the value of superior data over sophisticated modeling.
+Thus, the competition not only sparked interest in sports analytics, but also resulted in peer-reviewed research which, in that case, demonstrated the value of quality data over sophisticated modeling.
 

### 2022-08-01 sports_analytics.Rmd
editing

@@
 Similar insights are now being made in less obvious settings, including esports [@clark2020bayesian; @maymin2021smart]. 
-These insights come both from academia, where researchers typically use public data to produce high-caliber, peer-reviewed scientific work, as well as from industry, where highly-trained analysts work with with players, coaches, and team officials to gain immediate effect thanks to high-resolution, often proprietary data.
+These insights come both from academia, where researchers typically use public data to produce high-caliber, peer-reviewed scientific work, as well as from industry, where highly-trained analysts work with with players, coaches, and team officials to put new ideas into immediate effect thanks to high-resolution, often proprietary data.
 A growing pool of people move seamlessly between these two worlds, leading to the formation of partnerships and the cross-pollination of ideas.
@@
 
-In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a game state at any given point in a sporting contest. 
-Mathematically, we define $X$ to be a random variable indicating the number of points (or runs) that a team will score over some determined amount of time (e.g.
-remainder of game, quarter, period, or inning).
+In many sports, the first step towards an analytical understanding is the estimation of the expected value of a game state at any given point in it. 
+Mathematically, we define $X$ to be a random variable indicating the number of points (or runs) that a team will score over some determined amount of time (e.g., remainder of game, quarter, period, or inning).
 Let $s \in S$ be a tuple that encodes the *state* of a game. 
 Then our task is to estimate:
-$$
+
+\begin{equation}
   \mathbb{E}[X | s] = \sum_{X \geq 0} \Pr[X | s] \cdot X\,,
-$$
+\end{equation}
+
 for any state $s \in S$, where $\Pr[X | s]$ is the probability of scoring $X$ points given that the game is in state $s$ and $S$ is the set of all possible states.
@@
 Note how (when reading across the rows) the expected runs decrease as outs increase for the same configuration of baserunners, while (when reading down the columns) expected runs generally increase as baserunners advance.
+000 means no runners on base, and 110 means runners on second and third bases.
 \\label{tab:erm}") |>
@@
 In a high run-scoring environment, where there are many home runs, the value of a walk may be higher, since a player who walks is more likely to score on a subsequent home run. 
-Conversely, in a low run-scoring environment, stolen bases and sacrifice bunts may be comparatively more valuable. 
+Conversely, in a low run-scoring environment where hits are hard to come by, stolen bases and sacrifice bunts may be comparatively more valuable. 
 Thus, a careful estimate of $\mathbb{E}[X | s]$ would include a time parameter $t$, indicating when the estimate is appropriate.
@@
 
-The first effort of estimating expected points of possession in football goes back to @Carter1971operations. 
-In this technical note, the expected points for 1st and 10 plays in the NFL are estimated, given any yard line on the football field. 
+The task of estimating expected points of possession in football goes back to @Carter1971operations, who estimate the expected points for 1st and 10 plays in the NFL, given any yard line on the football field.
 Due to limitations regarding the amount of data collected, the authors divide football field into 10-yard buckets, centered at their midpoints (e.g.
@@
 
+XX need example.
+
 Early work on expected point values in American football can also be found in @Carroll1988hidden.
 In particular, the authors extend @Carter1971operations's work and propose a linear model for expected points in the NFL.
-It is determined that every extra 25 yards is associated with 2 more points scored on average for a football team.
+They determine that every extra 25 yards is associated with 2 more points scored on average for a football team.
 
-Another attempt at modeling expected points in football is done by @goldner2012markov and @goldner2017situational, which proposes a Markov framework.
+Other attempts at modeling expected points in football are @goldner2012markov and @goldner2017situational, who propose a Markov framework.
 In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
@@
 A more in-depth overview of the history of expected points in sports is provided in @Yurko2019nflwar (Section 1.1). 
-Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
-The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal down situation and whether there are less than two minutes remaining in the half.
-This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.
-The expected points for a play can then be calculated accordingly, by summing up the products of the scoring event point values and their associated probabilities.
+Most importantly, @Yurko2019nflwar use publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
+The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and indicators for goal down situation and whether there are less than two minutes remaining in the half.
+Their model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.
+The expected points for a play can then be calculated accordingly, by summing up the products of the scoring event point values and their associated probabilities (see Equation 1).
 
-In addition, @pelechrinis2019 develops an expected points framework in the same spirit as previous work, but accounts for the strength of the opponents in their method.
+In addition, @pelechrinis2019 develop an expected points framework in the same spirit as the previous work, but account for the strength of the opponents in their method.
 They state that by failing to account for opponent strength appropriately, about 124.8 points per team each season (or about 3.8 wins per season) are not credited correctly.
+
+XX Seems massive
 
@@
 @Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on examples from the first quarter of a game (to avoid issues with end-of-half and end-of-game decision making).
-The author concludes that teams don't go for it enough if they are trying to maximize their probability of winning the game.
+He concludes that teams don't go for it enough if they are trying to maximize their probability of winning the game.
 
@@
 
-Recently, @lopez2020bigger presented an introduction to NFL tracking data, and as an example, they examine 4th down behavior.
+Recently, @lopez2020bigger presents an introduction to NFL tracking data, and as an example, examines 4th down behavior.
 Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.
-They demonstrate that teams are still conservative when choosing to go for it or punt on 4th down, but the effect may be overstated.
+They demonstrate that teams are still conservative when choosing to go for it or punting on 4th down, but the effect may be overstated.
 
@@
 
-As for quarterback evaluation, @White2002 present a tiered logistic regression method that can be, in general, applied to any regression setting with a polychotomous response.
+For quarterback evaluation, @White2002 present a tiered logistic regression method that can be, in general, applied to any regression setting with a polychotomous response.
 Using this technique, they estimate the value of NFL plays using a simple expected points model with down, yards to go, and yards to goal as predictors.
@@
 However, rather than assessing each play on its own, the author evaluates the play in the context of the drive. 
-Based on play-by-play data from 2005 through 2008, it is determined that teams are under-utilizing passing plays in some situations.
+Based on play-by-play data from 2005 through 2008, he determines that teams are under-utilizing passing plays in some situations.
 
@@
 See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
+The strategy of maximizing expected points in basketball has lead directly to the proliferation of three-point shooting in the NBA.
 
-In American football, @Yurko2020going use tracking data provided by the [2019 NFL Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) to model the expected yards gained for a ball-carrier during the course of a play.
+In American football, @Yurko2020going use tracking data provided by the [2019 NFL Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) (see Section \ref{sec:opportunities}) to model the expected yards gained for a ball-carrier during the course of a play.
 As an extension to pre-existing approaches, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play, rather than only producing a single estimate for the expected yards gained.
@@
 For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs.
-If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy $u$ that maximizes the expected number of points will maximize the probability of winning:
+If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy $u \in U$ that maximizes the expected number of points will maximize the probability of winning:
 $$
@@
 Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
-Table \ref{tab:tied} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13. 
+Table \ref{tab:tied} (derived from Table \ref{tab:erm}) reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13.
 The defense is in a tight spot, facing an 87% probability of losing the game. 
 However, by walking the hitter to load the bases, they create the opportunity to force the lead runner at home and thus reduce the chances of scoring to 82%, even though they raise the expected number of runs scored to 2.22.
-In this case, the defensive team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.
+In this case, the *defensive* team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.
 
@@
   mutate(Base = int2bin(Base)) |>
-  kable(booktabs = TRUE, 
+  rename(`Pr(Scoring)` = Pr_0) |>
+  kable(
+    booktabs = TRUE, 
     caption = "Loading the bases can be a rational strategy.
-In this case, loading the bases maximizes both the expected number of runs scored *and* the probability of not scoring at all.
-\\label{tab:tied}")
+In this case, loading the bases maximizes both the expected number of runs scored and the probability of not scoring at all.
+\\label{tab:tied}"
+  )
 ```
@@
 
-The notion of win probability in baseball goes back to at least as early as @Lindsey1961.
-In this paper, the author calculates the expected win probability after each inning based on the distribution of runs scored in each inning. 
+The notion of win probability in baseball goes back to at least as early as @Lindsey1961, who calculates the expected win probability after each inning based on the distribution of runs scored in each inning.
 Inspired by @Lindsey1961's work, @Mills1970 utilize win probability to introduce Player Win Average (PWA), a measure of a player's contribution to the game outcome. 
@@
 $$
-Here, $\exp{\beta_i}$ is often viewed as a representation of player $i$'s ability.
+Here, $\exp{\beta_i}$ is often viewed as a representation of team $i$'s ability.
 
@@
 According to the abilities, between 2014 and 2018 the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray.
-To obtain win probabilities, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is
+In addition to an ordering, the magnitude of the coefficients in Table \ref{tab:tennisbtm} provide a measure of relative strength.
+
+To obtain win probabilities, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is:
 $$
-\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768.
+\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768 \,.
 $$
@@
 Betting moneylines are used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
+Both papers motivate the usefulness of model-based measures of team strength by demonstrating their superiority to low-resolution win-loss records. 
+Apart from sports gambling, having an accurate estimate of team strength is useful to team officials, who are constantly monitoring and forecasting their team's ability.
 
@@
 
-```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+```{r chess, eval=whoami::username() == 'bbsdaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
@@
 
-## Colors
+## Theming
 
@@
 The \pkg{teamcolors} R package [@R-teamcolors] provides color palettes and logos for men's and women's professional and collegiate sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
-For instance, the NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
+For example, the NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
 Figure \ref{fig:teamcolors} illustrates the full array of team color palettes for professional sports leagues.
@@
 
-# Opportunities
+# Opportunities {#sec:opportunities}
 
@@
 Perhaps motivated by his success in the Kaggle March Madness competition, Michael Lopez joined the National Football League and launched the [Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/) in 2019.
-This annual competition has similarly fueled new research directions in American football and a [*JQAS* special issue on player tracking data in the National Football League (NFL)](https://www.degruyter.com/journal/key/jqas/16/2/html) of [@lopez2020bigger].
+This annual competition has similarly fueled new research directions in American football and a [*JQAS* special issue on player tracking data in the National Football League](https://www.degruyter.com/journal/key/jqas/16/2/html) [@lopez2020bigger].
 Successful entries and their corresponding publications [@DeshpandeEvans2020; @Yurko2020going; @ChuReyersThomsonWu2020] have launched the careers of several of the most prominent early-career researchers in sports analytics.
@@
 However, as a maturing discipline, sports analytics has been able to address fundamental questions common to many sports. 
-In this paper, we explicate three of those big questions:
+In this paper, we explore three of those big questions:
 
@@
 
-# Acknowledgments
+# Acknowledgments {.unnumbered}
 

### 2022-08-01 sports_analytics.Rmd
added figure

@@
 5, 15, 25, 35, etc.), before averaging the value of the next scoring instance across the field to obtain the expected points.
+Figure \ref{fig:carter} shows a reproduction of @Carter1971operations's estimates. 
+As expected, the estimated expected points increases monotonically as the teams gets closer to the endzone.
 
-XX need example.
+```{r carter, fig.cap="Table 1 from Carter and Machol's original paper. Note the monotonic increase in expected point values as the team gets closer to the endzone. ", out.width="100%"}
+knitr::include_graphics("carter_table1.png")
+```
 

### 2022-08-01 sports_analytics.Rmd
Changed "deepest" to "longest"

@@
 Insights derived from the analysis of data have transformed the world of sports over the last few decades. 
-While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the deepest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact.
+While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the longest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact.
 In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 

### 2022-08-01 sports_analytics.Rmd
Corrected the expected value notation.

@@
 \begin{equation}
-  \mathbb{E}[X | s] = \sum_{X \geq 0} \Pr[X | s] \cdot X\,,
+  \mathbb{E}[X | s] = \sum_{x \geq 0} \Pr[X = x | s] \cdot x\,,
 \end{equation}
 
-for any state $s \in S$, where $\Pr[X | s]$ is the probability of scoring $X$ points given that the game is in state $s$ and $S$ is the set of all possible states.
+for any state $s \in S$, where $\Pr[X = x | s]$ is the probability of scoring $x$ points given that the game is in state $s$ and $S$ is the set of all possible states.
 

### 2022-08-01 sports_analytics.Rmd
Notation chance on expected value.

@@
 In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
-Thus, there $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
+Thus, there are $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning.
 In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 

### 2022-08-01 sports_analytics.Rmd
More Minor edits.

@@
 The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
-Figure \ref{fig:lindsey} shows a reproduction of George Lindsey's original calculations and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
+Figure \ref{fig:lindsey} shows a reproduction of George Lindsey's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
 
@@
 
-XX Seems massive
+<!-- XX Seems massive -->
 
@@
 
-The concept of expected points in American football has many applications.
-For example, one of the most notable and well-studied topic is the evaluation of 4th down strategy.
-There is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.
+The concept of expected points in American football has many applications.  One of the most notable and well-studied topic is the evaluation of 4th down strategy.
+There is near universal consensus in the literature that NFL teams have been too conservative in the past when making 4th down decisions.
 
 @Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on examples from the first quarter of a game (to avoid issues with end-of-half and end-of-game decision making).
-He concludes that teams don't go for it enough if they are trying to maximize their probability of winning the game.
+They concluded that teams don't go for it enough if teams are trying to maximize their probability of winning the game.
 
-Numerous papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy.
+Numerous other papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy.
 In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down.
@@
 
-Recently, @lopez2020bigger presents an introduction to NFL tracking data, and as an example, examines 4th down behavior.
+@lopez2020bigger presents an introduction to NFL tracking data, and as an example, examines 4th down behavior.
 Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.
@@
 
-In addition, @Alamar2010 implements an expected points framework to examine play calling in the NFL. 
-However, rather than assessing each play on its own, the author evaluates the play in the context of the drive. 
-Based on play-by-play data from 2005 through 2008, he determines that teams are under-utilizing passing plays in some situations.
+@Alamar2010 implements an expected points framework to examine play calling in the NFL. However, rather than assessing each play on its own, they evaluate the play in the context of the drive. 
+Based on play-by-play data from 2005 through 2008, they determine that teams are under-utilizing passing plays in some situations.
 
@@
 
-Even in sports where the concept of a state is not so easy to define, the value of a possession can be estimated with the help of tracking data. 
+Even in sports where the concept of a state is more difficult to define, the value of a possession can be estimated with the help of tracking data.
 Over the past decade or so, professional sports leagues have collected tracking data which record the locations of all players and the ball (or puck) throughout a game.
-This high-resolution data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game. 
+This high-resolution data allows researchers to produce advanced analyses of the captured spatiotemporal information and better understand the game.
 This is a great leap forward from older resources such as traditional box-score results and play-by-play data.

### 2022-08-01 sports_analytics.Rmd
Small edits.

@@
 
-Note that Miami starts the play with an EPV of approximately 1.0 points.
-This indicates their implied average points per possession. 
-Chalmers's shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
+Note that Miami starts the play with an EPV of approximately 1.0 points, which indicates their implied average points per possession. 
+Chalmers' shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
 A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court!

### 2022-08-02 sports_analytics.Rmd
Pr(no runs); close #23

@@
   mutate(Base = int2bin(Base)) |>
-  rename(`Pr(Scoring)` = Pr_0) |>
+  rename(`Pr(No Runs)` = Pr_0) |>
   kable(
@@
 
-```{r chess, eval=whoami::username() == 'bbsdaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}

### 2022-08-02 sports_analytics.Rmd
EPV is discrete; close #20
moved Lindsey citation; close #17

@@
 The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
-Figure \ref{fig:lindsey} shows a reproduction of George Lindsey's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
+Figure \ref{fig:lindsey} shows a reproduction of George @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
 
@@
 The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision. 
-The computation of the EPV statistic is done using a Markov model.
+The computation of the EPV statistic is done using a (technically discrete) Markov model conditioned on spatial locations.
 Consequently, the authors derive a metric called EPV-Added (EPVA), measuring a player's EPV contribution in a given situation relative to a league-average player.

### 2022-08-02 sports_analytics.Rmd
discrete vs continuous states; close #22

@@
 
-## In American football
+## In American football {#sec:football}
 
@@
 
-In American football, @Yurko2020going use tracking data provided by the [2019 NFL Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) (see Section \ref{sec:opportunities}) to model the expected yards gained for a ball-carrier during the course of a play.
+In Section \ref{sec:football}, we discussed advances in American football analytics based on discrete game states defined by down, yards to first down, field position, etc. 
+The advent of player tracking data makes it possible to analyze American football using continuous states. 
+For example, @Yurko2020going use tracking data provided by the [2019 NFL Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) (see Section \ref{sec:opportunities}) to model the expected yards gained for a ball-carrier during the course of a play.
 As an extension to pre-existing approaches, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play, rather than only producing a single estimate for the expected yards gained.

### 2022-08-02 sports_analytics.Rmd
aposthrophe; close #21

@@
 The plot consists of two elements: 
-1) (bottom) the player locations on the court at a particular moment in this possession: when the ball just left Chalmers' hands, 
+1) (bottom) the player locations on the court at a particular moment in this possession: when the ball just left Chalmers's hands, 
 and 2) (top) a line graph showing how the EPV changes continuously throughout the play until the three-point attempt. 

### 2022-09-28 response_to_reviewers.Rmd
fixed most of #25

---
title: "Response to Reviewers"
date: "2022-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Major comments from Reviewer 1

Comments to the Author(s)
Please see attached file for comments.

## Major comments from Reviewer 2

> While Sections 2 and 3 are well done and complete, Sections 4 and 5 are weak and seem incomplete and unpolished.

>Likewise the Opportunities section should be expanded to include the proliferation of Sports Analytics conferences.

> My largest concern is that a cursory discussion of player contribution or player ratings is given and this seems a major oversight of one of the “big ideas” in sports analytics given the focus on this topic in many sports.  See, for example, Mark Glickman’s recent presentation on this at the 2022 Joint Statistical Meetings.  I don’t think that Hvattum (2019) is sufficient to cover the area of player contribution.

## Minor comments from Reviewer 2

> Pages 3,4, 5: The use of historical tables and data is powerful.  However, I worry about the quality of the images particularly Figure 2.

We share this concern. We are also uncertain about any licensing restrictions. We have left the images as is for now, and will wait direction from Wiley.

> Page 6, line 8:  “the effect may be overstated” is somewhat ambiguous.  Is it the

XX Quang or Greg to respond

> Page 6, line 53: Is the “!” warranted here for conditioning on a 9-dimensional vector?

We have replaced the exclamation point with a period.

> Page 8, line 53: There is also work on expected possession in ice hockey, see Stokes et al. https://hockey-graphs.com/2021/07/06/bayesian-space-time-models-for-expected-possession-added-value-part-1-of-2/

We have added a reference to this work, but are also cognizant that Wiley policy may not permit a citation to this unpublished work:

>> "preprint articles should not be cited in WIREs manuscripts as review articles should discuss and draw conclusions only from peer-reviewed research. Remember that original research/unpublished work should also not be included as it has not yet been peer-reviewed and could put the work in jeopardy of getting published in the primary press."

> 9, 17: Is Player Win Average relevant to win probability?  This is more of a player contribution metric than a win probability metric.  I would rather see a brief exposition of Tango et al (2007) here.

Given that Albert (2015) previously published an article on this topic in this same journal series, we did not include a treatment of player contribution metrics in our paper. Our opinion is that this is redundant given Albert's article.

> 9, 47: Gridiron football is a lovely term but for this audience, I think it needs to be clarified.

We like the term too, but have changed this to "American football."

> 10, 52: “In this paper”, the “this” seems ambiguous.

We have replaced this with "Deshpande & Jensen"

> 11, 36: Table 3, why order this table alphabetically, why not by ability level?

We have ordered the rows of Table 3 by descending order of ability.

> 11, 50: Given how much fivethirtyeight.com uses Elo ratings for all kinds of sports, it is probably worth mentioning their work as examples.

We have added the sentence: "Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at FiveThirtyEight.com." But again, we acknowledge that Wiley policy may not allow such a reference.

> 12, 1: As noted above, Section 4 feels much weaker than Sections 2 and 3 which are strong.  Perhaps a look at Pythagorean methods as a historical introduction to this topic.

XX We could consider this.

> 12, 23: This is not much of a case study.  Perhaps some additional information on the data available, ways it could be used to calculate game state expected valued, win probability, player ratings….

XX Yeah

> 12, 37: Theming *teamcolors* is a great *R* package, not sure it is part of the “big ideas” in sports analytics.  A single line in the first paragraph of this section would seem to be better.

We have added examples from the `SportyR` package, which draws playing surfaces, to help flesh out this subsection.

> 13, 35: Add a citation for Big Data Cup, https://www.stathletes.com/big-data-cup/, for SABR Diamond Dollars Case Competition, https://sabr.org/analytics/case, Big Data Derby https://www.kaggle.com/competitions/big-data-derby-2022, and maybe more.

Thank you for these suggestions. We have added them.

> 13, 43-53: don’t think there is a need for a bulleted list here.

### 2022-09-28 sports_analytics.Rmd
fixed most of #25

@@
     "chess", 
-    "chessR"
+    "chessR",
+    "sportyR"
   ),
@@
 Chalmers' shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
-A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court!
+A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court.
 
@@
 Many applications can be derived from this framework, including predicting which footballer on the pitch is most likely to receive the next pass from the current on-ball player.
+
+@kumagai2021hockey presents an EPV metric for ice hockey.
 
@@
 The two win probability frameworks proposed by @Lock2014 and @Yurko2019nflwar were also implemented in @Yam2019 with minimal modifications.
-Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in gridiron football.
+Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in American football.
 
@@
 On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
-In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
+@Deshpande2016 propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
 This model assumes independence of observations and constant variability in win probability.
@@
   BradleyTerry2::BTabilities() |> 
-  round(3) |> 
-  kable(booktabs = TRUE, 
-        col.names = c("Ability", "SE"),
-        caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}") |> 
+  as_tibble(rownames = "Player") |>
+  arrange(desc(ability)) |>
+  mutate(
+    Ability = as.double(ability),
+    SE = as.double(s.e.)
+  ) |>
+  select(Player, Ability, SE) |>
+  kable(
+    booktabs = TRUE, 
+    digits = 3,
+    caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}"
+  ) |> 
   kable_styling(latex_options = "hold_position")
@@
 See @koning2017rating for more information on applications of the Elo rating in soccer. 
+Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at [FiveThirtyEight.com](https://fivethirtyeight.com).
 
@@
 
-```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+```{r chess, eval=whoami::username() == 'bsbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
@@
 
-## Theming
+## Graphical tools
+
+Creating effective data graphics is a key component of statistical communication, and sports is no exception. 
+We highlight a few packages that assist with the creation of data graphics about sports.
 
@@
 teamcolors::show_team_col()
+```
+
+Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, or pitch for a particular sport.
+Orienting these points graphically may require drawing a complex set of guidelines that provide that context to readers. 
+Thankfully, the \pkg{sportyR} package [@R-sportyR] contains generic playing surfaces for baseball, basketball, curling, American football, ice hockey, soccer, and tennis that can be added to \pkg{ggplot} graphics with a single function call.
+Consider, for example, how much time the surfaces plotted in Figure \ref{fig:sportyR} would save you, and how helpful they would be in contextualizing player tracking data (such as those shown in Figure \ref{fig:epv-demo}).
+With the increased availability of player tracking data, this particular tool should see increased usage.
+
+```{r sportyR, fig.show="hold", out.width="49%", fig.cap="At left, an NBA basketball court drawn by \\pkg{sportyR}. At right, an NHL hockey rink drawn by \\pkg{sportyR}."}
+library(sportyR)
+geom_basketball(league = "NBA")
+# geom_tennis(league = "USTA")
+geom_hockey(league = "NHL")
 ```
@@
 
-Similar competitions that provide opportunities for aspiring researchers to tackle sports analytics problems include the [Big Data Cup](https://www.stathletes.com/big-data-cup/) for ice hockey and the [SABR Diamond Dollars Case Competition](https://sabr.org/analytics/case) for baseball.
+Similar competitions that provide opportunities for aspiring researchers to tackle sports analytics problems include the [Big Data Cup](https://www.stathletes.com/big-data-cup/) for ice hockey and the [SABR Diamond Dollars Case Competition](https://sabr.org/analytics/case) for baseball, and the [Big Data Derby](https://www.kaggle.com/competitions/big-data-derby-2022) for horse racing.
 
@@
 In this paper, we explore three of those big questions: 
+Who are the best teams and how good are they? 
+What is the likelihood of each team winning the game at any given juncture? 
+Is there a generic framework for evaluating strategies at any given juncture in a game?
 
-- Who are the best teams and how good are they? 
-- What is the likelihood of each team winning the game at any given juncture? 
-- Is there a generic framework for evaluating strategies at any given juncture in a game?
-
-Other fundamental questions, such as:
-
-- How significant is the element of chance in a particular sport? 
-- Given that we know who the best teams are, who are the best players and how can we quantify their relative contributions? 
-- What combinations of players work best together in a particular sport?
-
-are addressed elsewhere. 
-In particular, see @Lopez2018 for estimations of the element of chance across four major sports. 
+Other fundamental questions are addressed elsewhere. 
+How significant is the element of chance in a particular sport? 
+Given that we know who the best teams are, who are the best players and how can we quantify their relative contributions? 
+What combinations of players work best together in a particular sport? In particular, see @Lopez2018 for estimations of the element of chance across four major sports. 
 The second question is often addressed using a formulation of *wins above replacement* (WAR)---see @baumer2015openwar and @Yurko2019nflwar for details in baseball and American football. 

### 2022-09-28 response_to_reviewers.Rmd
address #28; #30

@@
 I don’t think that Hvattum (2019) is sufficient to cover the area of player contribution.
+
+## Minor comments from Reviewer 1
+
+> In the second sentence, baseball is introduced as a “naturally discrete” sport, but there are certainly many things about baseball which are not discrete, as talked about later in the article.
+It would be helpful to clarify what is meant by discrete in this setting.
+
+We have added a paragraph clarify the distinction between "discrete" and "continuous" sports.
+In order to maintain the flow of the narrative, the paragraph does not appear here, but rather at the beginning of Section 2.
+
+> Also, later on in the first paragraph, it would be useful to have a citation or two about the changes being brought about in baseball due to analytics.
+
+We believe that the last sentence of the first paragraph does exactly that.
+
+> In Section 2.1, Figure 1 (George Lindsey’s calculations) is introduced before the reader even knows who George Lindsey is.
+
+We think this is mainly due to LaTeX floating the figure suboptimally, and this can be corrected in the page proofs.
+We have moved the sentence introducing Figure 1 and Table 1 until after the mention of Lindsey's work in the text to mitigate this problem.
+
+> Furthermore, the expected run matrix is displayed “in its more familiar form”, but I would venture for most readers of this article, no form would be familiar.
+
+We have replaced "more familiar" with "most common".
+
+> I think the caption does explain the matrix nicely, although it might be useful to point out from which column in Lindsey’s table the matrix is derived.
+
+Indeed, we note which column corresponds to the expected run matrix in the caption.
+
+> In Figure 2, I found it quite surprising that the expected point value was so negative     at the 95 yard line.
+Were safeties very common in this era of American football?
+This might be worth mentioning/discussing.
+
+XX: Greg or Quang to address?
+
+> The title of Section 2.3 seems to indicate that the aforementioned sports have only discrete states, but this is not always the case.
+Some clarification would be useful here.
+In general, the organization of Section 2 is confusing.
+It first seems to be going sport by sport, but then 2.3 is about several different sports and then 2.4 is on a general topic altogether.
+As is, it reads somewhat disjointed and is hard to follow.
+Section 3 flows much better in this sense.
+
+We thank the reviewer for this observation. We have reorganized the subsections of Section 2 with more explicit titles to improve the flow and readability of this Section.
+
+> Table 2 seems unnecessary, as the content is entirely summarized in the text.
+
+We have removed Table 2.
+
+> After the comprehensive views in Sections 2 and 3, Section 4 seems sparse on details.
+Much detail is given on the Bradley-Terry model, but only one example is given of a possible usage.
+Has this been used in other sports?
+Has it changed how we think about those sports or team strategies?
+An explanation of Elo rating should be given; I for one have no knowledge of how to compute it.
+
+We thank the reviewer for this suggestion. We have added a subsection on Elo rating.
+
+XX
+
+> There is a large body of literature on Bayesian state-space models in sports outside of the Glickman and Stern paper, mainly in soccer/association football.
+See for example Koopman and Lit (2015) and the various references in its introduction.
+
+Thank you for this suggestion. We have added a reference to Koopman's paper.
+
+> As mentioned earlier, Section 5 seems sparse on details, and furthermore it seems odd to choose chess as the case study when it has barely been mentioned previously in the article.
+I think this section could use a table of some sort, perhaps something that summarizes the CRAN task view at a high level.
+Another potential would be a figure which shows a typical workflow of sports analytics as explained in the text.
+
+XX
+
+> Figure 6 doesn’t add much to the text; I don’t think the reader better understands sports analytics by visualizing many team colors.
+Instead, it would be better to show  how the theming can lead to a more impactful graphic for sports, i.e.
+show how that package might be used in a real-world setting.
+
+XX
+
+> For Section 6, there is also now a “Big Data Derby” in horse racing.
+
+We thank the reviewer for this suggestion. We have added a reference to it.
+
+> All in all, I think the article explains the ideas well and covers a lot of ground, but more care should be taken towards the organization and what to include as figures.
+The last half of the paper has a very different pace than the first half.
+I think a common theme in the paper is the rise in tracking data and how that has impacted sports–perhaps that idea could be better unified and stated in a specific section of the article, or at the very least noted in the conclusion.
+
+XX
+
+> Page 4, line 56: Needs to be some sort of separation between the “absorbing states” and “touchdown”
+
+We have added "absorbing states that include touchdowns..."
+
+> Page 5, line 56: Typo for “causal”
+
+Fixed.
+
+> Page 5, line 57: Should be “too conservative”
+
+Fixed.
+
+> Page 7, line 37: Should be “led” not “lead”
+
+Fixed.
 

### 2022-09-28 sports_analytics.Rmd
address #28; #30

@@
 
-## In baseball, the expected run matrix
+The concept of a state is easier to grasp in a sport that can be modeled as *discrete* (in the sense of [discrete event simulation](https://en.wikipedia.org/wiki/Discrete-event_simulation)). 
+By discrete, we mean a sport that can be easily broken into short, distinct segments of action which are typically summarized categorically. 
+Each of these segments might represent a state $s$. 
+For example, each pitch in baseball is either a ball or a strike. 
+If the ball is put in play, then there may be a complex sequence of movements by the players, but ultimately (within a few seconds) that sequence will end and no more action will be permitted until the next pitch.
+At the beginning and end of each phase of action, we will know definitively which team is on offense and defense, which runners are on which bases, the score, how many outs there are, etc. 
+Tennis could similarly be viewed as a series of discrete actions defined by each point. 
+To say that a player is winning 6-2, 3-1, 40-15 and serving with one fault committed is to characterize the state of the match.
+In American football, the game can be broken down into a discrete sequence based on each down. 
+Contrast this to sports like lacrosse, soccer, or any variant of hockey, which feature largely running clocks and continuous player movement. 
+In these sports, it is not obvious how to break up the action into discrete chunks.
 
-For clarity, we start with baseball due to the discrete nature of the sport. 
+In this Section, we illustrate how the fundamental concept of the expected value of a game state leads to compelling findings in a variety of sports.
+
+## Discrete event analysis
+
+First, we explore results derived from the expected value of a state in sports where discrete event analysis is common. 
+We draw primarily on baseball and American football, but applications in other sports (e.g., tennis) are common (see for example, @kovalchik2019calibration).
+
+### In baseball, the expected run matrix
+
 In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
@@
 The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics.
-Figure \ref{fig:lindsey} shows a reproduction of George @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
 
@@
 For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.
+Figure \ref{fig:lindsey} shows a reproduction of @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its most common form.
+
 @Tango2007book (and many subsequent analyses) conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
@@
 
-## In American football {#sec:football}
+### In American football, expected points {#sec:football}
 
@@
 Other attempts at modeling expected points in football are @goldner2012markov and @goldner2017situational, who propose a Markov framework.
-In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
+In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* that include touchdowns, field goals, and other play outcomes. 
 For any given play, the expected points are calculated using the absorption probabilities for different scoring events.
@@
 
-### 4th down strategy
+### In American football, 4th down strategy
 
@@
 Numerous other papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy.
-In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down.
+In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a causal inference framework to reach similar conclusions that NFL teams are too conservative in going for it on 4th down.
 In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.
@@
 
-## In sports with continuous states
+## Continuous event analysis
 
@@
 This is a great leap forward from older resources such as traditional box-score results and play-by-play data.
+
+### In basketball, expected point value
 
@@
 See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
-The strategy of maximizing expected points in basketball has lead directly to the proliferation of three-point shooting in the NBA.
+The strategy of maximizing expected points in basketball has led directly to the proliferation of three-point shooting in the NBA.
+
+### In American football, yards gained
 
@@
 
+### In other sports
+
 The notion of expected possession value has also been extended to association football (soccer).
@@
 
-@kumagai2021hockey presents an EPV metric for ice hockey.
+@kumagai2021hockey offer an EPV metric for ice hockey.
 
@@
 Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
-Table \ref{tab:tied} (derived from Table \ref{tab:erm}) reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13.
+Information derived from Table \ref{tab:erm} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13. 
 The defense is in a tight spot, facing an 87% probability of losing the game. 
@@
 
-```{r}
+```{r, eval=FALSE}
 erm <- erm |>
@@
 A popular method for estimating team strength in sports is through pairwise evaluations.
+
+## Bradley-Terry models
+
 Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
@@
 
+## Elo ratings
+
 Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
-See @koning2017rating for more information on applications of the Elo rating in soccer. 
+See @koning2017rating for more information on applications of the Elo rating in soccer, and @kovalchik2019calibration and @kovalchik2016searching for Elo ratings in tennis. 
 Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at [FiveThirtyEight.com](https://fivethirtyeight.com).
+
+## Bayesian state-space models
 
@@
 See also @glickman2017estimating for more discussion on estimating team strengths in American football.
+
 More recently, @Lopez2018 extend @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
@@
 Apart from sports gambling, having an accurate estimate of team strength is useful to team officials, who are constantly monitoring and forecasting their team's ability.
+
+@koopman2015dynamic uses a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.
 

### 2022-10-16 sports_analytics.Rmd
ctv table (#37)

@@
 Software tools for sports analytics are similarly numerous. 
-For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network and organizes them by sport [@baumer2022ctv].
-A more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].
+For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network (CRAN) and organizes them by sport [@baumer2022ctv].
+Table \ref{tab:ctv} provides an overview of the currently available sport-specific CRAN packages.
+Further, a more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].
+
+```{r}
+tibble(
+  "American Football" = "nflverse, nflfastR, nflreadr, nfl4th, nflseedR,\n nflplotR, NFLSimulatoR, fflr, ffscrapr, ffsimulator, gsisdecoder, cfbfastR",
+  "Association Football (Soccer)" = "worldfootballR, engsoccerdata, socceR, ggsoccer, footballpenaltiesBL, footBayes, itscalledsoccer, FPLdata, EUfootball",
+  "Australian Rules Football" = "fitzRoy",
+  "Baseball/Softball" = "Lahman, retrosheet, pitchRx, mlbstats, baseballDBR, baseballr, runexp",
+  "Basketball" = "BAwiR, AdvancedBasketballStats, uncmbb, BasketballAnalyzeR, NBAloveR, wehoop, hoopR, toRvik",
+  "Chess" = "chess, stockfish, bigchess, rchess, chessR",
+  "Cricket" = "yorkr, cricketr, cricketdata, howzatR",
+  "Esports" = "CSGo, rbedrock, ROpenData, opendotaR, RDota2",
+  "GPS Activity Tracking" = "trackeR, trackeRapp, rStrava",
+  "Hockey" = "hockeyR, NHLData, nhlapi, nhlscrape, fastRhockey",
+  "Swimming" = "SwimmeR",
+  "Track and Field" = "combinedevents, JumpeR",
+  "Volleyball" = "volleystat"
+) |> 
+  pivot_longer(everything(),
+               names_to = "Sport",
+               values_to = "Packages") |>
+  transmute(Sport,
+            "Number of Packages" = str_count(Packages, ",") + 1,
+            "List of Packages" = Packages) |>
+  kable(booktabs = TRUE,
+        linesep = "",
+        caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN). \\label{tab:ctv}") |> 
+  column_spec(3, width = "7cm")
+```
 

### 2022-10-17 sports_analytics.Rmd
a couple of paragraphs on Elo (#36)

@@
 
-Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
+<!-- a basic Elo probability model for estimating the probability of $A$ beating $B$ is -->
+<!-- $$ -->
+<!-- P(A) = W(R_A - R_B) \,. -->
+<!-- $$ -->
+<!-- Here, $W$ is a continuous and monotonically increasing function that maps the difference in strength to (0, 1).  -->
+
+Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess.
+Given two players $i$ and $j$ with unknown ratings $R_i$ and $R_j$, the probability $\Pi_{ij}$ of $i$ beating $j$ is defined as
+$$
+\Pi_{ij} = \frac{1}{1 + K^{(R_j - R_i)/400}} \,.
+$$
+In this formula, $K$ is commonly known as the $K$-factor (or development coefficient).
+The International Chess Federation (FIDE) chooses $K=10$ for the calculation of Elo ratings.
+
+Another interpretation for $\Pi_{ij}$ is the expected score of the game for player $i$.
+The scores of 0, 0.5, and 1 are associated with three possible game outcomes loss, tie, and win, respectively.
+After a game, the updated Elo rating $R^*_i$ for player $i$ is
+$$
+R^*_i = R_i + K(S_i - \Pi_{ij}) \,,
+$$
+where $S_i \in \{0, 0.5, 1\}$.
+When a tournament concludes, a post-tournament rating is obtained for each player based on the rating updates for all games played.
+
+[Mention glicko as an alternative to improve Elo]
+
+Besides chess, the Elo system has also been implemented to estimate team strength in other sports. 
 See @koning2017rating for more information on applications of the Elo rating in soccer, and @kovalchik2019calibration and @kovalchik2016searching for Elo ratings in tennis. 
-Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at [FiveThirtyEight.com](https://fivethirtyeight.com).
+Furthermore, Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at [FiveThirtyEight.com](https://fivethirtyeight.com).
 

### 2022-10-17 sports_analytics.Rmd
Elo chess k-factor

@@
 $$
-In this formula, $K$ is commonly known as the $K$-factor (or development coefficient).
-The International Chess Federation (FIDE) chooses $K=10$ for the calculation of Elo ratings.
+In this formula, $K$ is commonly known as the $K$-factor, or development coefficient.
+The International Chess Federation (FIDE) uses $K=10$ for players with any previously achieved rating of at least 2400.
+For those that have never reached a rating of 2400 or better, $K=20$ is assigned. 
+Finally, $K=40$ is given to new players with under 30 games played, and players under the age of 18 with rating fewer than 2300 [@fide2022rating].
 
@@
 When a tournament concludes, a post-tournament rating is obtained for each player based on the rating updates for all games played.
-
-[Mention glicko as an alternative to improve Elo]
 

### 2022-10-17 response_to_reviewers.Rmd
response
address #35

@@
 
-> My largest concern is that a cursory discussion of player contribution or player ratings is given and this seems a major oversight of one of the “big ideas” in sports analytics given the focus on this topic in many sports.
-See, for example, Mark Glickman’s recent presentation on this at the 2022 Joint Statistical Meetings.
-I don’t think that Hvattum (2019) is sufficient to cover the area of player contribution.
+> My largest concern is that a cursory discussion of player contribution or player ratings is given and this seems a major oversight of one of the "big ideas" in sports analytics given the focus on this topic in many sports.
+See, for example, Mark Glickman's recent presentation on this at the 2022 Joint Statistical Meetings.
+I don't think that Hvattum (2019) is sufficient to cover the area of player contribution.
 
@@
 
-> In the second sentence, baseball is introduced as a “naturally discrete” sport, but there are certainly many things about baseball which are not discrete, as talked about later in the article.
+> In the second sentence, baseball is introduced as a "naturally discrete" sport, but there are certainly many things about baseball which are not discrete, as talked about later in the article.
 It would be helpful to clarify what is meant by discrete in this setting.
@@
 
-> In Section 2.1, Figure 1 (George Lindsey’s calculations) is introduced before the reader even knows who George Lindsey is.
+> In Section 2.1, Figure 1 (George Lindsey's calculations) is introduced before the reader even knows who George Lindsey is.
 
@@
 
-> Furthermore, the expected run matrix is displayed “in its more familiar form”, but I would venture for most readers of this article, no form would be familiar.
+> Furthermore, the expected run matrix is displayed "in its more familiar form", but I would venture for most readers of this article, no form would be familiar.
 
@@
 
-> I think the caption does explain the matrix nicely, although it might be useful to point out from which column in Lindsey’s table the matrix is derived.
+> I think the caption does explain the matrix nicely, although it might be useful to point out from which column in Lindsey's table the matrix is derived.
 
@@
 
-XX: Greg or Quang to address?
+We have added a sentence to address this. The expected value was unrealistic because the model is linear, which is a flaw of this approach.
 
@@
 
-XX
+We have added a table summarizing the sport-specific packages that are available on CRAN.
 
-> Figure 6 doesn’t add much to the text; I don’t think the reader better understands sports analytics by visualizing many team colors.
+> Figure 6 doesn't add much to the text; I don't think the reader better understands sports analytics by visualizing many team colors.
 Instead, it would be better to show  how the theming can lead to a more impactful graphic for sports, i.e.
@@
 
-> For Section 6, there is also now a “Big Data Derby” in horse racing.
+> For Section 6, there is also now a "Big Data Derby" in horse racing.
 
@@
 
-> Page 4, line 56: Needs to be some sort of separation between the “absorbing states” and “touchdown”
+> Page 4, line 56: Needs to be some sort of separation between the "absorbing states" and "touchdown"
 
@@
 
-> Page 5, line 56: Typo for “causal”
+> Page 5, line 56: Typo for "causal"
 
@@
 
-> Page 5, line 57: Should be “too conservative”
+> Page 5, line 57: Should be "too conservative"
 
@@
 
-> Page 7, line 37: Should be “led” not “lead”
+> Page 7, line 37: Should be "led" not "lead"
 
@@
 
-> Page 6, line 8:  “the effect may be overstated” is somewhat ambiguous.  Is it the
+> Page 6, line 8:  "the effect may be overstated" is somewhat ambiguous.  Is it the
 
@@
 
-> Page 6, line 53: Is the “!” warranted here for conditioning on a 9-dimensional vector?
+> Page 6, line 53: Is the "!" warranted here for conditioning on a 9-dimensional vector?
 
@@
 
-> 10, 52: “In this paper”, the “this” seems ambiguous.
+> 10, 52: "In this paper", the "this" seems ambiguous.
 
@@
 
-> 12, 37: Theming *teamcolors* is a great *R* package, not sure it is part of the “big ideas” in sports analytics.  A single line in the first paragraph of this section would seem to be better.
+> 12, 37: Theming *teamcolors* is a great *R* package, not sure it is part of the "big ideas" in sports analytics.  A single line in the first paragraph of this section would seem to be better.
 
@@
 
-> 13, 43-53: don’t think there is a need for a bulleted list here.
+> 13, 43-53: don't think there is a need for a bulleted list here.
+
+We have removed the bulleted lists.

### 2022-10-17 sports_analytics.Rmd
address #35
#32 and #36: elo with greg chess example

@@
 As expected, the estimated expected points increases monotonically as the teams gets closer to the endzone.
+One limitation of this approach is the linearity assumption, which results in a high negative expected point value when the offensive team is 95 yards away from the opponent's goal line.
 
@@
 
-## Elo ratings
+## Elo ratings {#sec:elo}
 
@@
 When a tournament concludes, a post-tournament rating is obtained for each player based on the rating updates for all games played.
+
+```{r}
+library(chessR)
+us <- c("fatgreggy")
+greg <- get_raw_chessdotcom(
+  usernames = us,
+  year_month = 202206
+)
+# greg: black
+```
+
+To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR]. 
+Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas `r greg[6, "WhiteElo"]` was the opponent's rating.
+Since both ratings are below 2400, we apply a development coefficient of $K=20$ for this example.
+The probability of author ($a$) defeating their opponent ($b$) is
+<!-- don't know how to include `r` inside latex -->
+$$
+\Pi_{ab} = \frac{1}{1 + 20^{(1683 - 1732)/400}} = 0.591 \,.
+$$
+The author that played this game ended up with a victory, which means the associated score is $S_a = 1$.
+The post-game Elo rating for this author is
+$$
+R^*_a = 1732 + 20(1 - 0.591) = 1740 \,.
+$$
 
@@
 
-Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in a game recently won by one of the authors (downloaded via the \pkg{chessR} package).
+Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package).
 

### 2022-10-17 sports_analytics.Rmd
added `nflplotR` figure (#31)

@@
 library(tidyverse)
+theme_set(theme_bw())
 library(kableExtra)
@@
     "chessR",
-    "sportyR"
+    "sportyR",
+    "nflplotR"
   ),
@@
   ) +
-  theme_bw() +
   theme(legend.position = "top")
@@
 Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas `r greg[6, "WhiteElo"]` was the opponent's rating.
-Since both ratings are below 2400, we apply a development coefficient of $K=20$ for this example.
+Since both ratings are below 2400, we apply a development coefficient of $K=20$ to this example.
 The probability of author ($a$) defeating their opponent ($b$) is
@@
 $$
-The author that played this game ended up with a victory, which means the associated score is $S_a = 1$.
+The playing author finished with a victory, which is associated with a score of $S_a = 1$.
 The post-game Elo rating for this author is
@@
 
+Sharing a similar goal with \pkg{teamcolors} is \pkg{nflplotR} [@R-nflplotR], which also provides \pkg{ggplot2} extensions but is designed specifically for the NFL.
+A great feature of \pkg{nflplotR} is the \pkg{ggplot2} "geom" (geometric object) functions that enhance high-quality plotting of NFL team logos and player images.
+Figure \ref{fig:nflplotr} shows a scatterplot of offensive and defensive expected points added for NFL teams in the 2021 regular season.
+The logos of all 32 American football clubs are plotted in place of the usual dots, giving a better description of which team each data point represents.
+
+```{r nflplotr, fig.cap="Offensive and defensive per play for the 2021 NFL regular season, plotted with \\pkg{nflplotR} using data from \\pkg{nflfastR}."}
+pbp <- nflreadr::load_pbp(2021) %>% 
+  dplyr::filter(season_type == "REG") %>%
+  dplyr::filter(!is.na(posteam) & (rush == 1 | pass == 1))
+
+offense <- pbp %>%
+  dplyr::group_by(team = posteam) %>%
+  dplyr::summarise(off_epa = mean(epa, na.rm = TRUE))
+
+defense <- pbp %>%
+  dplyr::group_by(team = defteam) %>%
+  dplyr::summarise(def_epa = mean(epa, na.rm = TRUE))
+
+combined <- offense %>%
+  dplyr::inner_join(defense, by = "team")
+
+combined |> 
+  ggplot(aes(off_epa, def_epa)) +
+  nflplotR::geom_mean_lines(aes(v_var = off_epa , h_var = def_epa)) +
+  nflplotR::geom_nfl_logos(aes(team_abbr = team), width = 0.08, alpha = 0.8) +
+  labs(x = "Offensive EPA/play",
+       y = "Defensive EPA/play") +
+  scale_y_reverse()
+```
+
 Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, or pitch for a particular sport.

### 2022-10-17 sports_analytics.Rmd
nfl scatterplot caption

@@
   - datasets
+link-citations: yes
 bibliography: [refs.bib, pkgs.bib]
@@
 
+<!-- 
 ```{r teamcolors, warning=FALSE, fig.cap="Professional sports team color palettes provided by the \\pkg{teamcolors} package. ", fig.height=9}
@@
 ``` 
+-->
 
@@
 
-```{r nflplotr, fig.cap="Offensive and defensive per play for the 2021 NFL regular season, plotted with \\pkg{nflplotR} using data from \\pkg{nflfastR}."}
+```{r nflplotr, fig.cap="Offensive and defensive expected points added per play for the 2021 NFL regular season, plotted with \\pkg{nflplotR} using data from \\pkg{nflfastR}."}
 pbp <- nflreadr::load_pbp(2021) %>% 

### 2022-10-24 response_to_reviewers.Rmd
#32 and #36: elo with greg chess example
close #38

@@
 
-We thank the reviewer for this suggestion. We have added a subsection on Elo rating.
-
-XX
+We thank the reviewer for this suggestion. We have added a subsection on Elo rating, with a mathematical defination followed by a calculation example.
 
@@
 We have added examples from the `SportyR` package, which draws playing surfaces, to help flesh out this subsection.
+We also replaced the previous Figure 6 with a new Figure 5 that illustrates how the use of team colors is particularly impactful for sports-releated data graphics.
 

### 2022-10-24 sports_analytics.Rmd
close #38

@@
 For example, the NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
-Figure \ref{fig:teamcolors} illustrates the full array of team color palettes for professional sports leagues.
+Figure \ref{fig:mlb_standings} illustrates how the use of team colors, which have a natural association for many sports fans, can help to untangle what would otherwise be messy data graphics. 
+In Figure \ref{fig:mlb_standings}, 30 different lines are plotted on top of one another, crisscrossing and intersecting in various unpredictable ways. 
+However, the use of team colors to identify the lines makes it possible to follow the trajectory of most teams over the course of the season.
 
-<!-- 
-```{r teamcolors, warning=FALSE, fig.cap="Professional sports team color palettes provided by the \\pkg{teamcolors} package. ", fig.height=9}
-teamcolors::show_team_col() 
+```{r}
+# remotes::install_github("beanumber/retro")
+library(retrosheet)
+retro_gl <- get_retrosheet("game", 2021)
+games_2021 <- retro_gl |>
+  mutate(home_win = HmRuns > VisRuns) %>%
+  select(Date, VisTm, HmTm, home_win) %>%
+  pivot_longer(
+    -c(Date, home_win),
+    names_to = "is_home", values_to = "team_id", 
+  ) %>%
+  mutate(
+    is_win = ifelse(
+      is_home == "HmTm", home_win, abs(1 - home_win))
+  ) %>%
+  group_by(team_id) %>%
+  mutate(
+    cum_W = cumsum(is_win), 
+    cum_L = cumsum(!is_win), 
+    cum_G = cum_W + cum_L, 
+    G_over_500 = cum_W - cum_L, 
+    last_game = cum_G == max(cum_G)
+  )
+retro_teams <- retrosheet::getTeamIDs(2021) |>
+  enframe(name = "mascot", value = "retro_id")
+mlb_pal <- teamcolors |>
+  filter(league == "mlb", str_detect(division, "NL")) |>
+  inner_join(retro_teams, by = "mascot") |>
+  select(retro_id, primary) |>
+  deframe()
+games_2021 <- games_2021 |>
+  filter(team_id %in% names(mlb_pal))
 ```
--->
 
-Sharing a similar goal with \pkg{teamcolors} is \pkg{nflplotR} [@R-nflplotR], which also provides \pkg{ggplot2} extensions but is designed specifically for the NFL.
-A great feature of \pkg{nflplotR} is the \pkg{ggplot2} "geom" (geometric object) functions that enhance high-quality plotting of NFL team logos and player images.
+```{r mlb_standings, fig.cap="The progression of National Leauge team standings during the 2021 Major League Baseball season.
+Note how the use of team colors makes it possible to untangle what would otherwise be a messy jumble of indistinguishable lines.
+Data provided by \\pkg{retrosheet} and colors provided by \\pkg{teamcolors}."}
+ggplot(
+    games_2021, 
+    aes(x = cum_G, y = G_over_500, color = team_id)
+  ) + 
+  geom_hline(yintercept = 18, color = "darkgray", lty = 2) + 
+  geom_line() +  
+  ggrepel::geom_label_repel(
+    data = filter(games_2021, last_game), 
+    aes(label = team_id)
+  ) + 
+  scale_color_manual(
+    guide = "none", name = "Team",
+    values = mlb_pal
+  ) +
+  scale_y_continuous("Games above .500") + 
+  scale_x_continuous("Cumulative Games Played") + 
+  labs(title = "Daily MLB Standings, 2021", 
+       caption = "Source: Retrosheet")
+```
+
+\pkg{nflplotR} [@R-nflplotR] has a similar goal to \pkg{teamcolors}.
+It also provides \pkg{ggplot2} extensions but is designed specifically for the NFL.
+A great feature of \pkg{nflplotR} is the collection of `geom_*()` (geometric object) functions that enhance high-quality plotting of NFL team logos and player images with \pkg{ggplot2}.
 Figure \ref{fig:nflplotr} shows a scatterplot of offensive and defensive expected points added for NFL teams in the 2021 regular season.
-The logos of all 32 American football clubs are plotted in place of the usual dots, giving a better description of which team each data point represents.
+The logos of all 32 American football clubs are plotted in place of the usual dots, making it easier for the reader to identify which team each data point represents.
 

### 2022-10-24 sports_analytics.Rmd
added new subsection on baseball tools; #31; #32

@@
     "sportyR",
-    "nflplotR"
+    "nflplotR",
+    "baseballr"
   ),
@@
 ) |> 
-  pivot_longer(everything(),
+  pivot_longer(
+    everything(),
     names_to = "Sport",
-               values_to = "Packages") |>
-  transmute(Sport,
+    values_to = "Packages"
+  ) |>
+  transmute(
+    Sport,
     "Number of Packages" = str_count(Packages, ",") + 1,
-            "List of Packages" = Packages) |>
-  kable(booktabs = TRUE,
+    "List of Packages" = Packages
+  ) |>
+  arrange(desc(`Number of Packages`)) |>
+  kable(
+    booktabs = TRUE,
     linesep = "",
-        caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN). \\label{tab:ctv}") |> 
+    caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN).
+While the major North American sports dominate the list, perhaps the fastest-growing collection is for esports.
+\\label{tab:ctv}"
+  ) |> 
   column_spec(3, width = "7cm")
@@
 
-## Graphical tools
+## Graphical tools {#sec:graphics}
 
@@
 
+## Case study in the evolution of tools and research: baseball
+
+As the granularity of baseball data has evolved over time, so too have the statistical methodologies for modeling that data, and the tools for working with it. 
+For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
+Over time, the resolution of baseball data has improved to include play-by-play data, pitch-by-pitch data, and now player tracking data. 
+The \pkg{Lahman} package [@R-Lahman] has provided seasonal data for years. 
+The \pkg{retrosheet} package [@R-retrosheet] now provides access to the historical play-by-play data available from Retrosheet (this is a comprehensive version of what Lindsay collected for his research).
+The \pkg{pitchRx} package [@R-pitchRx] provides access to pitch-by-pitch data that fueled innovative research into catcher framing [@deshpande2017hierarchical], pitch values [@healey2019bayesian], and pitch classification [@sidle2018using].
+The \pkg{baseballr} package [@R-baseballr] now provides access to player tracking data from Statcast.
+These data have led to investigations into how defensive shifts affect batting performance [@bouzarth2021swing], as well as how launch angles affect the probability of hitting a home run [@marchi2018analyzing].
+
+As we saw above with chess, the packages in baseball fit together in creative ways. 
+In Figure \ref{fig:mlb_standings}, we showed how \pkg{teamcolors} can illuminate data pulled from \pkg{retrosheet} to make an informative data graphic. 
+One could just as easily use \pkg{sportyR} to generate a field graphic, and then overlay player tracking data obtained from \pkg{baseballr} to depict defensive shifts.
+
+This, these R packages enable research by making data more easily available. 
+Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
+[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts.]
+
 # Opportunities {#sec:opportunities}

### 2022-10-26 sports_analytics.Rmd
test

@@
 
-@lopez2020bigger presents an introduction to NFL tracking data, and as an example, examines 4th down behavior.
+@lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
+In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is really only a rounded approximation of the true distance "to go" (i.e.
+1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1).
+Whereas a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard.
+With tracking data, the "to go" distance can be much more accurately assessed.
+
 Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.

### 2022-10-27 sports_analytics.Rmd
Updated the mike lopez paper section.

@@
 In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is really only a rounded approximation of the true distance "to go" (i.e.
-1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1).
-Whereas a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard.
-With tracking data, the "to go" distance can be much more accurately assessed.
-
-Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.
-They demonstrate that teams are still conservative when choosing to go for it or punting on 4th down, but the effect may be overstated.
+1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1.
+In fact, anything up to 2 yard will recorded as 4th and 1 @lopezDM).
+Whereas a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making.
+With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data.
+Many past analyses of the decision to go for it or not on 4th down conclude that coached in the NFL are too conservative in their decisions making.
+@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but the effect may be overstated due to the way in which to go yardage has been recorded in the past.
 

### 2022-11-01 response_to_reviewers.Rmd
added new subsection on baseball tools; #31; #32
added pythag; close #33

@@
 
-XX We could consider this.
+We thank the reviewer for this excellent idea. We have added a new subsection on Expected Winning Percentages models to the beginning of Section 4.
 
@@
 
-XX Yeah
+We have expanded the case study in chess to include Elo ratings...
+
+In addition, we have added new subsection with a case study in baseball about the evolution of data, tools, and statistical methodologies.
+We hope that these additional will help to flesh out the value of these tools.
 

### 2022-11-01 sports_analytics.Rmd
added pythag; close #33

@@
 A third crucial idea in sports analytics is the estimation of team strength. 
-A popular method for estimating team strength in sports is through pairwise evaluations.
+First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
+Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations.
+
+The impetus for all methods for estimating team strength is the realization that won-loss records are a noisy measure of team strength. 
+As binary outcomes, and with all sports (except perhaps chess) involving some element of chance, wins and losses carry some signal of team strength, but we can do better.
+
+## Expected winning percentage
+
+The first method for estimating team strength to become popular in sports analytics in a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
+Later, @miller2007derivation derived the formula as a consequence of assuming that runs are generated by two independent Weibull processes.
+
+Expected winning percentage is just:
+$$
+  \widehat{wpct} = \frac{X^\alpha}{X^\alpha + Y^\alpha} \,,
+$$
+where $X$ is the number of points (runs) that a team has scored, and $Y$ is the number of points (runs) that they have allowed, over some specified time period. 
+James's work was originally in baseball, and he posited the value of $\alpha = 2$. 
+The resemblance to the formula for computing the length of the hypotenuse in a right triangle provides the nod to Pythagoras.
+
+Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
+This can be done with a few lines of code, after observing that 
+$$
+  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
+$$
+and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
+Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball.
+
+```{r}
+library(tidyverse)
+library(Lahman)
+exp_wpct <- function(x, k = 2) { 
+  return(1/(1 + (1/x)^k))
+}
+
+TeamRuns <- Teams %>% 
+  filter(yearID >= 1954) %>%
+  rename(RS = R) %>% 
+  mutate(WPct = W / (W + L), run_ratio = RS/RA) %>%
+  select(yearID, teamID, lgID, WPct, run_ratio)
+
+data_space <- ggplot(
+  data = TeamRuns, aes(x = run_ratio, y = WPct)
+) +
+  geom_vline(xintercept = 1, color = "darkgray", linetype = 2) +
+  geom_hline(yintercept = 0.5, color = "darkgray", linetype = 2) +
+  geom_point(alpha = 0.2) + 
+  xlab("Ratio of Runs Scored to Runs Allowed") + 
+  ylab("Winning Percentage")
+```
+
+```{r}
+optimal <- TeamRuns %>%
+  nls(
+    formula = WPct ~ 1/(1 + (1/run_ratio)^k), 
+    start = list(k = 2)
+  ) %>%
+  coef()
+```
+
+```{r pythag, fig.cap="Winning percentages vs.
+runs scored and runs allowed in baseball, 1954--2021.
+The navy line represents the expected winning percentage model posited by James, with the exponent value of 2.
+The gold line shows the same model with an optimal exponent of 1.84."}
+data_space +
+  stat_function(fun = exp_wpct, args = list(2), size = 2, color = "navy") +
+  stat_function(fun = exp_wpct, args = list(optimal), size = 2, color = "gold")
+```
+
+Many authors have fit expected winning percentage models to other sports---too many to cite here. 
+See, for example, @hamilton2011extension for association football (soccer), @caro2013testing for Division I college football, and notably, future NBA general manager Daryl Morey for basketball [@dewan1993stats].
 
@@
 
-Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
+Perhaps the most widely-used probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
 For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 

### 2022-11-01 response_to_reviewers.Rmd
close #36

@@
 
-We thank the reviewer for this suggestion. We have added a subsection on Elo rating, with a mathematical defination followed by a calculation example.
+We thank the reviewer for this excellent suggestion.
+We have added a subsection on Elo ratings, with a mathematical definition followed by a calculation example.
+This example uses the `chessR` package, which ties into and strengthens our discussion of tools in the next chapter.
 
@@
 
-We have added the sentence: "Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at FiveThirtyEight.com." But again, we acknowledge that Wiley policy may not allow such a reference.
+We have added the sentence: "Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at FiveThirtyEight.com."
 

### 2022-11-01 sports_analytics.Rmd
close #36

@@
 
-To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR]. 
-Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas `r greg[6, "WhiteElo"]` was the opponent's rating.
+To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR] (see Section \ref{sec:chess}).
+Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas his opponent was rated `r greg[6, "WhiteElo"]`.
 Since both ratings are below 2400, we apply a development coefficient of $K=20$ to this example.
-The probability of author ($a$) defeating their opponent ($b$) is
+The probability of the author ($a$) defeating their opponent ($b$) was
 <!-- don't know how to include `r` inside latex -->
@@
 $$
-The playing author finished with a victory, which is associated with a score of $S_a = 1$.
-The post-game Elo rating for this author is
+The author won the match: that outcome is associated with a score of $S_a = 1$.
+The post-game Elo rating for the author is thus
 $$
@@
 
-## Case study in how tools fit together: chess
+## Case study in how tools fit together: chess {#sec:chess}
 

### 2022-11-01 response_to_reviewers.Rmd
close #32

@@
 > As mentioned earlier, Section 5 seems sparse on details, and furthermore it seems odd to choose chess as the case study when it has barely been mentioned previously in the article.
-I think this section could use a table of some sort, perhaps something that summarizes the CRAN task view at a high level.
+
+Our new subsection on Elo ratings also discusses chess, so now, the reader has already seen a methodological application to chess by the time they get to the case study.
+We have also added a new subsection to this Section about baseball, in the hopes of fleshing it out.
+
+> I think this section could use a table of some sort, perhaps something that summarizes the CRAN task view at a high level.
 Another potential would be a figure which shows a typical workflow of sports analytics as explained in the text.
@@
 
-We have expanded the case study in chess to include Elo ratings...
-
-In addition, we have added new subsection with a case study in baseball about the evolution of data, tools, and statistical methodologies.
-We hope that these additional will help to flesh out the value of these tools.
+We have connected the case study in chess to the new subsection on Elo ratings and the subsequent subsection on graphical tools.
+We hope that these explicit connections to additional content will make this subsection feel more substantial.
+In addition, we have added new subsection to this Section with a different case study in baseball about the evolution of data, tools, and statistical methodologies.
+We hope that this additional content will help to flesh out the value of these tools.
 

### 2022-11-01 sports_analytics.Rmd
close #32

@@
 This type of infrastructure, where one package is the "workhorse" that facilitates common generic data operations, and other packages layer on specific functionality, is common in sports analytics.
+See Section \ref{sec:elo} for an example of how the \pkg{chessR} package can be used to compute Elo ratings.
 
 Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package).
+We note how the contextual information provided by the chessboard is instrumental in helping the reader contextualize the data. 
+(How many of us can visualize PGN directly?)
+In Section \ref{sec:graphics}, we outline a collection of graphical tools that provide similar context for different playing surfaces.
 
@@
 
-Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, or pitch for a particular sport.
+Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, board, or pitch for a particular sport.
 Orienting these points graphically may require drawing a complex set of guidelines that provide that context to readers. 

### 2022-11-01 sports_analytics.Rmd
clean up #34

@@
 @lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
-In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is really only a rounded approximation of the true distance "to go" (i.e.
+In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e.
 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1.
 In fact, anything up to 2 yard will recorded as 4th and 1 @lopezDM).
-Whereas a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making.
+However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making.
 With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data.

### 2022-11-01 response_to_reviewers.Rmd
clean up #34
close #39

@@
 
-XX
+We hope that this revision addresses the reviewer's concerns about the pacing of the second half of the paper.
+In particular, we have added new subsections to both Sections 4 and 5, and added language tying together other parts of these Sections to ideas explored earlier in the paper.
+
+Respectfully, we do not see player tracking data as a theme of the paper.
+However, one of the new subsections we added traces the evolution of data alongside the evolution in methodology and insights made in baseball.
+In the conclusion, we have pointed to this discussion as representative of similar evolutions happening in other sports.
+We hope that his will satisfy the reviewer's desire for clarity on this issue.
 
@@
 
-XX Quang or Greg to respond
+We have clarified our language on this subject. The reason for the comment was related to the increased resolution of player tracking data. We have made this reasoning explicit.
 

### 2022-11-01 sports_analytics.Rmd
close #39

@@
 While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible. 
+Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
+We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper.
 
@@
 
-## Expected winning percentage
+## Expected winning percentage {#sec:wpct}
 
@@
 
-## Case study in the evolution of tools and research: baseball
+## Case study in the evolution of tools and research: baseball {#sec:baseball}
 
 As the granularity of baseball data has evolved over time, so too have the statistical methodologies for modeling that data, and the tools for working with it.
+
 For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
+These data, now available through the \pkg{Lahman} package [@R-Lahman], were sufficient to study broad trends in baseball, and led to insights such as the value of expected winning percentage (see Section \ref{sec:wpct}) and the importance of on-base percentage relative to batting average.
+These relatively simple insights fueled the "Moneyball" [@lewis2004moneyball] era revolution in sports analytics [@baumer2014sabermetric].
+
 Over time, the resolution of baseball data has improved to include play-by-play data, pitch-by-pitch data, and now player tracking data.
-The \pkg{Lahman} package [@R-Lahman] has provided seasonal data for years. 
+
 The \pkg{retrosheet} package [@R-retrosheet] now provides access to the historical play-by-play data available from Retrosheet (this is a comprehensive version of what Lindsay collected for his research).
+This play-by-play data allowed researchers to learn about strategies, like those that we discussed in Section \ref{sec:ev}.
+In baseball, this deepened our understanding of bunting, stolen bases, handedness, batting order, and many other aspects of the game. 
+Play-by-play data underlies much of the analysis in @Tango2007book.
+
 The \pkg{pitchRx} package [@R-pitchRx] provides access to pitch-by-pitch data that fueled innovative research into catcher framing [@deshpande2017hierarchical], pitch values [@healey2019bayesian], and pitch classification [@sidle2018using].
+Catcher framing is a notable example of a concept that scouts talked about for decades, but that could not be quantified by analysts until data of the appropriate resolution became available.
+
+While play-by-play data allows us to make valuations *between plays*, player tracking data allows us to make valuations *within plays*. 
 The \pkg{baseballr} package [@R-baseballr] now provides access to player tracking data from Statcast.
@@
 
-This, these R packages enable research by making data more easily available. 
+Thus, these R packages enable research by making data more easily available. 
 Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
-[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts.]
+[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts (see Section \ref{sec:opportunities}).]
 
@@
 
-# Conclusion
+# Conclusion {#sec:conclusion}
 
@@
 In drawing together these three big ideas in sports analytics, we have also drawn attention to some computational tools for doing the work and opportunities to showcase that work. 
+Our discussion in Section \ref{sec:baseball} shows how the increased resolution of available data has catalyzed new research directions in baseball, but this same dynamic is playing out in all sports.
 It is through these exchanges of ideas, tools, models, and data that analytics moves our collective understanding of sports forward.

### 2022-11-07 response_to_reviewers.Rmd
close #40

@@
 I don't think that Hvattum (2019) is sufficient to cover the area of player contribution.
+
+We thank the reviewer for this comment. We have addressed this important concern in multiple ways:
+
+- We do not consider individual player ratings to be within the scope of this paper, because they are covered by Albert (2015) in this same journal, and because their inclusion would double the length of this paper.
+We have added the following clarification in scope to the Introduction:
+
+"Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by Albert (2015), and its inclusion would double the length of this manuscript.
+We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section 4."
+
+- We have added the following sentence to our Section on team strength: "The methods in this Section apply equally well to multiplayer teams and single-player teams."
+
+- We have added to the existing clarification in scope that we are not considering individual player ratings in team sports in the Conclusion by citing a recent paper from Glickman on this subject.
+"Recent work by Che & Glickman (2022) also addresses this question across sports."
+
+We hope that these changes will satisfy the reviewer's largest concern.
 

### 2022-11-07 sports_analytics.Rmd
close #40

@@
 We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper. 
+Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by @Albert2015, and its inclusion would double the length of this manuscript.
+We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section \ref{sec:strength}.
 
@@
 Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations. 
+The methods in this Section apply equally well to multiplayer teams and single-player teams.
 
@@
 
-The first method for estimating team strength to become popular in sports analytics in a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
-Later, @miller2007derivation derived the formula as a consequence of assuming that runs are generated by two independent Weibull processes.
+The first method for estimating team strength to become popular in sports analytics is a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
+Later, @miller2007derivation derived the formula as a consequence of assuming that runs (in baseball) are generated by two independent Weibull processes.
 
@@
 and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
-Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball.
+Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954.
 
@@
 $$
-\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768 \,.
+\hat\Pi_{24} = \frac{\exp(\hat\beta_2 - \hat\beta_4)}{1 + \exp(\hat\beta_2 - \hat\beta_4)} = \frac{\exp(1.136 + 0.062)}{1 + \exp(1.136 + 0.062)} = 0.768 \,.
 $$
@@
 The second question is often addressed using a formulation of *wins above replacement* (WAR)---see @baumer2015openwar and @Yurko2019nflwar for details in baseball and American football. 
+Recent work by @che2022athlete also addresses this question across sports. 
 The third question is most compelling in sports like basketball, ice hockey, and soccer, where substitutions are common and it is obvious that different combinations of players with different sets of skills will result in squad of varying strengths and weaknesses.

### 2022-11-07 response_to_reviewers.Rmd
updated response

@@
 
-Comments to the Author(s)
-Please see attached file for comments.
+> Overall, I thought the authors were successful in introducing the big ideas in sports
+analytics at a general and non-technical level, and many interesting aspects of the literature
+were discussed. However, I think the second part of the title ”Statistical Tools for Their
+Investigation” was not properly explored. Several packages are mentioned throughout, and
+a brief sketch of a possible analysis workflow is given in Section 5, but I think it lacks sufficient
+detail. Beyond that, I think there are some extraneous figures and possible improvements
+with the overall flow.
+
+Our revision includes a significant bolstering of Sections 4, 5, and 6.
+In particular, we have added a subsection on Elo ratings, added a subsection on baseball tools, expanded and linked the discussion of graphical tools, and replaced some weak figures with more relevant substitutes.
 
@@
 > While Sections 2 and 3 are well done and complete, Sections 4 and 5 are weak and seem incomplete and unpolished.
+
+As noted above, Sections 4 and 5 are significantly stronger in this revision.
 
@@
 
-We believe that the last sentence of the first paragraph does exactly that.
+We have added a citation to Healey (2017) to this effect.
 
@@
 
-XX
+We thank the reviewer for this excellent suggestion.
+We have replaced Figure 6 with a new figure that illustrates in the `retrosheet` and `teamcolors` package can be combined to produce a meaningful data graphic.
+In particular, we note how the use of team colors clarifies what would otherwise be a messy jumble of indistinguishable lines.
 
@@
 We have added examples from the `SportyR` package, which draws playing surfaces, to help flesh out this subsection.
-We also replaced the previous Figure 6 with a new Figure 5 that illustrates how the use of team colors is particularly impactful for sports-releated data graphics.
+We also replaced the previous Figure 6 with a new Figure 6 that illustrates how the use of team colors is particularly impactful for sports-releated data graphics.
 

### 2022-11-07 sports_analytics.Rmd
updated response

@@
 In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
-And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact.
+And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact [@healey2017new].
 
@@
   ) + 
+  annotate("text", x = 2, y = 22, label = "90 wins", color = "darkgray", hjust = "left") +
   scale_color_manual(

### 2022-11-07 response_to_reviewers.Rmd
conferences; close #41

@@
 
->Likewise the Opportunities section should be expanded to include the proliferation of Sports Analytics conferences.
+> Likewise the Opportunities section should be expanded to include the proliferation of Sports Analytics conferences.
+
+We thank the reviewer for this suggestion. We have added two paragraphs on sports analytics conferences to Section 6.
 

### 2022-11-07 sports_analytics.Rmd
conferences; close #41

@@
 
-Public research in sports analytics is driven in part by several notable competitions. 
-These competitions have been an important source of new ideas and have contributed to the diversification of the field by breaking down barriers to entry.
+Public research in sports analytics is driven in part by several notable competitions and conferences. 
+These venues have been an important source of new ideas and have contributed to the diversification of the field by breaking down barriers to entry.
 
@@
 Similar competitions that provide opportunities for aspiring researchers to tackle sports analytics problems include the [Big Data Cup](https://www.stathletes.com/big-data-cup/) for ice hockey and the [SABR Diamond Dollars Case Competition](https://sabr.org/analytics/case) for baseball, and the [Big Data Derby](https://www.kaggle.com/competitions/big-data-derby-2022) for horse racing.
+
+The biennial [New England Symposium on Statistics in Sports](https://www.nessis.org/) is likely the longest-running academic conference devoted to sports analytics. 
+It's West Coast counterpart is [The Cascadia Symposium on Statistics in Sports](http://www.cascadiasports.com/). 
+Many influential results have been showcased for the first time at these conferences. 
+Other prominent sports analytics conferences include the  [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2022/), [UConn Sports Analytics Symposium](https://statds.org/events/ucsas2022/), and [MathSport International](http://www.mathsportinternational.com/).
+
+The highest-profile sports analytics conference is undoubtedly the [Sloan Sports Analytics Conference](https://www.sloansportsconference.com/), which draws academics, industry professionals, vendors, and media organizations.
+While the conference holds a research competition and has certainly drawn attention to sports analytics, it has also been criticized for a variety of shortcomings. 
+These criticisms include a lack of emphasis on reproducibility in the research competition, high ticket prices, the large salaries taken by the organizers, and the lack of diversity among attendees and presenters [@funt2022].
 

### 2022-11-07 sports_analytics.Rmd
updated

--- 2022-11-07 sports_analytics.Rmd
@@
 While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible.
-Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
-We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper. 
-Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by @Albert2015, and its inclusion would double the length of this manuscript.
-We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section \ref{sec:strength}.
 
@@
 @lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
-In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e.
+In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is really only a rounded approximation of the true distance "to go" (i.e.
 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1.
 In fact, anything up to 2 yard will recorded as 4th and 1 @lopezDM).
-However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making.
+Whereas a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making.
 With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data.
@@
 A third crucial idea in sports analytics is the estimation of team strength. 
-First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
-Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations. 
-The methods in this Section apply equally well to multiplayer teams and single-player teams.
-
-The impetus for all methods for estimating team strength is the realization that won-loss records are a noisy measure of team strength. 
-As binary outcomes, and with all sports (except perhaps chess) involving some element of chance, wins and losses carry some signal of team strength, but we can do better.
-
-## Expected winning percentage {#sec:wpct}
-
-The first method for estimating team strength to become popular in sports analytics is a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
-Later, @miller2007derivation derived the formula as a consequence of assuming that runs (in baseball) are generated by two independent Weibull processes.
-
-Expected winning percentage is just:
-$$
-  \widehat{wpct} = \frac{X^\alpha}{X^\alpha + Y^\alpha} \,,
-$$
-where $X$ is the number of points (runs) that a team has scored, and $Y$ is the number of points (runs) that they have allowed, over some specified time period. 
-James's work was originally in baseball, and he posited the value of $\alpha = 2$. 
-The resemblance to the formula for computing the length of the hypotenuse in a right triangle provides the nod to Pythagoras.
-
-Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
-This can be done with a few lines of code, after observing that 
-$$
-  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
-$$
-and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
-Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954.
-
-```{r}
-library(tidyverse)
-library(Lahman)
-exp_wpct <- function(x, k = 2) { 
-  return(1/(1 + (1/x)^k))
-}
-
-TeamRuns <- Teams %>% 
-  filter(yearID >= 1954) %>%
-  rename(RS = R) %>% 
-  mutate(WPct = W / (W + L), run_ratio = RS/RA) %>%
-  select(yearID, teamID, lgID, WPct, run_ratio)
-
-data_space <- ggplot(
-  data = TeamRuns, aes(x = run_ratio, y = WPct)
-) +
-  geom_vline(xintercept = 1, color = "darkgray", linetype = 2) +
-  geom_hline(yintercept = 0.5, color = "darkgray", linetype = 2) +
-  geom_point(alpha = 0.2) + 
-  xlab("Ratio of Runs Scored to Runs Allowed") + 
-  ylab("Winning Percentage")
-```
-
-```{r}
-optimal <- TeamRuns %>%
-  nls(
-    formula = WPct ~ 1/(1 + (1/run_ratio)^k), 
-    start = list(k = 2)
-  ) %>%
-  coef()
-```
-
-```{r pythag, fig.cap="Winning percentages vs.
-runs scored and runs allowed in baseball, 1954--2021.
-The navy line represents the expected winning percentage model posited by James, with the exponent value of 2.
-The gold line shows the same model with an optimal exponent of 1.84."}
-data_space +
-  stat_function(fun = exp_wpct, args = list(2), size = 2, color = "navy") +
-  stat_function(fun = exp_wpct, args = list(optimal), size = 2, color = "gold")
-```
-
-Many authors have fit expected winning percentage models to other sports---too many to cite here. 
-See, for example, @hamilton2011extension for association football (soccer), @caro2013testing for Division I college football, and notably, future NBA general manager Daryl Morey for basketball [@dewan1993stats].
+A popular method for estimating team strength in sports is through pairwise evaluations.
 
@@
 
-Perhaps the most widely-used probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
+Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
 For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
@@
 $$
-\hat\Pi_{24} = \frac{\exp(\hat\beta_2 - \hat\beta_4)}{1 + \exp(\hat\beta_2 - \hat\beta_4)} = \frac{\exp(1.136 + 0.062)}{1 + \exp(1.136 + 0.062)} = 0.768 \,.
+\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768 \,.
 $$
@@
 
-To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR] (see Section \ref{sec:chess}).
-Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas his opponent was rated `r greg[6, "WhiteElo"]`.
+To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR]. 
+Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas `r greg[6, "WhiteElo"]` was the opponent's rating.
 Since both ratings are below 2400, we apply a development coefficient of $K=20$ to this example.
-The probability of the author ($a$) defeating their opponent ($b$) was
+The probability of author ($a$) defeating their opponent ($b$) is
 <!-- don't know how to include `r` inside latex -->
@@
 $$
-The author won the match: that outcome is associated with a score of $S_a = 1$.
-The post-game Elo rating for the author is thus
+The playing author finished with a victory, which is associated with a score of $S_a = 1$.
+The post-game Elo rating for this author is
 $$
@@
 @koopman2015dynamic uses a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.
+
+# Sports Betting
+@LopezBaumerMatthews2018 uses betting market data to [describe paper]
+
+@Boulier2017
+
+Another popular type of betting pool is a survivor pool.  
+@Bergman2017SurvivingAN presents formal optimization approaches for NFL survivor pools and concludes that planning for only part of the season yields optimal results in terms of maximizing survival probability.
 
@@
 
-## Case study in how tools fit together: chess {#sec:chess}
+## Case study in how tools fit together: chess
 
@@
 This type of infrastructure, where one package is the "workhorse" that facilitates common generic data operations, and other packages layer on specific functionality, is common in sports analytics.
-See Section \ref{sec:elo} for an example of how the \pkg{chessR} package can be used to compute Elo ratings.
 
 Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package).
-We note how the contextual information provided by the chessboard is instrumental in helping the reader contextualize the data. 
-(How many of us can visualize PGN directly?)
-In Section \ref{sec:graphics}, we outline a collection of graphical tools that provide similar context for different playing surfaces.
 
@@
 
-Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, board, or pitch for a particular sport.
+Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, or pitch for a particular sport.
 Orienting these points graphically may require drawing a complex set of guidelines that provide that context to readers. 
@@
 
-## Case study in the evolution of tools and research: baseball {#sec:baseball}
+## Case study in the evolution of tools and research: baseball
 
 As the granularity of baseball data has evolved over time, so too have the statistical methodologies for modeling that data, and the tools for working with it. 
-
 For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
-These data, now available through the \pkg{Lahman} package [@R-Lahman], were sufficient to study broad trends in baseball, and led to insights such as the value of expected winning percentage (see Section \ref{sec:wpct}) and the importance of on-base percentage relative to batting average.
-These relatively simple insights fueled the "Moneyball" [@lewis2004moneyball] era revolution in sports analytics [@baumer2014sabermetric].
-
 Over time, the resolution of baseball data has improved to include play-by-play data, pitch-by-pitch data, and now player tracking data. 
-
+The \pkg{Lahman} package [@R-Lahman] has provided seasonal data for years. 
 The \pkg{retrosheet} package [@R-retrosheet] now provides access to the historical play-by-play data available from Retrosheet (this is a comprehensive version of what Lindsay collected for his research).
-This play-by-play data allowed researchers to learn about strategies, like those that we discussed in Section \ref{sec:ev}.
-In baseball, this deepened our understanding of bunting, stolen bases, handedness, batting order, and many other aspects of the game. 
-Play-by-play data underlies much of the analysis in @Tango2007book.
-
 The \pkg{pitchRx} package [@R-pitchRx] provides access to pitch-by-pitch data that fueled innovative research into catcher framing [@deshpande2017hierarchical], pitch values [@healey2019bayesian], and pitch classification [@sidle2018using].
-Catcher framing is a notable example of a concept that scouts talked about for decades, but that could not be quantified by analysts until data of the appropriate resolution became available.
-
-While play-by-play data allows us to make valuations *between plays*, player tracking data allows us to make valuations *within plays*. 
 The \pkg{baseballr} package [@R-baseballr] now provides access to player tracking data from Statcast.
@@
 
-Thus, these R packages enable research by making data more easily available. 
+This, these R packages enable research by making data more easily available. 
 Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
-[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts (see Section \ref{sec:opportunities}).]
+[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts.]
 
@@
 
-# Conclusion {#sec:conclusion}
+# Conclusion
 
@@
 The second question is often addressed using a formulation of *wins above replacement* (WAR)---see @baumer2015openwar and @Yurko2019nflwar for details in baseball and American football. 
-Recent work by @che2022athlete also addresses this question across sports. 
 The third question is most compelling in sports like basketball, ice hockey, and soccer, where substitutions are common and it is obvious that different combinations of players with different sets of skills will result in squad of varying strengths and weaknesses.
@@
 In drawing together these three big ideas in sports analytics, we have also drawn attention to some computational tools for doing the work and opportunities to showcase that work. 
-Our discussion in Section \ref{sec:baseball} shows how the increased resolution of available data has catalyzed new research directions in baseball, but this same dynamic is playing out in all sports.
 It is through these exchanges of ideas, tools, models, and data that analytics moves our collective understanding of sports forward.

### 2022-11-07 sports_analytics.Rmd

--- 2022-11-07 sports_analytics.Rmd
@@
 
+# Sports Betting
+@LopezBaumerMatthews2018 uses betting market data to [describe paper]
+
+@Boulier2017
+
+Another popular type of betting pool is a survivor pool.  
+@Bergman2017SurvivingAN presents formal optimization approaches for NFL survivor pools and concludes that planning for only part of the season yields optimal results in terms of maximizing survival probability.
+
 # Tools {#sec:tools}

### 2022-11-07 sports_analytics.Rmd
Added a bunch of papers on gambling.

@@
 
-@Boulier2017
+@Boulier2006: Testing the efficiency of the National Football League betting market: Look at 1994 - 2000.
+Conclude there is no evidence to conclude the market was inefficient over this period of time.
+
+Boulier, B.
+L.
+and Stekler, H.
+O.
+(2003).
+Predicting the outcomes of National Football League games.
+International Journal of Forecasting 19 257{270.
+: They test several predictors for NFL games and find that the betting market is the best predictor.
+
+Buttrey, S. E. (2016). Beating the market betting on NHL hockey games. Journal of Quantitative
+Analysis in Sports 12 87{98. : They demonstrate a method that produced a positive ROI
+
+Gandar, J., Zuber, R., O'brien, T. and Russo, B. (1988). Testing rationality in the point spread
+betting market. The Journal of Finance 43 995{1008. : They conclude that Stats test are too weak to reject the hypothesis of rationality.  But they do reject rationalist with ``economic tests".
+
+Lacey, N.
+J.
+(1990).
+An estimation of market efficiency in the NFL point spread betting market.
+Applied Economics 22 117{129.
+: Finds the market to be efficient, though they did find some profitable betting rules over the period of time between 1984 and 1986
 
@@
 @Bergman2017SurvivingAN presents formal optimization approaches for NFL survivor pools and concludes that planning for only part of the season yields optimal results in terms of maximizing survival probability.
+
+#Betting pools
+Breiter DJ, Carlin BP (1997) How to play office pools if you must.Chance10(1):5–11
+
+Clair B, Letscher D (2007) Optimal strategies for sports betting pools.Oper. Res.55(6):1163–1177.
+
+Kaplan EH, Garstka SJ (2001) March Madness and the office pool.Management Sci.47(3):369–382
+
+Metrick A (1996) March Madness? Strategic behavior in {NCAA}basketball tournament betting pools.J. Econom. Behav. Organ.30(2):159–172.
+
+Niemi JB, Carlin BP, Alexander JM (2008) Contrarian strategies forNCAA tournament pools: A cure for March Madness?Chance21(1):35–42
+
+Sauer RD (1998) The economics of wagering markets.J. Econom. Lit-erature36(4):2021–2064
+
+Computing the Number of Winning NFL Survivor Pool Entries
+J Imbrogno, D Bergman - The College Mathematics Journal, 2022:
+Description: The average number of winning survivor pool entries is computed for the 2019 National Football League (NFL) season.
+The probability of having to share the winning pot, given realistic expectations on game outcomes, is also shown.
+
+Nichols, M.
+W.
+(2012).
+The impact of visiting team travel on game outcome and biases in NFL betting markets.
+Journal of Sports Economics 15 78{96.
+: They conclude that the impact of travel os not completely incorporated into the betting markets, but that any effect is too small to find anye profitable advantage.
+
+Paul, R.
+J.
+and Weinbach, A.
+P.
+(2014).
+Market efficiency and behavioral biases in the wnba betting market.
+International Journal of Financial Studies 2 193{202.: Over the period of 2007–2012, they find that the market is still hard to beat.
+
+Soebbing, B. P. and Humphreys, B. R. (2013). Do gamblers think that teams tank? Evidence from the NBA. Contemporary Economic Policy 31 301{313.
+Spann, M. and Skiera, B. (2009). Sports forecasting: a: NBA bettors believe that tanking is occurring.
 

### 2022-11-07 sports_analytics.Rmd
Working on the gambling section.

@@
 
-# Sports Betting
-@LopezBaumerMatthews2018 uses betting market data to [describe paper]
+# Sports betting 
+While @lopez2018 uses betting market data to model team strengths, it does not directly address strategies for betting or inefficiencies in betting markets.  Work such as @lacey1990, @Boulier2006,
 
@@
 
-Lacey, N.
-J.
-(1990).
-An estimation of market efficiency in the NFL point spread betting market.
-Applied Economics 22 117{129.
-: Finds the market to be efficient, though they did find some profitable betting rules over the period of time between 1984 and 1986
+@Lacey1990 : Finds the market to be efficient, though they did find some profitable betting rules over the period of time between 1984 and 1986
 
@@
 
-Paul, R.
-J.
-and Weinbach, A.
-P.
-(2014).
-Market efficiency and behavioral biases in the wnba betting market.
-International Journal of Financial Studies 2 193{202.: Over the period of 2007–2012, they find that the market is still hard to beat.
+@PaulWeinbach2014 Market efficiency and behavioral biases in the wnba betting market
 

### 2022-11-15 response_to_reviewers.Rmd
close #34
close #42

@@
 Our revision includes a significant bolstering of Sections 4, 5, and 6.
-In particular, we have added a subsection on Elo ratings, added a subsection on baseball tools, expanded and linked the discussion of graphical tools, and replaced some weak figures with more relevant substitutes.
+In particular, we have added a subsection on Elo ratings, added a subsection on baseball tools, expanded and linked the discussion of graphical tools, replaced some weak figures with more relevant substitutes, and added supporting references.
 
@@
 
-We have clarified our language on this subject. The reason for the comment was related to the increased resolution of player tracking data. We have made this reasoning explicit.
+We have clarified our language on this subject.
+The reason for the comment was related to the increased resolution of player tracking data.
+We have made this reasoning explicit and added citations to support it.
 

### 2022-11-15 sports_analytics.Rmd
close #34
close #42
references for sports betting; #29

--- 2022-11-07 sports_analytics.Rmd
@@
 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1.
-In fact, anything up to 2 yard will recorded as 4th and 1 @lopezDM).
+In fact, anything up to 2 yard will recorded as 4th and 1 [@lopezDM].
 However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making.
@@
 
-# Sports Betting
-@LopezBaumerMatthews2018 uses betting market data to [describe paper]
-
-@Boulier2017
+# Sports betting
+
+While @Lopez2018 uses betting market data to model team strengths, it does not directly address strategies for betting or inefficiencies in betting markets. 
+Work such as @lacey1990estimation, @Boulier2006,
+
+@Boulier2006 that there is insufficient evidence to conclude that betting markets were inefficient from 1994--2000.
+@boulier2003predicting test several predictors for NFL games and find that the betting market is the best predictor.
+@Buttrey2016 demonstrate a method that produces a positive ROI.
+@gandar1988testing conclude that stats test are too weak to reject the hypothesis of rationality.  But they do reject rationalist with ``economic tests". 
+@lacey1990estimation : Finds the market to be efficient, though they did find some profitable betting rules over the period of time between 1984 and 1986
+
+## Betting pools
 
@@
 @Bergman2017SurvivingAN presents formal optimization approaches for NFL survivor pools and concludes that planning for only part of the season yields optimal results in terms of maximizing survival probability.
+
+@breiter1997play
+
+@clair2007optimal
+
+@kaplan2001march
+
+@metrick1996march
+
+@niemi2008contrarian
+
+@sauer1998economics
+
+@imbrogno2022computing
+Description: The average number of winning survivor pool entries is computed for the 2019 National Football League (NFL) season.
+The probability of having to share the winning pot, given realistic expectations on game outcomes, is also shown.
+
+@nichols2014impact conclude that the impact of travel is not completely incorporated into the betting markets, but that any effect is too small to find any profitable advantage.
+
+@PaulWeinbach2014 Market efficiency and behavioral biases in the wnba betting market
+
+@soebbing2013gamblers
+
+@spann2009sports NBA bettors believe that tanking is occurring.
 
@@
 Table \ref{tab:ctv} provides an overview of the currently available sport-specific CRAN packages.
+@casals2022systematic offers a systematic review of sport-related packages on CRAN.
 Further, a more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].

### 2022-11-15 sports_analytics.Rmd
framing for sports betting; #29

@@
 
-# Sports betting
+# Non-performance data
+
+Most of the research in sports analytics is fueled by the analysis of data recorded from the outcome of sports contests.
+However, a growing body of literature is informed by non-performance data about sports. 
+This research includes fields like sports economics and kinesiology that are subfields of mature disciplines (a treatment of either is well beyond the scope of the this paper).
+In addition to this body of research, we note that today, professional sports teams are collecting biometric data on their own players, and presumably using that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance.
+While this research may constitute "emerging methodologies," it will likely never be published, because each team will use it to their competitive advantage.
+
+Nevertheless, in this Section we discuss concepts stemming from one notable, relatively new source of (semi-)public non-performance sports data: data from betting markets.
+Since the 2018 United States Supreme Court decision in [*Murphy v.
+National Collegiate Athletic Association*](https://en.wikipedia.org/wiki/Murphy_v._National_Collegiate_Athletic_Association), sports gambling has exploded in the U.S.
+The increasing interest in sports gambling has led to increasing interest in sports gambling *data*, and that data has proven useful to researchers in two major ways.
+
+First, betting market data is probably the best source for estimating the true probability of a team winning a game. 
+The efficiency of betting market data in this respect has been demonstrated time and time again.
+The utility of these estimates have then informed research that has helped us learn about the sports themselves.
+In this sense, data generated by sports gambling has been an important source of data useful for sports analytics (see Section \ref{sec:betting-market-data}).
+
+Second, sports analytics researchers have studied various types of sports gambling outlets (including fantasy sports). 
+This research has estimated probabilities, evaluated common strategies, and offered optimized approaches for a variety of different games of chance (see Section \ref{sec:pools}).
+
+## Sports betting market data {#sec:betting-market-data}
 
@@
 
-## Betting pools
+## Analytics for sports betting pools {#sec:pools}
 

### 2022-11-25 sports_analytics.Rmd
some edits.

@@
 
-While @Lopez2018 uses betting market data to model team strengths, it does not directly address strategies for betting or inefficiencies in betting markets. 
-Work such as @lacey1990estimation, @Boulier2006,
+While @Lopez2018 uses betting market data to model team strengths, it does not directly address strategies for betting or inefficiencies in betting markets. Early work by @gandar1988testing examines 
+Related work such as @lacey1990estimation and @Boulier2006 both explore the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
+Neither paper finds strong evidence for inefficiencies in the markets.
+@boulier2003predicting compared the predictive performance of power rankings and media experts to the betting market for NFL games and found that the betting market was the best for predicting winners.
+More recently @Buttrey2016 explored the NHL betting market.
+They produced a model to predict win probabilities in given games, and games where the predicted probability differed from the market price bets were placed.
+They find that using their methods, that they were able to produce a positive return on investment.
 

### 2022-11-28 sports_analytics.Rmd
close #29

--- 2022-11-15 sports_analytics.Rmd
@@
 However, a growing body of literature is informed by non-performance data about sports. 
-This research includes fields like sports economics and kinesiology that are subfields of mature disciplines (a treatment of either is well beyond the scope of the this paper).
+This research appears in fields like sports economics and kinesiology that are subfields of mature disciplines (a treatment of either is well beyond the scope of the this paper).
 In addition to this body of research, we note that today, professional sports teams are collecting biometric data on their own players, and presumably using that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance.
@@
 
-Nevertheless, in this Section we discuss concepts stemming from one notable, relatively new source of (semi-)public non-performance sports data: data from betting markets.
+Nevertheless, in this Section we discuss concepts stemming from one notable, relatively new source of (semi-)public non-performance sports data: betting markets.
 Since the 2018 United States Supreme Court decision in [*Murphy v.
 National Collegiate Athletic Association*](https://en.wikipedia.org/wiki/Murphy_v._National_Collegiate_Athletic_Association), sports gambling has exploded in the U.S.
-The increasing interest in sports gambling has led to increasing interest in sports gambling *data*, and that data has proven useful to researchers in two major ways.
+The increasing interest in sports gambling has led to increasing interest in sports gambling *data*, and that data has proven useful to researchers in at least two major ways.
 
@@
 This research has estimated probabilities, evaluated common strategies, and offered optimized approaches for a variety of different games of chance (see Section \ref{sec:pools}). 
-
-## Sports betting market data {#sec:betting-market-data}
-
-While @Lopez2018 uses betting market data to model team strengths, it does not directly address strategies for betting or inefficiencies in betting markets. 
-Work such as @lacey1990estimation, @Boulier2006,
-
-@Boulier2006 that there is insufficient evidence to conclude that betting markets were inefficient from 1994--2000.
-@boulier2003predicting test several predictors for NFL games and find that the betting market is the best predictor.
-@Buttrey2016 demonstrate a method that produces a positive ROI.
-@gandar1988testing conclude that stats test are too weak to reject the hypothesis of rationality.  But they do reject rationalist with ``economic tests". 
-@lacey1990estimation : Finds the market to be efficient, though they did find some profitable betting rules over the period of time between 1984 and 1986
-
-## Analytics for sports betting pools {#sec:pools}
-
-Another popular type of betting pool is a survivor pool.
-@Bergman2017SurvivingAN presents formal optimization approaches for NFL survivor pools and concludes that planning for only part of the season yields optimal results in terms of maximizing survival probability.
-
-@breiter1997play
-
-@clair2007optimal
-
-@kaplan2001march
-
-@metrick1996march
-
-@niemi2008contrarian
-
-@sauer1998economics
-
-@imbrogno2022computing
-Description: The average number of winning survivor pool entries is computed for the 2019 National Football League (NFL) season.
-The probability of having to share the winning pot, given realistic expectations on game outcomes, is also shown.
-
-@nichols2014impact conclude that the impact of travel is not completely incorporated into the betting markets, but that any effect is too small to find any profitable advantage.
-
-@PaulWeinbach2014 Market efficiency and behavioral biases in the wnba betting market
-
-@soebbing2013gamblers
-
-@spann2009sports NBA bettors believe that tanking is occurring.
+Some researchers have then tried to demonstrate a positive return on some of these betting strategies, with very limited success.
+
+## The use of sports betting market data {#sec:betting-market-data}
+
+While @Lopez2018 uses betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
+Early work by @gandar1988testing examines the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
+Related work such as @lacey1990estimation and @Boulier2006 explores the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
+Neither paper finds strong evidence for inefficiencies in the markets. 
+@boulier2003predicting compare the predictive performance of power rankings and media experts to the betting market for NFL games and found that the betting market is the best for predicting winners. 
+@lopez2015building show that betting market data was most useful in predicting men's college basketball tournament outcomes.
+
+Sports betting market data has also been used to investigate competitive behavior within leagues. 
+@soebbing2013gamblers find evidence that sports bettors *think* tanking in the NBA is occurring, although the evidence for whether it actually is remains mixed.
+
+## Analytics for sports betting {#sec:pools}
+
+### Betting pools
+
+One popular type of betting pool is a survivor pool, in which participants stay in the competition as long as they continue to successfully pick winners.
+@Bergman2017SurvivingAN present formal optimization approaches for NFL survivor pools and conclude that planning for only part of the season yields optimal results in terms of maximizing survival probability.
+@imbrogno2022computing estimate the probability of having to share the winning pot in NFL survivor pools.
+
+Perhaps the most commonly-studied sports betting market surrounds the NCAA men's college basketball tournament. 
+@breiter1997play use Monte Carlo methods to study the standard "office pool."
+@kaplan2001march consider a variety of NCAA college basketball pools, and find that the simple strategy of picking the team with the better seed is generally, but not always, optimal. 
+@metrick1996march finds that bettors overback the heaviest favorites. 
+@niemi2008contrarian show an improved return on investment by picking an undervalued champion and then completing the rest of one's bracket by using published odds. 
+@clair2007optimal develop and test strategies for maximizing expected return in both survivor and tournament-style pools.
+
+### Beating the market
+
+Naturally, after studying the efficiency of sports betting markets, researchers try to find inefficiencies that can be exploited for financial gain. 
+Not surprisingly (given the efficiency of these markets), such gains are hard to come by.
+
+@sauer1998economics finds that while racetrack betting markets are generally efficient, information asymmetry plays a role in creating inefficient markets. 
+@nichols2014impact concludes that the impact of travel is not completely incorporated into the betting markets, but that any effect is too small to find any profitable advantage. 
+@PaulWeinbach2014 investigate the less-saturated betting market for the WNBA and fail to find strategies for positive return on investment. 
+@spann2009sports show no way to beat the market in the German premier soccer league, given the high fees associated with placing bets.
+
+More successfully, @Buttrey2016 explores the NHL betting market and produces a model to predict win probabilities in given games, then tests the model by placing market price bets in games where the predicted probability differs from the market.
+They find that their methods were able to produce a positive return on investment.
 

### 2022-11-28 response_to_reviewers.Rmd
finishing touches

@@
 
-Our revision includes a significant bolstering of Sections 4, 5, and 6.
-In particular, we have added a subsection on Elo ratings, added a subsection on baseball tools, expanded and linked the discussion of graphical tools, replaced some weak figures with more relevant substitutes, and added supporting references.
+Our revision includes a significant bolstering of Sections 4, 5, 6, and 7.
+In particular, we have added a section on the use of non-performance data, added a subsection on Elo ratings, added a subsection on baseball tools, expanded and linked the discussion of graphical tools, including information about sports analytics conferences, replaced some weak figures with more relevant substitutes, and added supporting references.
+We hope these additions will satisfy the reviewer.
 
@@
 We thank the reviewer for this excellent suggestion.
-We have replaced Figure 6 with a new figure that illustrates in the `retrosheet` and `teamcolors` package can be combined to produce a meaningful data graphic.
+We have replaced Figure 6 with a new figure that illustrates how the `retrosheet` and `teamcolors` package can be combined to produce a meaningful data graphic.
 In particular, we note how the use of team colors clarifies what would otherwise be a messy jumble of indistinguishable lines.
@@
 We have added examples from the `SportyR` package, which draws playing surfaces, to help flesh out this subsection.
-We also replaced the previous Figure 6 with a new Figure 6 that illustrates how the use of team colors is particularly impactful for sports-releated data graphics.
+We also replaced the previous Figure 6 with a new Figure 6 that illustrates how the use of team colors is particularly impactful for sports-related data graphics.
 

### 2022-11-28 sports_analytics.Rmd
finishing touches

@@
 Doing the work of sports analytics requires computing with data. 
-While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible. 
+While the sources of sports data are too numerous to list, we discuss the use of non-performance data (most notably, sports market betting data) in Section \ref{sec:betting}, and in Section \ref{sec:tools} we highlight a few computational tools (including a table of R packages) that make this kind of work possible.
 Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
@@
   ) + 
-  geom_line(size = 1.5, alpha = 0.8) +
+  geom_line(linewidth = 1.5, alpha = 0.8) +
   labs(
@@
 
-The first method for estimating team strength to become popular in sports analytics is a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
+A simple method for estimating team strength that has become popular in sports analytics is expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
 Later, @miller2007derivation derived the formula as a consequence of assuming that runs (in baseball) are generated by two independent Weibull processes.
@@
 and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
-Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954.
+Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954, where the optimal value of $\alpha$ is 1.84.
 
@@
 data_space +
-  stat_function(fun = exp_wpct, args = list(2), size = 2, color = "navy") +
-  stat_function(fun = exp_wpct, args = list(optimal), size = 2, color = "gold")
+  stat_function(fun = exp_wpct, args = list(2), linewidth = 2, color = "navy") +
+  stat_function(fun = exp_wpct, args = list(optimal), linewidth = 2, color = "gold")
 ```
@@
 $$
-Here, $\exp{\beta_i}$ is often viewed as a representation of team $i$'s ability.
+Here, $\exp{(\beta_i)}$ is often viewed as a representation of team $i$'s ability.
 
@@
 
-# Non-performance data
+# Non-performance data {#sec:betting}
 
@@
 
-## The use of sports betting market data {#sec:betting-market-data}
+## The use of betting market data for sports analytics {#sec:betting-market-data}
 
@@
 
-```{r chess, eval=whoami::username() == 'bsbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
@@
 
-In drawing together these three big ideas in sports analytics, we have also drawn attention to some computational tools for doing the work and opportunities to showcase that work. 
+In drawing together these three big ideas in sports analytics, we have also drawn attention to new uses of sports betting market data, some computational tools for doing sports analytics work, and opportunities to showcase that work.
 Our discussion in Section \ref{sec:baseball} shows how the increased resolution of available data has catalyzed new research directions in baseball, but this same dynamic is playing out in all sports.

### 2023-01-03 response_to_reviewers_R1.Rmd
start on #44

---
title: "Response to Reviewers"
date: "2023-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Major comments from Reviewer 1

## Major comments from Reviewer 2

Overall Summary:

This revised version is a significant improvement over the previous version and the reviewers have addressed my major concerns with the paper with one exception.

Suggested Changes:

The authors have now made clearer the limitation in scope of this manuscript related to player evaluation. Perhaps, a statement regarding this can be added to the abstract.

I think that parts of the first paragraph of Section 5 regarding biometric data is a bit out of place. This could be moved to Section 7 (Opportunities). Likewise, Section 5 could be retitled something like‘Betting Data and Sports Analytics” or “Betting Data Analysis”.

## Minor Changes from Reviewer 2

- Page 1, line 47 “And, of course, in baseball” add commas before and - after “of course”

> Done.

- Page 2, line 5, “games share” instead of “game shares”

> Fixed.

- Page 2 line 6, “anthropologically’ seems odd here. Maybe “historically” instead

> Fixed.

- Page 8, Macdonald (2012?) used expected possession value for ice hockey to evaluate players.

> We have a added a citation to this work, but note that whereas Macdonald is using expected *goals*, he did not have access to player tracking data necessary to compute an EPV metric.

- Section 6.1, line 53, Figure ??
- Section 6.1, line 56, contextual is used twice in that sentence, perhaps “helping the reader visualize” or “helping the reader understand”

> We have changed this to "understand".

- Table 3, the third column of this table is off by a line with the entries in the first two columns.

> This is an artifact of the attempt to track changes and does not appear in the unannotated manuscript.

- Section 6.2, line 52, Figure 6 should be Figure 7; likewise the figure number in the next sentence

> This is a downstream consequence of the previous error with Figure ??

- Figure 6, in my version the images are stacked rather than left/right
- Section 6.2, line 46 This sentence that starts “Consider, for example, “ seems out of place with the rest of the tone of this article
- Section 7, line 7 or 8, it might be worth mentioning the growth of small conferences like UCSAS that have sprung up, something like “As the field of sports analytics has grown, there has been a proliferation of sports specific and regional sports analytics conferences.”

> We have added a similar sentence to this effect.

### 2023-01-03 sports_analytics_R0.Rmd
start on #44

--- 2022-11-25 sports_analytics.Rmd
@@
   - datasets
-link-citations: yes
 bibliography: [refs.bib, pkgs.bib]
@@
 csl: apa.csl
-output: rticles::arxiv_article
+output: 
+  rticles::arxiv_article:
+    keep_md: yes
 header-includes:
@@
 library(tidyverse)
-theme_set(theme_bw())
 library(kableExtra)
@@
     "chess", 
-    "chessR",
-    "sportyR",
-    "nflplotR",
-    "baseballr"
+    "chessR"
   ),
@@
 In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
-And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact [@healey2017new].
+And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact.
 
@@
 
-The concept of a state is easier to grasp in a sport that can be modeled as *discrete* (in the sense of [discrete event simulation](https://en.wikipedia.org/wiki/Discrete-event_simulation)). 
-By discrete, we mean a sport that can be easily broken into short, distinct segments of action which are typically summarized categorically. 
-Each of these segments might represent a state $s$. 
-For example, each pitch in baseball is either a ball or a strike. 
-If the ball is put in play, then there may be a complex sequence of movements by the players, but ultimately (within a few seconds) that sequence will end and no more action will be permitted until the next pitch.
-At the beginning and end of each phase of action, we will know definitively which team is on offense and defense, which runners are on which bases, the score, how many outs there are, etc. 
-Tennis could similarly be viewed as a series of discrete actions defined by each point. 
-To say that a player is winning 6-2, 3-1, 40-15 and serving with one fault committed is to characterize the state of the match.
-In American football, the game can be broken down into a discrete sequence based on each down. 
-Contrast this to sports like lacrosse, soccer, or any variant of hockey, which feature largely running clocks and continuous player movement. 
-In these sports, it is not obvious how to break up the action into discrete chunks.
+## In baseball, the expected run matrix
 
-In this Section, we illustrate how the fundamental concept of the expected value of a game state leads to compelling findings in a variety of sports.
-
-## Discrete event analysis
-
-First, we explore results derived from the expected value of a state in sports where discrete event analysis is common. 
-We draw primarily on baseball and American football, but applications in other sports (e.g., tennis) are common (see for example, @kovalchik2019calibration).
-
-### In baseball, the expected run matrix
-
+For clarity, we start with baseball due to the discrete nature of the sport. 
 In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
@@
 The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
+Figure \ref{fig:lindsey} shows a reproduction of George @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.
 
@@
 For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.
-Figure \ref{fig:lindsey} shows a reproduction of @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its most common form.
-
 @Tango2007book (and many subsequent analyses) conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.
@@
 
-### In American football, expected points {#sec:football}
+## In American football {#sec:football}
 
@@
 As expected, the estimated expected points increases monotonically as the teams gets closer to the endzone.
-One limitation of this approach is the linearity assumption, which results in a high negative expected point value when the offensive team is 95 yards away from the opponent's goal line.
 
@@
 Other attempts at modeling expected points in football are @goldner2012markov and @goldner2017situational, who propose a Markov framework.
-In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* that include touchdowns, field goals, and other play outcomes. 
+In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
 For any given play, the expected points are calculated using the absorption probabilities for different scoring events.
@@
 
-### In American football, 4th down strategy
+### 4th down strategy
 
@@
 Numerous other papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy.
-In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a causal inference framework to reach similar conclusions that NFL teams are too conservative in going for it on 4th down.
+In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down.
 In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.
 
-@lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
-In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e.
-1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1.
-In fact, anything up to 2 yard will recorded as 4th and 1 [@lopezDM].
-However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making.
-With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data.
-Many past analyses of the decision to go for it or not on 4th down conclude that coached in the NFL are too conservative in their decisions making. 
-@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but the effect may be overstated due to the way in which to go yardage has been recorded in the past.
+@lopez2020bigger presents an introduction to NFL tracking data, and as an example, examines 4th down behavior.
+Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.
+They demonstrate that teams are still conservative when choosing to go for it or punting on 4th down, but the effect may be overstated.
 
@@
 
-## Continuous event analysis
+## In sports with continuous states
 
@@
 This is a great leap forward from older resources such as traditional box-score results and play-by-play data.
-
-### In basketball, expected point value
 
@@
 Chalmers' shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
-A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court.
+A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court!
 
@@
 See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
-The strategy of maximizing expected points in basketball has led directly to the proliferation of three-point shooting in the NBA.
-
-### In American football, yards gained
+The strategy of maximizing expected points in basketball has lead directly to the proliferation of three-point shooting in the NBA.
 
@@
 
-### In other sports
-
 The notion of expected possession value has also been extended to association football (soccer).
@@
 Many applications can be derived from this framework, including predicting which footballer on the pitch is most likely to receive the next pass from the current on-ball player.
-
-@kumagai2021hockey offer an EPV metric for ice hockey.
 
@@
 Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
-Information derived from Table \ref{tab:erm} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13. 
+Table \ref{tab:tied} (derived from Table \ref{tab:erm}) reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13.
 The defense is in a tight spot, facing an 87% probability of losing the game. 
@@
 
-```{r, eval=FALSE}
+```{r}
 erm <- erm |>
@@
 The two win probability frameworks proposed by @Lock2014 and @Yurko2019nflwar were also implemented in @Yam2019 with minimal modifications.
-Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in American football.
+Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in gridiron football.
 
@@
   ) +
+  theme_bw() +
   theme(legend.position = "top")
@@
 On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
-@Deshpande2016 propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
+In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
 This model assumes independence of observations and constant variability in win probability.
@@
 A third crucial idea in sports analytics is the estimation of team strength. 
-First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
-Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations. 
-The methods in this Section apply equally well to multiplayer teams and single-player teams.
-
-The impetus for all methods for estimating team strength is the realization that won-loss records are a noisy measure of team strength. 
-As binary outcomes, and with all sports (except perhaps chess) involving some element of chance, wins and losses carry some signal of team strength, but we can do better.
-
-## Expected winning percentage {#sec:wpct}
-
-The first method for estimating team strength to become popular in sports analytics is a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
-Later, @miller2007derivation derived the formula as a consequence of assuming that runs (in baseball) are generated by two independent Weibull processes.
-
-Expected winning percentage is just:
-$$
-  \widehat{wpct} = \frac{X^\alpha}{X^\alpha + Y^\alpha} \,,
-$$
-where $X$ is the number of points (runs) that a team has scored, and $Y$ is the number of points (runs) that they have allowed, over some specified time period. 
-James's work was originally in baseball, and he posited the value of $\alpha = 2$. 
-The resemblance to the formula for computing the length of the hypotenuse in a right triangle provides the nod to Pythagoras.
-
-Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
-This can be done with a few lines of code, after observing that 
-$$
-  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
-$$
-and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
-Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954.
-
-```{r}
-library(tidyverse)
-library(Lahman)
-exp_wpct <- function(x, k = 2) { 
-  return(1/(1 + (1/x)^k))
-}
-
-TeamRuns <- Teams %>% 
-  filter(yearID >= 1954) %>%
-  rename(RS = R) %>% 
-  mutate(WPct = W / (W + L), run_ratio = RS/RA) %>%
-  select(yearID, teamID, lgID, WPct, run_ratio)
-
-data_space <- ggplot(
-  data = TeamRuns, aes(x = run_ratio, y = WPct)
-) +
-  geom_vline(xintercept = 1, color = "darkgray", linetype = 2) +
-  geom_hline(yintercept = 0.5, color = "darkgray", linetype = 2) +
-  geom_point(alpha = 0.2) + 
-  xlab("Ratio of Runs Scored to Runs Allowed") + 
-  ylab("Winning Percentage")
-```
-
-```{r}
-optimal <- TeamRuns %>%
-  nls(
-    formula = WPct ~ 1/(1 + (1/run_ratio)^k), 
-    start = list(k = 2)
-  ) %>%
-  coef()
-```
-
-```{r pythag, fig.cap="Winning percentages vs.
-runs scored and runs allowed in baseball, 1954--2021.
-The navy line represents the expected winning percentage model posited by James, with the exponent value of 2.
-The gold line shows the same model with an optimal exponent of 1.84."}
-data_space +
-  stat_function(fun = exp_wpct, args = list(2), size = 2, color = "navy") +
-  stat_function(fun = exp_wpct, args = list(optimal), size = 2, color = "gold")
-```
-
-Many authors have fit expected winning percentage models to other sports---too many to cite here. 
-See, for example, @hamilton2011extension for association football (soccer), @caro2013testing for Division I college football, and notably, future NBA general manager Daryl Morey for basketball [@dewan1993stats].
-
-## Bradley-Terry models
-
-Perhaps the most widely-used probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
+A popular method for estimating team strength in sports is through pairwise evaluations. 
+Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
 For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
@@
   BradleyTerry2::BTabilities() |> 
-  as_tibble(rownames = "Player") |>
-  arrange(desc(ability)) |>
-  mutate(
-    Ability = as.double(ability),
-    SE = as.double(s.e.)
-  ) |>
-  select(Player, Ability, SE) |>
-  kable(
-    booktabs = TRUE, 
-    digits = 3,
-    caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}"
-  ) |> 
+  round(3) |> 
+  kable(booktabs = TRUE, 
+        col.names = c("Ability", "SE"),
+        caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}") |> 
   kable_styling(latex_options = "hold_position")
@@
 
-## Elo ratings {#sec:elo}
-
-<!-- a basic Elo probability model for estimating the probability of $A$ beating $B$ is -->
-<!-- $$ -->
-<!-- P(A) = W(R_A - R_B) \,. -->
-<!-- $$ -->
-<!-- Here, $W$ is a continuous and monotonically increasing function that maps the difference in strength to (0, 1).  -->
-
-Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess.
-Given two players $i$ and $j$ with unknown ratings $R_i$ and $R_j$, the probability $\Pi_{ij}$ of $i$ beating $j$ is defined as
-$$
-\Pi_{ij} = \frac{1}{1 + K^{(R_j - R_i)/400}} \,.
-$$
-In this formula, $K$ is commonly known as the $K$-factor, or development coefficient.
-The International Chess Federation (FIDE) uses $K=10$ for players with any previously achieved rating of at least 2400.
-For those that have never reached a rating of 2400 or better, $K=20$ is assigned. 
-Finally, $K=40$ is given to new players with under 30 games played, and players under the age of 18 with rating fewer than 2300 [@fide2022rating].
-
-Another interpretation for $\Pi_{ij}$ is the expected score of the game for player $i$.
-The scores of 0, 0.5, and 1 are associated with three possible game outcomes loss, tie, and win, respectively.
-After a game, the updated Elo rating $R^*_i$ for player $i$ is
-$$
-R^*_i = R_i + K(S_i - \Pi_{ij}) \,,
-$$
-where $S_i \in \{0, 0.5, 1\}$.
-When a tournament concludes, a post-tournament rating is obtained for each player based on the rating updates for all games played.
-
-```{r}
-library(chessR)
-us <- c("fatgreggy")
-greg <- get_raw_chessdotcom(
-  usernames = us,
-  year_month = 202206
-)
-# greg: black
-```
-
-To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR] (see Section \ref{sec:chess}).
-Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas his opponent was rated `r greg[6, "WhiteElo"]`.
-Since both ratings are below 2400, we apply a development coefficient of $K=20$ to this example.
-The probability of the author ($a$) defeating their opponent ($b$) was
-<!-- don't know how to include `r` inside latex -->
-$$
-\Pi_{ab} = \frac{1}{1 + 20^{(1683 - 1732)/400}} = 0.591 \,.
-$$
-The author won the match: that outcome is associated with a score of $S_a = 1$.
-The post-game Elo rating for the author is thus
-$$
-R^*_a = 1732 + 20(1 - 0.591) = 1740 \,.
-$$
-
-Besides chess, the Elo system has also been implemented to estimate team strength in other sports. 
-See @koning2017rating for more information on applications of the Elo rating in soccer, and @kovalchik2019calibration and @kovalchik2016searching for Elo ratings in tennis. 
-Furthermore, Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at [FiveThirtyEight.com](https://fivethirtyeight.com).
-
-## Bayesian state-space models
+Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
+See @koning2017rating for more information on applications of the Elo rating in soccer.
 
@@
 See also @glickman2017estimating for more discussion on estimating team strengths in American football. 
-
 More recently, @Lopez2018 extend @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
@@
 Apart from sports gambling, having an accurate estimate of team strength is useful to team officials, who are constantly monitoring and forecasting their team's ability.
-
-@koopman2015dynamic uses a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.
-
-# Non-performance data
-
-Most of the research in sports analytics is fueled by the analysis of data recorded from the outcome of sports contests.
-However, a growing body of literature is informed by non-performance data about sports. 
-This research includes fields like sports economics and kinesiology that are subfields of mature disciplines (a treatment of either is well beyond the scope of the this paper).
-In addition to this body of research, we note that today, professional sports teams are collecting biometric data on their own players, and presumably using that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance.
-While this research may constitute "emerging methodologies," it will likely never be published, because each team will use it to their competitive advantage.
-
-Nevertheless, in this Section we discuss concepts stemming from one notable, relatively new source of (semi-)public non-performance sports data: data from betting markets.
-Since the 2018 United States Supreme Court decision in [*Murphy v.
-National Collegiate Athletic Association*](https://en.wikipedia.org/wiki/Murphy_v._National_Collegiate_Athletic_Association), sports gambling has exploded in the U.S.
-The increasing interest in sports gambling has led to increasing interest in sports gambling *data*, and that data has proven useful to researchers in two major ways.
-
-First, betting market data is probably the best source for estimating the true probability of a team winning a game. 
-The efficiency of betting market data in this respect has been demonstrated time and time again.
-The utility of these estimates have then informed research that has helped us learn about the sports themselves.
-In this sense, data generated by sports gambling has been an important source of data useful for sports analytics (see Section \ref{sec:betting-market-data}).
-
-Second, sports analytics researchers have studied various types of sports gambling outlets (including fantasy sports). 
-This research has estimated probabilities, evaluated common strategies, and offered optimized approaches for a variety of different games of chance (see Section \ref{sec:pools}).
-
-## Sports betting market data {#sec:betting-market-data}
-
-While @Lopez2018 uses betting market data to model team strengths, it does not directly address strategies for betting or inefficiencies in betting markets. Early work by @gandar1988testing examines 
-Related work such as @lacey1990estimation and @Boulier2006 both explore the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
-Neither paper finds strong evidence for inefficiencies in the markets.
-@boulier2003predicting compared the predictive performance of power rankings and media experts to the betting market for NFL games and found that the betting market was the best for predicting winners.
-More recently @Buttrey2016 explored the NHL betting market.
-They produced a model to predict win probabilities in given games, and games where the predicted probability differed from the market price bets were placed.
-They find that using their methods, that they were able to produce a positive return on investment.
-
-@Boulier2006 that there is insufficient evidence to conclude that betting markets were inefficient from 1994--2000.
-@boulier2003predicting test several predictors for NFL games and find that the betting market is the best predictor.
-@Buttrey2016 demonstrate a method that produces a positive ROI.
-@gandar1988testing conclude that stats test are too weak to reject the hypothesis of rationality.  But they do reject rationalist with ``economic tests". 
-@lacey1990estimation : Finds the market to be efficient, though they did find some profitable betting rules over the period of time between 1984 and 1986
-
-## Analytics for sports betting pools {#sec:pools}
-
-Another popular type of betting pool is a survivor pool.
-@Bergman2017SurvivingAN presents formal optimization approaches for NFL survivor pools and concludes that planning for only part of the season yields optimal results in terms of maximizing survival probability.
-
-@breiter1997play
-
-@clair2007optimal
-
-@kaplan2001march
-
-@metrick1996march
-
-@niemi2008contrarian
-
-@sauer1998economics
-
-@imbrogno2022computing
-Description: The average number of winning survivor pool entries is computed for the 2019 National Football League (NFL) season.
-The probability of having to share the winning pot, given realistic expectations on game outcomes, is also shown.
-
-@nichols2014impact conclude that the impact of travel is not completely incorporated into the betting markets, but that any effect is too small to find any profitable advantage.
-
-@PaulWeinbach2014 Market efficiency and behavioral biases in the wnba betting market
-
-@soebbing2013gamblers
-
-@spann2009sports NBA bettors believe that tanking is occurring.
 
@@
 Software tools for sports analytics are similarly numerous. 
-For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network (CRAN) and organizes them by sport [@baumer2022ctv].
-Table \ref{tab:ctv} provides an overview of the currently available sport-specific CRAN packages.
-@casals2022systematic offers a systematic review of sport-related packages on CRAN.
-Further, a more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].
-
-```{r}
-tibble(
-  "American Football" = "nflverse, nflfastR, nflreadr, nfl4th, nflseedR,\n nflplotR, NFLSimulatoR, fflr, ffscrapr, ffsimulator, gsisdecoder, cfbfastR",
-  "Association Football (Soccer)" = "worldfootballR, engsoccerdata, socceR, ggsoccer, footballpenaltiesBL, footBayes, itscalledsoccer, FPLdata, EUfootball",
-  "Australian Rules Football" = "fitzRoy",
-  "Baseball/Softball" = "Lahman, retrosheet, pitchRx, mlbstats, baseballDBR, baseballr, runexp",
-  "Basketball" = "BAwiR, AdvancedBasketballStats, uncmbb, BasketballAnalyzeR, NBAloveR, wehoop, hoopR, toRvik",
-  "Chess" = "chess, stockfish, bigchess, rchess, chessR",
-  "Cricket" = "yorkr, cricketr, cricketdata, howzatR",
-  "Esports" = "CSGo, rbedrock, ROpenData, opendotaR, RDota2",
-  "GPS Activity Tracking" = "trackeR, trackeRapp, rStrava",
-  "Hockey" = "hockeyR, NHLData, nhlapi, nhlscrape, fastRhockey",
-  "Swimming" = "SwimmeR",
-  "Track and Field" = "combinedevents, JumpeR",
-  "Volleyball" = "volleystat"
-) |> 
-  pivot_longer(
-    everything(),
-    names_to = "Sport",
-    values_to = "Packages"
-  ) |>
-  transmute(
-    Sport,
-    "Number of Packages" = str_count(Packages, ",") + 1,
-    "List of Packages" = Packages
-  ) |>
-  arrange(desc(`Number of Packages`)) |>
-  kable(
-    booktabs = TRUE,
-    linesep = "",
-    caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN).
-While the major North American sports dominate the list, perhaps the fastest-growing collection is for esports.
-\\label{tab:ctv}"
-  ) |> 
-  column_spec(3, width = "7cm")
-```
+For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network and organizes them by sport [@baumer2022ctv].
+A more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].
 
@@
 This type of infrastructure, where one package is the "workhorse" that facilitates common generic data operations, and other packages layer on specific functionality, is common in sports analytics.
-See Section \ref{sec:elo} for an example of how the \pkg{chessR} package can be used to compute Elo ratings.
 
-Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package).
-We note how the contextual information provided by the chessboard is instrumental in helping the reader contextualize the data. 
-(How many of us can visualize PGN directly?)
-In Section \ref{sec:graphics}, we outline a collection of graphical tools that provide similar context for different playing surfaces.
+Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in a game recently won by one of the authors (downloaded via the \pkg{chessR} package).
 
-```{r chess, eval=whoami::username() == 'bsbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
 At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
@@
 
-## Graphical tools {#sec:graphics}
-
-Creating effective data graphics is a key component of statistical communication, and sports is no exception. 
-We highlight a few packages that assist with the creation of data graphics about sports.
+## Theming
 
@@
 For example, the NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
-Figure \ref{fig:mlb_standings} illustrates how the use of team colors, which have a natural association for many sports fans, can help to untangle what would otherwise be messy data graphics. 
-In Figure \ref{fig:mlb_standings}, 30 different lines are plotted on top of one another, crisscrossing and intersecting in various unpredictable ways. 
-However, the use of team colors to identify the lines makes it possible to follow the trajectory of most teams over the course of the season.
+Figure \ref{fig:teamcolors} illustrates the full array of team color palettes for professional sports leagues.
 
-```{r}
-# remotes::install_github("beanumber/retro")
-library(retrosheet)
-retro_gl <- get_retrosheet("game", 2021)
-games_2021 <- retro_gl |>
-  mutate(home_win = HmRuns > VisRuns) %>%
-  select(Date, VisTm, HmTm, home_win) %>%
-  pivot_longer(
-    -c(Date, home_win),
-    names_to = "is_home", values_to = "team_id", 
-  ) %>%
-  mutate(
-    is_win = ifelse(
-      is_home == "HmTm", home_win, abs(1 - home_win))
-  ) %>%
-  group_by(team_id) %>%
-  mutate(
-    cum_W = cumsum(is_win), 
-    cum_L = cumsum(!is_win), 
-    cum_G = cum_W + cum_L, 
-    G_over_500 = cum_W - cum_L, 
-    last_game = cum_G == max(cum_G)
-  )
-retro_teams <- retrosheet::getTeamIDs(2021) |>
-  enframe(name = "mascot", value = "retro_id")
-mlb_pal <- teamcolors |>
-  filter(league == "mlb", str_detect(division, "NL")) |>
-  inner_join(retro_teams, by = "mascot") |>
-  select(retro_id, primary) |>
-  deframe()
-games_2021 <- games_2021 |>
-  filter(team_id %in% names(mlb_pal))
+```{r teamcolors, warning=FALSE, fig.cap="Professional sports team color palettes provided by the \\pkg{teamcolors} package. ", fig.height=9}
+teamcolors::show_team_col()
 ```
-
-```{r mlb_standings, fig.cap="The progression of National Leauge team standings during the 2021 Major League Baseball season.
-Note how the use of team colors makes it possible to untangle what would otherwise be a messy jumble of indistinguishable lines.
-Data provided by \\pkg{retrosheet} and colors provided by \\pkg{teamcolors}."}
-ggplot(
-    games_2021, 
-    aes(x = cum_G, y = G_over_500, color = team_id)
-  ) + 
-  geom_hline(yintercept = 18, color = "darkgray", lty = 2) + 
-  geom_line() +  
-  ggrepel::geom_label_repel(
-    data = filter(games_2021, last_game), 
-    aes(label = team_id)
-  ) + 
-  annotate("text", x = 2, y = 22, label = "90 wins", color = "darkgray", hjust = "left") +
-  scale_color_manual(
-    guide = "none", name = "Team",
-    values = mlb_pal
-  ) +
-  scale_y_continuous("Games above .500") + 
-  scale_x_continuous("Cumulative Games Played") + 
-  labs(title = "Daily MLB Standings, 2021", 
-       caption = "Source: Retrosheet")
-```
-
-\pkg{nflplotR} [@R-nflplotR] has a similar goal to \pkg{teamcolors}.
-It also provides \pkg{ggplot2} extensions but is designed specifically for the NFL.
-A great feature of \pkg{nflplotR} is the collection of `geom_*()` (geometric object) functions that enhance high-quality plotting of NFL team logos and player images with \pkg{ggplot2}.
-Figure \ref{fig:nflplotr} shows a scatterplot of offensive and defensive expected points added for NFL teams in the 2021 regular season.
-The logos of all 32 American football clubs are plotted in place of the usual dots, making it easier for the reader to identify which team each data point represents.
-
-```{r nflplotr, fig.cap="Offensive and defensive expected points added per play for the 2021 NFL regular season, plotted with \\pkg{nflplotR} using data from \\pkg{nflfastR}."}
-pbp <- nflreadr::load_pbp(2021) %>% 
-  dplyr::filter(season_type == "REG") %>%
-  dplyr::filter(!is.na(posteam) & (rush == 1 | pass == 1))
-
-offense <- pbp %>%
-  dplyr::group_by(team = posteam) %>%
-  dplyr::summarise(off_epa = mean(epa, na.rm = TRUE))
-
-defense <- pbp %>%
-  dplyr::group_by(team = defteam) %>%
-  dplyr::summarise(def_epa = mean(epa, na.rm = TRUE))
-
-combined <- offense %>%
-  dplyr::inner_join(defense, by = "team")
-
-combined |> 
-  ggplot(aes(off_epa, def_epa)) +
-  nflplotR::geom_mean_lines(aes(v_var = off_epa , h_var = def_epa)) +
-  nflplotR::geom_nfl_logos(aes(team_abbr = team), width = 0.08, alpha = 0.8) +
-  labs(x = "Offensive EPA/play",
-       y = "Defensive EPA/play") +
-  scale_y_reverse()
-```
-
-Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, board, or pitch for a particular sport.
-Orienting these points graphically may require drawing a complex set of guidelines that provide that context to readers. 
-Thankfully, the \pkg{sportyR} package [@R-sportyR] contains generic playing surfaces for baseball, basketball, curling, American football, ice hockey, soccer, and tennis that can be added to \pkg{ggplot} graphics with a single function call.
-Consider, for example, how much time the surfaces plotted in Figure \ref{fig:sportyR} would save you, and how helpful they would be in contextualizing player tracking data (such as those shown in Figure \ref{fig:epv-demo}).
-With the increased availability of player tracking data, this particular tool should see increased usage.
-
-```{r sportyR, fig.show="hold", out.width="49%", fig.cap="At left, an NBA basketball court drawn by \\pkg{sportyR}. At right, an NHL hockey rink drawn by \\pkg{sportyR}."}
-library(sportyR)
-geom_basketball(league = "NBA")
-# geom_tennis(league = "USTA")
-geom_hockey(league = "NHL")
-```
-
-## Case study in the evolution of tools and research: baseball {#sec:baseball}
-
-As the granularity of baseball data has evolved over time, so too have the statistical methodologies for modeling that data, and the tools for working with it.
-
-For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
-These data, now available through the \pkg{Lahman} package [@R-Lahman], were sufficient to study broad trends in baseball, and led to insights such as the value of expected winning percentage (see Section \ref{sec:wpct}) and the importance of on-base percentage relative to batting average.
-These relatively simple insights fueled the "Moneyball" [@lewis2004moneyball] era revolution in sports analytics [@baumer2014sabermetric].
-
-Over time, the resolution of baseball data has improved to include play-by-play data, pitch-by-pitch data, and now player tracking data.
-
-The \pkg{retrosheet} package [@R-retrosheet] now provides access to the historical play-by-play data available from Retrosheet (this is a comprehensive version of what Lindsay collected for his research).
-This play-by-play data allowed researchers to learn about strategies, like those that we discussed in Section \ref{sec:ev}.
-In baseball, this deepened our understanding of bunting, stolen bases, handedness, batting order, and many other aspects of the game. 
-Play-by-play data underlies much of the analysis in @Tango2007book.
-
-The \pkg{pitchRx} package [@R-pitchRx] provides access to pitch-by-pitch data that fueled innovative research into catcher framing [@deshpande2017hierarchical], pitch values [@healey2019bayesian], and pitch classification [@sidle2018using].
-Catcher framing is a notable example of a concept that scouts talked about for decades, but that could not be quantified by analysts until data of the appropriate resolution became available.
-
-While play-by-play data allows us to make valuations *between plays*, player tracking data allows us to make valuations *within plays*. 
-The \pkg{baseballr} package [@R-baseballr] now provides access to player tracking data from Statcast.
-These data have led to investigations into how defensive shifts affect batting performance [@bouzarth2021swing], as well as how launch angles affect the probability of hitting a home run [@marchi2018analyzing].
-
-As we saw above with chess, the packages in baseball fit together in creative ways. 
-In Figure \ref{fig:mlb_standings}, we showed how \pkg{teamcolors} can illuminate data pulled from \pkg{retrosheet} to make an informative data graphic. 
-One could just as easily use \pkg{sportyR} to generate a field graphic, and then overlay player tracking data obtained from \pkg{baseballr} to depict defensive shifts.
-
-Thus, these R packages enable research by making data more easily available. 
-Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
-[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts (see Section \ref{sec:opportunities}).]
 
@@
 
-Public research in sports analytics is driven in part by several notable competitions and conferences. 
-These venues have been an important source of new ideas and have contributed to the diversification of the field by breaking down barriers to entry.
+Public research in sports analytics is driven in part by several notable competitions. 
+These competitions have been an important source of new ideas and have contributed to the diversification of the field by breaking down barriers to entry.
 
@@
 
-Similar competitions that provide opportunities for aspiring researchers to tackle sports analytics problems include the [Big Data Cup](https://www.stathletes.com/big-data-cup/) for ice hockey and the [SABR Diamond Dollars Case Competition](https://sabr.org/analytics/case) for baseball, and the [Big Data Derby](https://www.kaggle.com/competitions/big-data-derby-2022) for horse racing.
+Similar competitions that provide opportunities for aspiring researchers to tackle sports analytics problems include the [Big Data Cup](https://www.stathletes.com/big-data-cup/) for ice hockey and the [SABR Diamond Dollars Case Competition](https://sabr.org/analytics/case) for baseball.
 
-The biennial [New England Symposium on Statistics in Sports](https://www.nessis.org/) is likely the longest-running academic conference devoted to sports analytics. 
-It's West Coast counterpart is [The Cascadia Symposium on Statistics in Sports](http://www.cascadiasports.com/). 
-Many influential results have been showcased for the first time at these conferences. 
-Other prominent sports analytics conferences include the  [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2022/), [UConn Sports Analytics Symposium](https://statds.org/events/ucsas2022/), and [MathSport International](http://www.mathsportinternational.com/).
-
-The highest-profile sports analytics conference is undoubtedly the [Sloan Sports Analytics Conference](https://www.sloansportsconference.com/), which draws academics, industry professionals, vendors, and media organizations.
-While the conference holds a research competition and has certainly drawn attention to sports analytics, it has also been criticized for a variety of shortcomings. 
-These criticisms include a lack of emphasis on reproducibility in the research competition, high ticket prices, the large salaries taken by the organizers, and the lack of diversity among attendees and presenters [@funt2022].
-
-# Conclusion {#sec:conclusion}
+# Conclusion
 
@@
 In this paper, we explore three of those big questions:
-Who are the best teams and how good are they? 
-What is the likelihood of each team winning the game at any given juncture? 
-Is there a generic framework for evaluating strategies at any given juncture in a game?
 
-Other fundamental questions are addressed elsewhere. 
-How significant is the element of chance in a particular sport? 
-Given that we know who the best teams are, who are the best players and how can we quantify their relative contributions? 
-What combinations of players work best together in a particular sport? In particular, see @Lopez2018 for estimations of the element of chance across four major sports. 
+- Who are the best teams and how good are they? 
+- What is the likelihood of each team winning the game at any given juncture? 
+- Is there a generic framework for evaluating strategies at any given juncture in a game?
+
+Other fundamental questions, such as:
+
+- How significant is the element of chance in a particular sport? 
+- Given that we know who the best teams are, who are the best players and how can we quantify their relative contributions? 
+- What combinations of players work best together in a particular sport?
+
+are addressed elsewhere. 
+In particular, see @Lopez2018 for estimations of the element of chance across four major sports. 
 The second question is often addressed using a formulation of *wins above replacement* (WAR)---see @baumer2015openwar and @Yurko2019nflwar for details in baseball and American football. 
-Recent work by @che2022athlete also addresses this question across sports. 
 The third question is most compelling in sports like basketball, ice hockey, and soccer, where substitutions are common and it is obvious that different combinations of players with different sets of skills will result in squad of varying strengths and weaknesses.

### 2023-01-03 response_to_reviewers_R1.Rmd
add chess figures; #44

@@
 - Section 6.1, line 53, Figure ??
+
+> This is an artifact of the attempt to track changes and does not appear in the unannotated manuscript.
+
 - Section 6.1, line 56, contextual is used twice in that sentence, perhaps “helping the reader visualize” or “helping the reader understand”
@@
 - Figure 6, in my version the images are stacked rather than left/right
+
+> This is an artifact of the attempt to track changes and does not appear in the unannotated manuscript.
+
 - Section 6.2, line 46 This sentence that starts “Consider, for example, “ seems out of place with the rest of the tone of this article
+
+> TODO
+
 - Section 7, line 7 or 8, it might be worth mentioning the growth of small conferences like UCSAS that have sprung up, something like “As the field of sports analytics has grown, there has been a proliferation of sports specific and regional sports analytics conferences.”

### 2023-01-03 sports_analytics.Rmd
add chess figures; #44

@@
 In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
-And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact [@healey2017new].
+And, of course, in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact [@healey2017new].
 
@@
 Every sport is different, with its own set of rules, strategies, methods of data collection, number of players, and the magnitude of the role of chance. 
-At the same time, many sports are similar, either because one evolved from the other, or the structure of the game shares certain attributes. 
-Sports that are closely related anthropologically may or may not share common applications of analytical methods. 
+At the same time, many sports are similar, either because one evolved from the other, or the structure of the games share certain attributes. 
+Sports that are closely related historically may or may not share common applications of analytical methods. 
 For example, despite belonging to the same bat-and-ball family, baseball and cricket differ in strategies such as batting order or sacrifice plays.
@@
 
+@macdonald2012expected uses expected goals to evaluate ice hockey players, but does not have access to player tracking data necessary to evaluate possessions.
 @kumagai2021hockey offer an EPV metric for ice hockey.
@@
 Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package).
-We note how the contextual information provided by the chessboard is instrumental in helping the reader contextualize the data. 
+We note how the contextual information provided by the chessboard is instrumental in helping the reader understand the data. 
 (How many of us can visualize PGN directly?)
@@
 
-```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
-At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
+```{r chess-fig, eval=whoami::username() == 'bsbaumer'}
 library(chess)
 # Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
+pdf("chess-1.pdf")
 game() |> plot()
+dev.off()
 
@@
 # plot_moves(m, interactive = FALSE)
+
+pdf("chess-2.pdf")
 plot(m)
+dev.off()
+```
+
+```{r chess, fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package.
+At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
+knitr::include_graphics("chess-1.pdf")
+knitr::include_graphics("chess-2.pdf")
 ```
@@
 
+As the field of sports analytics has grown, a proliferation of sports specific and regional sports analytics conferences have arisen.
 The biennial [New England Symposium on Statistics in Sports](https://www.nessis.org/) is likely the longest-running academic conference devoted to sports analytics. 

### 2023-01-03 response_to_reviewers_R1.Rmd
address R1; #43

@@
 ## Major comments from Reviewer 1
+
+Overall, the paper is now much more comprehensive, complete, and clear; I think it
+will serve as a very useful introduction to the field of sports analytics. I have only a few
+suggestions and points of confusion to share, which I hope the authors will examine before
+submitting the final version. They are all fairly minor.
+
+- If there is still space allowed in the abstract, I think you could consider explicitly
+mentioning the three big ideas (expected value of a game state, win probability, and
+measures of team strength). This would give readers a better idea of what is to come
+and also allow people to find your article more easily if they are interested in one of
+those specific topics.
+- In Section 2.1.2, I am a little confused as to what Carroll et. al. did to extend the
+previous model of Carter and Machol, since both efforts seem to involve some sort of
+linearity assumption.
+- Also in Section 2.1.2, it might be worthwhile to briefly define absorbing states, as I for
+one didn’t know the definition before looking it up (and as it turns out the concept is
+quite simple)
+- In Section 2.1.3, you make the comment “they estimate that a better strategy would
+be worth about 0.4 wins per season on average, a substantial amount in a season with
+only 16 games.” This makes sense, but previously in Section 2.1.2 you mentioned
+a potential difference of 3.8 wins per season, yet didn’t comment on the significant
+magnitude of that effect. I’m not advocating for a particular change here, but the
+authors could consider adding a small comment about the size of the effect Pelechrinis
+et. al. estimated.
+- At the end of Section 2.1.3, you write “Lopez (2020) also concludes that coaches are
+too conservative on 4th down decision making, but the effect may be overstated due
+to the way in which to go yardage has been recorded in the past.” It’s unclear to me
+whether the overstated effect is the one Lopez estimates in his paper, or if you are
+referring to the estimated effects past studies which used less granular data.
+- In Section 2.3, you write “NHL coaches do not always employ the optimal strategies”,
+but it isn’t clear what those strategies are in reference to the previous discussion. Is
+pulling the goalie optimal?
+- In the second formula of Section 4.1, the right hand side of the equation has a 2, but
+it should be an α.
+
+> Thank you. This is fixed.
+
+- I think Section 5.2 would benefit from a short introductory paragraph, perhaps which
+introduces different types of sports betting, thus leading into the betting pools discus-
+sion of Section 5.2.1.
+- I notice you have several hyperlinks throughout–I’m not sure what the Wiley standards
+are, but it might be helpful to distinguish links with a different color.
+
+> Indeed. We await Wiley's instructions when we reach the typesetting phase.
+
+## Minor comments from Reviewer 1
+
+- Page 5, lines 33-34: ‘might reasonably’ is used twice in the same sentence
+
+> We have changed the first "might reasonably" to "could"
+
+- Page 6, line 23: ‘topics’ instead of ‘topic’
+
+> Fixed.
+
+- Page 6, line 38: ‘go for it’ instead of ‘got for it’
+
+> Fixed.
+
+- Page 6, lines 45-46: ‘coaches’ instead of ‘coached’ and ‘decision making’ instead of
+‘decisions making’
+
+> Fixed.
+
+- Page 7, line 7: ‘kickoffs’ instead of ‘kickoff’
+
+> Fixed.
+
+- Page 8, line 51: Might be better to say “Earlier in Section 2” as opposed to just “In
+Section 2”
+
+> Fixed.
+
+- Page 11, line 56 and Page 12, line 5: ‘win-loss’ instead of ‘won-loss’
+
+> Fixed.
+
+- Page 13, line 53: This is probably a matter of personal preference but “rating less than
+2300” sounds more natural than “rating fewer than 2300”
+
+> Fixed.
+
+- Page 14, line 35: ‘use’ instead of ‘uses’
+
+> Fixed.
+
+- Page 17, line 27: I think the question in parentheses should be placed before the period
+of the previous sentence
+
+> We moved this.
+
+- Page 18, line 40: ‘Lindsey’ instead of ‘Lindsay’
+
+> Fixed.
+
+- Page 20, lines 13-15: Not sure why there are brackets around this sentence.
+
+> We removed the brackets.
+
+- Page 20, line 41: ‘Its’ not ‘It’s’
+
+> Fixed.
 

### 2023-01-03 sports_analytics.Rmd
address R1; #43

@@
 Second, the characterization of $S$ as having 24 states is only the simplest possible. 
-The inning, or the score of the game, or even the weather, might reasonably be incorporated into $S$, as those conditions might reasonably affect the estimate of $\mathbb{E}[X | s]$. 
+The inning, or the score of the game, or even the weather, could be incorporated into $S$, as those conditions might reasonably affect the estimate of $\mathbb{E}[X | s]$. 
 More definitively, the identity of the current batter, pitcher, or batter on deck, might also affect the estimate of $\mathbb{E}[X | s]$. 
@@
 
-The concept of expected points in American football has many applications.  One of the most notable and well-studied topic is the evaluation of 4th down strategy.
+The concept of expected points in American football has many applications.  One of the most notable and well-studied topics is the evaluation of 4th down strategy.
 There is near universal consensus in the literature that NFL teams have been too conservative in the past when making 4th down decisions.
@@
 @lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
-In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e.
+In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "go for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e.
 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1.
@@
 With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data.
-Many past analyses of the decision to go for it or not on 4th down conclude that coached in the NFL are too conservative in their decisions making. 
+Many past analyses of the decision to go for it or not on 4th down conclude that coaches in the NFL are too conservative in their decision making. 
 @lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but the effect may be overstated due to the way in which to go yardage has been recorded in the past.
@@
 Another application of expected points is to evaluate kickoff decisions made by football coaches, as demonstrated by @urschel2011. 
-Specifically, they look at surprise on-sides kicks versus regular kickoff and the decision to accept a touchback versus returning the kickoff.
+Specifically, they look at surprise on-sides kicks versus regular kickoffs and the decision to accept a touchback versus returning the kickoff.  
 Using data from the 2009 NFL season, they conclude, as many have, that coaches in the NFL tend to make conservative decisions.
@@
 
-In Section \ref{sec:ev}, we defined the expected value of a possession based on the state $s$ of the game in terms of the expected number of points (runs) $X$ that would be scored in the remainder of some period of time.
+Earlier in Section \ref{sec:ev}, we defined the expected value of a possession based on the state $s$ of the game in terms of the expected number of points (runs) $X$ that would be scored in the remainder of some period of time.
 We then showed how this value could be used to analyze the relative effectiveness of certain strategies, with the simple idea that strategies that yield higher expected values are preferable. 
@@
 A third crucial idea in sports analytics is the estimation of team strength. 
-First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
+First, we briefly introduce a simple method for estimating team strength based on win-loss record. 
 Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations. 
@@
 
-The impetus for all methods for estimating team strength is the realization that won-loss records are a noisy measure of team strength. 
+The impetus for all methods for estimating team strength is the realization that win-loss records are a noisy measure of team strength. 
 As binary outcomes, and with all sports (except perhaps chess) involving some element of chance, wins and losses carry some signal of team strength, but we can do better.
@@
 $$
-  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
+  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^\alpha}
 $$
@@
 The International Chess Federation (FIDE) uses $K=10$ for players with any previously achieved rating of at least 2400.
-For those that have never reached a rating of 2400 or better, $K=20$ is assigned. 
-Finally, $K=40$ is given to new players with under 30 games played, and players under the age of 18 with rating fewer than 2300 [@fide2022rating].
+Finally, $K=40$ is given to new players with under 30 games played, and players under the age of 18 with rating less than than 2300 [@fide2022rating].
 
@@
 
-@koopman2015dynamic uses a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.
+@koopman2015dynamic use a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.
 
@@
 
-While @Lopez2018 uses betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
+While @Lopez2018 use betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
 Early work by @gandar1988testing examines the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
@@
 Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package).
-We note how the contextual information provided by the chessboard is instrumental in helping the reader understand the data. 
-(How many of us can visualize PGN directly?)
+We note how the contextual information provided by the chessboard is instrumental in helping the reader understand the data
+(How many of us can visualize PGN directly?).
 In Section \ref{sec:graphics}, we outline a collection of graphical tools that provide similar context for different playing surfaces.
@@
 
-For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
+For example, before George Lindsey's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
 These data, now available through the \pkg{Lahman} package [@R-Lahman], were sufficient to study broad trends in baseball, and led to insights such as the value of expected winning percentage (see Section \ref{sec:wpct}) and the importance of on-base percentage relative to batting average.
@@
 Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
-[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts (see Section \ref{sec:opportunities}).]
+Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts (see Section \ref{sec:opportunities}).
 
@@
 The biennial [New England Symposium on Statistics in Sports](https://www.nessis.org/) is likely the longest-running academic conference devoted to sports analytics. 
-It's West Coast counterpart is [The Cascadia Symposium on Statistics in Sports](http://www.cascadiasports.com/). 
+Its West Coast counterpart is [The Cascadia Symposium on Statistics in Sports](http://www.cascadiasports.com/). 
 Many influential results have been showcased for the first time at these conferences. 

### 2023-01-03 sports_analytics.Rmd
added big ideas to abstract; #43

@@
   Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years.
-  In this paper, we connect three big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
+  In this paper, we connect three big ideas that are common across multiple sports: the expected value of a game state, win probability, and measures of team strength.
+  For each, we explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
   While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources. 

### 2023-01-03 response_to_reviewers_R1.Rmd
added big ideas to abstract; #43
added scope to abstract; #44

@@
 
+> Thank you for your feedback.
+
 - If there is still space allowed in the abstract, I think you could consider explicitly
@@
 those specific topics.
+
+> Thank you for this suggestion. We have added the three ideas explicitly to the abstract.
+
 - In Section 2.1.2, I am a little confused as to what Carroll et. al. did to extend the
@@
 
-The authors have now made clearer the limitation in scope of this manuscript related to player evaluation. Perhaps, a statement regarding this can be added to the abstract.
+- The authors have now made clearer the limitation in scope of this manuscript related to player evaluation. Perhaps, a statement regarding this can be added to the abstract.
 
-I think that parts of the first paragraph of Section 5 regarding biometric data is a bit out of place.
+> We have a statement to this effect to the abstract.
+
+- I think that parts of the first paragraph of Section 5 regarding biometric data is a bit out of place.
 This could be moved to Section 7 (Opportunities).

### 2023-01-03 response_to_reviewers_R1.Rmd
reworded sentence; #44

@@
 
-> TODO
+> Thank you for this suggestion.
+We have reworded this sentence to read more like the rest of the article: "The surfaces plotted in Figure 9 are helpful in contextualizing player tracking data (such as those shown in Figure 3) and would be laborious for each analyst to have to create on their
+own."
 
@@
 
-> We have added a similar sentence to this effect.
+> Thank you for this suggestion. We have added a similar sentence to this effect: "As the field of sports analytics has grown, a proliferation of sports specific and regional sports analytics
+conferences have arisen."

### 2023-01-03 sports_analytics.Rmd
added scope to abstract; #44
reworded sentence; #44

@@
   Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
+  We also describe the use of non-performance data and opportunities to share analytical work, but omit an in-depth discussion of individual player evaluation as beyond our scope.
   This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
@@
 Thankfully, the \pkg{sportyR} package [@R-sportyR] contains generic playing surfaces for baseball, basketball, curling, American football, ice hockey, soccer, and tennis that can be added to \pkg{ggplot} graphics with a single function call.
-Consider, for example, how much time the surfaces plotted in Figure \ref{fig:sportyR} would save you, and how helpful they would be in contextualizing player tracking data (such as those shown in Figure \ref{fig:epv-demo}).
+The surfaces plotted in Figure \ref{fig:sportyR} are helpful in contextualizing player tracking data (such as those shown in Figure \ref{fig:epv-demo}) and would be laborious for each analyst to have to create on their own.
 With the increased availability of player tracking data, this particular tool should see increased usage.

### 2023-01-03 response_to_reviewers_R1.Rmd
defined absorbing state; #43

@@
 quite simple)
+
+> We have added a sentence defining what an absorbing state is: "An absorbing state is a link in a Markov chain from which there are no possible transitions (i.e., it is the end of the chain)."
+
 - In Section 2.1.3, you make the comment “they estimate that a better strategy would

### 2023-01-03 sports_analytics.Rmd
defined absorbing state; #43

@@
 
-[^state]: 25, if you include the absorbing state of 3 outs.
+[^state]: 25, if you include the absorbing state of 3 outs that describes the end of an inning.
 
@@
 Other attempts at modeling expected points in football are @goldner2012markov and @goldner2017situational, who propose a Markov framework.
-In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* that include touchdowns, field goals, and other play outcomes. 
+In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* that include touchdowns, field goals, and other possession outcomes. 
+An absorbing state is a link in a Markov chain from which there are no possible transitions (i.e., it is the end of the chain).
 For any given play, the expected points are calculated using the absorption probabilities for different scoring events.

### 2023-01-03 response_to_reviewers_R1.Rmd
added intro paragraph on betting; #43

@@
 - I think Section 5.2 would benefit from a short introductory paragraph, perhaps which
-introduces different types of sports betting, thus leading into the betting pools discus-
-sion of Section 5.2.1.
+introduces different types of sports betting, thus leading into the betting pools discussion of Section 5.2.1.
+
+> Thank you for this suggestion. We have a added a short paragraph to this effect.
+
 - I notice you have several hyperlinks throughout–I’m not sure what the Wiley standards

### 2023-01-03 sports_analytics.Rmd
added intro paragraph on betting; #43

@@
 
+Many different types of bets can be placed on sports. 
+For individual contests, bets involve point spreads, moneylines, odds, or other ways of handicapping which team will win. 
+Money can also be wagered on futures, where odds are given in advance for events that may or may not transpire (e.g., a certain team making the playoffs, or a certain player winning the MVP award). 
+Here, we focus on betting pools, in which a group of people compete to predict winners in multiple contests (often a tournament), and the inevitable question of whether strategies exist that will consistently beat the market.
+
 ### Betting pools

### 2023-01-03 sports_analytics.Rmd
goalie-pulling; #43
fix wording; close #47

@@
 Early work on expected point values in American football can also be found in @Carroll1988hidden.
-In particular, the authors extend @Carter1971operations's work and propose a linear model for expected points in the NFL.
+In particular, the authors consider a similar approach to @Carter1971operations and propose a linear model for expected points in the NFL.
 They determine that every extra 25 yards is associated with 2 more points scored on average for a football team.
@@
 The hope is to score a quick goal to get back in the game, but the risk is falling further behind. 
-@beaudoin2010strategies show that NHL coaches do not always employ the optimal strategies. 
+@beaudoin2010strategies show that NHL coaches do not always employ the optimal strategies, usually by waiting too long to pull their goalies. 
 @skinner2011scoring develops a general framework for these desperation strategies, which include the onside kick in American football, pulling the infield and/or outfield in baseball, and of course, the fabled Hack-a-Shaq strategy in basketball.

### 2023-01-03 sports_analytics.Rmd
ctv access date; close #50

@@
 While @Lopez2018 use betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
-Early work by @gandar1988testing examines the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
+Early work by @gandar1988testing examine the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
 Related work such as @lacey1990estimation and @Boulier2006 explores the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
@@
 Table \ref{tab:ctv} provides an overview of the currently available sport-specific CRAN packages.
-@casals2022systematic offers a systematic review of sport-related packages on CRAN.
+Recently, @casals2022systematic offer a systematic review of sport-related packages on CRAN.
 Further, a more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].
@@
     linesep = "",
-    caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN).
+    caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN) as of October 16, 2022.
 While the major North American sports dominate the list, perhaps the fastest-growing collection is for esports.

### 2023-01-09 response_to_reviewers_R1.Rmd
goalie-pulling; #43
fix wording; close #47
added betting example; close #46

@@
 linearity assumption.
+
+> We have fixed the wording of this sentence. Instead of "extend", it now says "consider a similar approach".
+
 - Also in Section 2.1.2, it might be worthwhile to briefly define absorbing states, as I for
@@
 pulling the goalie optimal?
+
+> We have added a clarification that coaches usually wait too long to pull their goalies.
+
 - In the second formula of Section 4.1, the right hand side of the equation has a 2, but
@@
 
+> Thank you for this suggestion.
+We renamed Section 5: Sports betting market data and added this as a 4th "big idea".
+We moved the short discussion of the use of proprietary biometric data to the end of Section 7.
+We also added a short example of how betting lines could be used to compute implied team strengths.
+
 ## Minor Changes from Reviewer 2

### 2023-01-09 sports_analytics.Rmd
small citation fixes
added betting example; close #46

@@
   Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years.
-  In this paper, we connect three big ideas that are common across multiple sports: the expected value of a game state, win probability, and measures of team strength.
+  In this paper, we connect four big ideas that are common across multiple sports: the expected value of a game state, win probability, measures of team strength, and the use of sports betting market data.
   For each, we explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
@@
   Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
-  We also describe the use of non-performance data and opportunities to share analytical work, but omit an in-depth discussion of individual player evaluation as beyond our scope.
+  We also describe opportunities to share analytical work, but omit an in-depth discussion of individual player evaluation as beyond our scope.
   This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
@@
 
-In this paper, we explore three key ideas that have widespread applicability across many sports: the expected value of a game state (Section \ref{sec:ev}), win probability (Section \ref{sec:wp}), and measures of team strength (Section \ref{sec:strength}).
+In this paper, we explore four key ideas that have widespread applicability across many sports: the expected value of a game state (Section \ref{sec:ev}), win probability (Section \ref{sec:wp}), measures of team strength (Section \ref{sec:strength}), and the use of sports betting market data (Section \ref{sec:betting}).
 In each case, we define the concept mathematically, explain how it originated, and give examples of its applications in multiple sports. 
@@
 Doing the work of sports analytics requires computing with data. 
-While the sources of sports data are too numerous to list, we discuss the use of non-performance data (most notably, sports market betting data) in Section \ref{sec:betting}, and in Section \ref{sec:tools} we highlight a few computational tools (including a table of R packages) that make this kind of work possible.
+While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools (including a table of R packages) that make this kind of work possible. 
 Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
@@
 
-# Non-performance data {#sec:betting}
+# Sports betting market data {#sec:betting}
 
 Most of the research in sports analytics is fueled by the analysis of data recorded from the outcome of sports contests.
-However, a growing body of literature is informed by non-performance data about sports. 
-This research appears in fields like sports economics and kinesiology that are subfields of mature disciplines (a treatment of either is well beyond the scope of the this paper).
-In addition to this body of research, we note that today, professional sports teams are collecting biometric data on their own players, and presumably using that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance.
-While this research may constitute "emerging methodologies," it will likely never be published, because each team will use it to their competitive advantage.
-
-Nevertheless, in this Section we discuss concepts stemming from one notable, relatively new source of (semi-)public non-performance sports data: betting markets.
+However, a growing body of literature is informed by data from sports betting markets. 
 Since the 2018 United States Supreme Court decision in [*Murphy v.
@@
 Some researchers have then tried to demonstrate a positive return on some of these betting strategies, with very limited success.
+
+## Example of how to use betting market data
+
+To see how betting market data can be used to estimate team strengths, consider the betting odds posted on [FanDuel Sportsbook for the 2023 NBA Champion](https://sportsbook.fanduel.com/navigation/nba?tab=nba-finals), posted on January 9, 2023 and shown in Figure \ref{tab:odds}.
+This is a futures market, because the actual NBA champion will not be determined until June 2023. 
+The Boston Celtics are the favorite to win, with posted odds of $+390$, meaning that a \$100 bet on the Celtics to win the championship will pay out \$390 if the Celtics win it all. 
+Conversely, six teams share the lowest odds at $+50000$.
+
+```{r odds}
+library(tidyverse)
+odds <- read_csv("odds.csv")
+odds |>
+  mutate(
+    prob = 100 / (100 + line),
+    `Prob. Normalized` = prob / sum(prob)
+  ) |>
+  rename(Team = team, Line = line, `Prob.` = prob) |>
+  mutate(Rank = row_number(), .before = Team) |>
+  janitor::adorn_totals() |>
+  slice(c(1:6, 25:31)) |>
+  knitr::kable(
+    booktabs = TRUE, 
+    linesep = c('', '', '', '', '', '\\addlinespace'),
+    digits = 3, 
+    caption = "2023 NBA Championship line odds for the top 6 and bottom 6 teams. Retrieved from FanDuel Sportsbook on January 9, 2023. ")
+```
+
+These lines ($\ell_i$) can be converted into an implied probability ($p_i$) using the formula:
+$$
+  p_i = \frac{100}{100 + \ell_i} \,.
+$$
+The sum of those probabilities is greater than one---this is why the sportsbook makes money regardless of who wins the championship. 
+However, the implied probabilities can be normalized by dividing by their sum to recover true probabilities of each team winning the championship. 
+Many different researchers have shown that these normalized implied probabilities are accurate, unbiased, and efficient estimates of the true unknowable probabilities (see @Lopez2018 for discussion and an extensive list of references).
+
+In this case, the FanDuel futures market suggests that the Celtics have a 16.3\% chance of winning the championship, while the Milwaukee Bucks have the second best chance, at 13.3\%. 
+These implied probabilities can be used to fit various models for team strength, as described in Section \ref{sec:strength}.
 
@@
 
+It is also worth noting that a significant, but unknown, proportion of the most innovative research is being conducted by professional sports teams. 
+This research will likely never be published, because each team will use it to their competitive advantage. 
+Part of what enables this research is better data. 
+For example, professional sports teams can collect biometric data on their own players, and use that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance.
+While this research may constitute "emerging methodologies," it unfortunately will take years, if at all, before the public benefits from it.
+
 # Conclusion {#sec:conclusion}

### 2023-01-09 sports_analytics.Rmd
connected comments on effect sizes; close #48

@@
 They state that by failing to account for opponent strength appropriately, about 124.8 points per team each season (or about 3.8 wins per season) are not credited correctly.
+This is a substantial amount in a 16-game season.
 
@@
 In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a causal inference framework to reach similar conclusions that NFL teams are too conservative in going for it on 4th down.
-In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.
+In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount comparable to the effect size reported by @pelechrinis2019 above.
 

### 2023-01-09 sports_analytics.Rmd
Made the Lopez statement clearer.

@@
 Many past analyses of the decision to go for it or not on 4th down conclude that coaches in the NFL are too conservative in their decision making. 
-@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but the effect may be overstated due to the way in which to go yardage has been recorded in the past.
+@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but further notes that past estimates of the magnitude of how conservative coaches are on 4th down may be overstated due to the way in which to go yardage was recorded only approximately in the past.
 

### 2023-01-09 sports_analytics.Rmd
small updates on bayesian state-space

@@
 @macdonald2012expected uses expected goals to evaluate ice hockey players, but does not have access to player tracking data necessary to evaluate possessions.
-@kumagai2021hockey offer an EPV metric for ice hockey.
+@kumagai2021hockey offer an EPV metric for ice hockey via a Bayesian space-time framework.
 
@@
 
-@koopman2015dynamic use a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.
+In a similar Bayesian setting, @koopman2015dynamic studies English Premier League soccer match results by assuming a bivariate Poisson distribution with time-varying team abilities.
+This state-space approach appears to improve on bookmaker's odds.
 

### 2023-01-09 sports_analytics.Rmd
plural verb

@@
 
-In a similar Bayesian setting, @koopman2015dynamic studies English Premier League soccer match results by assuming a bivariate Poisson distribution with time-varying team abilities.
+In a similar Bayesian setting, @koopman2015dynamic study English Premier League soccer match results by assuming a bivariate Poisson distribution with time-varying team abilities.
 This state-space approach appears to improve on bookmaker's odds.

### 2023-01-10 response_to_reviewers_R1.Rmd
connected comments on effect sizes; close #48
close #49

@@
 et. al. estimated.
+
+> We have moved the comment up and connected the second comment to the first.
+
 - At the end of Section 2.1.3, you write “Lopez (2020) also concludes that coaches are
@@
 referring to the estimated effects past studies which used less granular data.
+
+> We have clarified this language.
+It now reads: "Lopez (2020) also concludes that coaches are too conservative on 4th down decision making, but notes further that past estimates of the magnitude of how conservative coaches are on 4th down may be overstated due to the way in which to go yardage was recorded only approximately in the past."
+
 - In Section 2.3, you write “NHL coaches do not always employ the optimal strategies”,

### 2023-01-10 sports_analytics.Rmd
close #49

@@
 Many past analyses of the decision to go for it or not on 4th down conclude that coaches in the NFL are too conservative in their decision making. 
-@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but further notes that past estimates of the magnitude of how conservative coaches are on 4th down may be overstated due to the way in which to go yardage was recorded only approximately in the past.
+@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but notes further that past estimates of the magnitude of how conservative coaches are on 4th down may be overstated due to the way in which to go yardage was recorded only approximately in the past.
 
@@
 
-## Example of how to use betting market data
-
-To see how betting market data can be used to estimate team strengths, consider the betting odds posted on [FanDuel Sportsbook for the 2023 NBA Champion](https://sportsbook.fanduel.com/navigation/nba?tab=nba-finals), posted on January 9, 2023 and shown in Figure \ref{tab:odds}.
+## Example: Win probabilities from betting market data
+
+To see how betting market data can be used to estimate team strengths, consider the betting lines posted on [FanDuel Sportsbook for the 2023 NBA Champion](https://sportsbook.fanduel.com/navigation/nba?tab=nba-finals) on January 9, 2023 and shown in Table \ref{tab:odds}.
 This is a futures market, because the actual NBA champion will not be determined until June 2023. 
-The Boston Celtics are the favorite to win, with posted odds of $+390$, meaning that a \$100 bet on the Celtics to win the championship will pay out \$390 if the Celtics win it all. 
+The Boston Celtics are the favorite to win, with a moneyline of $+390$, meaning that a \$100 bet on the Celtics to win the championship will pay back the original bet and an additional \$390 if the Celtics win it all.
+This style of odds are sometimes called [*American odds*](https://en.wikipedia.org/wiki/Fixed-odds_betting#Moneyline_odds). 
+The corresponding fractional odds have the Celtics at 4.9:1 to win the championship. 
 Conversely, six teams share the lowest odds at $+50000$.
@@
   mutate(
+    Odds = (line / 100) + 1,
     prob = 100 / (100 + line),
@@
     digits = 3, 
-    caption = "2023 NBA Championship line odds for the top 6 and bottom 6 teams. Retrieved from FanDuel Sportsbook on January 9, 2023. ")
-```
-
-These lines ($\ell_i$) can be converted into an implied probability ($p_i$) using the formula:
+    caption = "2023 NBA Championship odds for the top 6 and bottom 6 teams. Retrieved from FanDuel Sportsbook on January 9, 2023. ")
+```
+
+These moneylines ($\ell_i$) can be converted into an implied probability ($p_i$) using the formula:
 $$

### 2023-01-10 sports_analytics.Rmd
final touches

@@
 
-## Example: Win probabilities from betting market data
+## Example: Win probabilities from betting market data {#sec:moneylines}
 
@@
 Many different types of bets can be placed on sports. 
-For individual contests, bets involve point spreads, moneylines, odds, or other ways of handicapping which team will win. 
+For individual contests, bets involve point spreads, moneylines (see Section \ref{sec:moneylines} for an example), odds, or other ways of handicapping which team will win. 
 Money can also be wagered on futures, where odds are given in advance for events that may or may not transpire (e.g., a certain team making the playoffs, or a certain player winning the MVP award). 
-Here, we focus on betting pools, in which a group of people compete to predict winners in multiple contests (often a tournament), and the inevitable question of whether strategies exist that will consistently beat the market.
+Here, we focus on betting pools, in which a group of people compete to predict winners in multiple contests (often a tournament).
+We also address the inevitable question of whether strategies exist that will consistently beat the market.
 

### 2023-01-10 response_to_reviewers_R1.Rmd
final touches
source figure

@@
 date: "2023-01-03"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 
-> Thank you for this suggestion. We have added the three ideas explicitly to the abstract.
+> Thank you for this suggestion. We have added the three (now four) big ideas explicitly to the abstract.
 
@@
 - In the second formula of Section 4.1, the right hand side of the equation has a 2, but
-it should be an α.
+it should be an $\alpha$.
 
@@
 
-> We have a statement to this effect to the abstract.
+> We have a statement to this effect to the abstract. It now reads: "We also describe opportunities to share analytical work, but
+omit an in-depth discussion of individual player evaluation as beyond our scope."
 

### 2023-01-10 sports_analytics.Rmd
source figure

@@
 Brooklyn Nets (black) NBA game on November 1, 2013.
-The captured moment is when Miami's Mario Chalmers just releases a three-point shot, which ends up as a missed field goal."}
+The captured moment is when Miami's Mario Chalmers just releases a three-point shot, which ends up as a missed field goal.
+Figure created by Cervone, et al.
+\\url{https://github.com/dcervone/EPVDemo}"}
 # Source: https://github.com/dcervone/EPVDemo/blob/master/gifs/poss_12.gif

### 2023-04-11 sports_analytics.Rmd
typo
proofreading comments

@@
   - name: Benjamin S. Baumer
-#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
+    thanks: "Corresponding author, <bbaumer@smith.edu>"
     department: Statistical & Data Sciences
@@
   In this paper, we connect four big ideas that are common across multiple sports: the expected value of a game state, win probability, measures of team strength, and the use of sports betting market data.
-  For each, we explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
+  For each, we explore both the shared similarities and individual idiosyncrasies of analytical approaches in each sport. 
   While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources.
